---
output: html_document
editor_options:
  chunk_output_type: inline
---
# Time resolution {#sec-time}

Chapter 1 lived in a single year, a single slice, a single region — enough to see
the energy balance and nothing to distract from it. This chapter adds **time**:
a horizon of several years to invest across, a discount rate to weigh the future,
hours and seasons within the year, ageing capacity (**vintages**), and the two
things that only make sense once time is resolved — **storage** and
**intermittent** renewables.

We keep **one region** (`R1`) throughout; regions arrive in @sec-regions.

```{r}
source("R/workshop-model.R")
library(dplyr) # filter(), select(), joins used in the exercises
```

::: {.callout-note appearance="simple"}
Outputs shown below are illustrative. With `eval: false` set for the book,
nothing runs at render time — copy each chunk into your own session to see the
real numbers, which is also how you check the ones in the prose.
:::

## A model that spans years {#sec-horizon}

So far the "horizon" was one year. `newHorizon()` describes a **study period** and
how it is divided into **intervals**, each solved as one representative
*milestone* year. Investment decisions are made once per milestone and carried
forward.

```{r}
ELC <- newCommodity(
  name      = "ELC",
  desc      = "Electricity",
  unit      = "PJ",
  timeframe = "ANNUAL"
)

COA <- newCommodity(
  name      = "COA",
  desc      = "Coal",
  unit      = "PJ",
  timeframe = "ANNUAL",
  emis      = data.frame(comm = "CO2", unit = "kt/PJ", emis = 95)
)

GAS <- newCommodity(
  name      = "GAS",
  desc      = "Natural gas",
  unit      = "PJ",
  timeframe = "ANNUAL",
  emis      = data.frame(comm = "CO2", unit = "kt/PJ", emis = 56)
)

CO2 <- newCommodity(
  name      = "CO2",
  desc      = "Carbon dioxide",
  unit      = "kt",
  timeframe = "ANNUAL"
)

SUP_COA <- newSupply(
  name      = "SUP_COA",
  commodity = "COA",
  unit      = "PJ",
  supply    = data.frame(cost = 2.5)
)

SUP_GAS <- newSupply(
  name      = "SUP_GAS",
  commodity = "GAS",
  unit      = "PJ",
  supply    = data.frame(cost = 6.0)
)

# demand grows from 50 PJ (2025) to 80 PJ (2050); energyRt interpolates between
DEM_ELC <- newDemand(
  name      = "DEM_ELC",
  commodity = "ELC",
  unit      = "PJ",
  demand    = data.frame(year = c(2025, 2050), demand = c(50, 80))
)

ECOA <- newTechnology(
  name    = "ECOA",
  desc    = "Coal plant",
  input   = list(comm = "COA", unit = "PJ", combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "COA", cinp2use = 0.40),
  invcost = list(invcost = 2000),
  fixom   = 55,
  cap2act = 31.536,
  olife   = 30L
)

EGAS <- newTechnology(
  name    = "EGAS",
  desc    = "Gas plant",
  input   = list(comm = "GAS", unit = "PJ", combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "GAS", cinp2use = 0.58),
  invcost = list(invcost = 900),
  fixom   = 25,
  cap2act = 31.536,
  olife   = 25L
)

draw(ECOA) # each technology's reference-energy-system diagram: coal in, ELC out
draw(EGAS)

repo_YR <- newRepository(
  "repo_YR",
  ELC, COA, GAS, CO2,
  SUP_COA, SUP_GAS, DEM_ELC, ECOA, EGAS
)
```

The horizon below begins with a one-year base period and adds three ten-year
intervals. `autoplot()` shows the milestone years it produces.

```{r}
horizon_2025_2050_by10 <- newHorizon(
  period    = 2025:2055,
  intervals = c(1, 10, 10, 10)
)

autoplot(horizon_2025_2050_by10)
```

`solve_model()` does in one step what you did by hand in Chapter 1 — it expands
the description over years and slices and sends it to the solver.

```{r}
mod_YR <- newModel(
  name     = "YR",
  data     = repo_YR,
  region   = "R1",
  discount = WS_DISCOUNT,
  horizon  = horizon_2025_2050_by10
)

scen_YR <- solve_model(mod_YR)

getData(scen_YR, "vTechCap", merge = TRUE)   # capacity by technology and milestone
getData(scen_YR, "vObjective", merge = TRUE) # total discounted cost, MEUR
```

A handful of milestone years — not one per calendar year — keeps the model small
while still spanning several decades of investment.

**2.1 — Temporal aggregation and problem size.** Re-solve the model with a fully
annual horizon, `newHorizon(period = 2025:2055)`, in which each calendar year
constitutes its own period. Compare the resulting problem size (`model_size()`)
and objective value with those of the milestone formulation. Assess whether the
additional temporal detail materially alters the investment path, and whether it
justifies the increase in problem dimension.

::: {.callout-tip collapse="true" title="Solution 2.1"}
```{r}
horizon_annual <- newHorizon(period = 2025:2055)

mod_annual <- newModel(
  name     = "YR_annual",
  data     = repo_YR,
  region   = "R1",
  discount = WS_DISCOUNT,
  horizon  = horizon_annual
)

scen_annual <- solve_model(mod_annual)

model_size(scen_YR)                             # four milestones
model_size(scen_annual)                         # one period per year

getData(scen_YR, "vObjective", merge = TRUE)
getData(scen_annual, "vObjective", merge = TRUE)
```
The annual formulation is several times larger, yet the investment path is nearly
identical: with smooth, monotone demand there is little that a yearly grid
captures which four milestones miss. Milestones are the default precisely because
they recover most of the accuracy at a fraction of the size. The detail that
*does* matter — hours and seasons — lies **within** the year, and occupies the
remainder of this chapter.
:::

## Discounting: business vs social planner {#sec-discount}

Spanning years forces a question Chapter 1 could ignore: a euro spent in 2050 is
not a euro spent today. energyRt uses **two** rates, and it is worth keeping them
apart:

- **SDR** (social discount rate) — discounts the *stream of system costs* in the
  objective. This is the planner's view of the future.
- **WACC** (weighted average cost of capital) — annuitises a lump investment into
  an *equivalent annual cost* (EAC). This is how expensive capital is to the
  entity that builds.

The `discount =` argument is a shorthand that sets **both** rates equal
(`discount = 0.05` ⇒ `wacc = sdr = 0.05`). A social planner adopts one low rate
over each asset's full life. A private investor applies a **higher WACC** and
often recovers capital over a **shorter payback** than the asset operates.

```{r}
# social planner: one low rate, capital recovered over the full technical life
mod_soc <- newModel(
  name     = "SOC",
  data     = repo_YR,
  region   = "R1",
  discount = 0.03,
  horizon  = horizon_2025_2050_by10
)

# investor: split rates, and the coal plant must repay its capital in 15 years
ECOA_inv <- update(
  ECOA,
  invcost = list(invcost = 2000, wacc = 0.10, payback = 15)
)

repo_inv <- newRepository(
  "repo_inv",
  ELC, COA, GAS, CO2,
  SUP_COA, SUP_GAS, DEM_ELC, ECOA_inv, EGAS
)

mod_inv <- newModel(
  name     = "INV",
  data     = repo_inv,
  region   = "R1",
  discount = data.frame(wacc = 0.10, sdr = 0.03),
  horizon  = horizon_2025_2050_by10
)

scen_soc <- solve_model(mod_soc)
scen_inv <- solve_model(mod_inv)

getData(scen_soc, "vTechEac", merge = TRUE)  # annualised capital charge
getData(scen_inv, "vTechEac", merge = TRUE)  # higher under investor terms
```

Same `invcost`, very different annual capital charge: a 10% WACC recovered over 15
years annuitises far more steeply than 3% over 30 years, so the investor's coal
plant appears dearer and the model leans harder on gas.

::: {.callout-note appearance="simple"}
The `wacc`/`sdr` split works on every backend, but the finite **`payback`**
feature is, as of energyRt 0.80.x, implemented **only on GLPK** for testing. Its support in other backends (GAMS, Pyomo, JuMP) is on the way.  
Omitting `discount` entirely leaves both rates at **0** (an undiscounted model), so always set it.  
There is no per-technology `sdr` — the social rate is a property of the model.
:::

The same divergence can be shown without solving, using `levcost()` (whose
`discount` argument *is* the WACC).

**2.2 — The cost of capital and technology choice.** A photovoltaic plant is
almost entirely capital and carries no fuel cost; a gas peaker is the converse.
Using `levcost()`, compute the levelised cost of a capital-intensive generator
under (a) the social-planner assumption (3%, full technical life) and (b) the
investor assumption (10%, 15-year payback). Determine which technology's LCOE is
more sensitive to the cost of capital, and explain how that sensitivity shifts the
least-cost generation mix toward or away from renewables.

::: {.callout-tip collapse="true" title="Solution 2.2"}
```{r}
# a capital-heavy generator: high invcost, no fuel
SOLARish <- newTechnology(
  name    = "SOLARish",
  desc    = "Capital-heavy generator",
  input   = list(comm = "SOL", unit = "PJ"),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "SOL", cinp2use = 1),
  invcost = list(invcost = meur_gw(650)),
  fixom   = 12,
  cap2act = 31.536,
  olife   = 25L
)

lc_soc <- levcost(SOLARish, discount = 0.03) # planner

SOLAR_inv <- update(
  SOLARish,
  invcost = list(invcost = meur_gw(650), wacc = 0.10, payback = 15)
)

lc_inv <- levcost(SOLAR_inv, discount = 0.10) # investor

lc_soc$levcost_npv
lc_inv$levcost_npv
```
The capital-intensive plant's LCOE rises sharply under investor assumptions —
most of its lifetime cost is the annuity, and the annuity is precisely what the
discount rate acts upon. A gas peaker scarcely moves, because most of *its* cost
is fuel, which no discount rate touches. A high cost of capital is therefore a
quiet thumb on the scale **against** renewables and **in favour of** fuel-burning
plant: the business case and the social case can rank the same technologies
differently from identical physics.
:::

## Within the year: seasons and hours {#sec-subannual}

A commodity balanced on `ANNUAL` need only add up over the whole year — a solar
panel could "supply" midnight demand. Real electricity balances **every hour**. A
**calendar** divides the year into time slices; a commodity's `timeframe` states
which level it balances on.

```{r}
WS_CAL # calendars$utopia_s4h24: 4 seasons x 24 h = 96 slices

head(WS_CAL@timeslice_share) # each slice's share of the year

autoplot(WS_CAL) # the calendar's nested season / hour structure
```

::: {.callout-note appearance="simple"}
Calendars come from the **timeslices** package; its successor **timescales**
generalises them to arbitrary nested timeframes, and its spatial twin
**geoscales** does the same for regions — the time and space "bricks" of the
*optimal2050* ecosystem that @sec-regions builds on.
:::

To balance electricity hourly, set `timeframe = "HOUR"` on `ELC` and attach the
calendar to the model.

```{r}
ELC_h <- update(ELC, timeframe = "HOUR")
```

**2.3 — Temporal resolution and the visibility of storage.** Solve the
electricity model on three calendars of increasing resolution —
`calendars$utopia_annual` (1 slice), `calendars$utopia_seasons` (12), and
`calendars$utopia_s4h24` (96) — supplied through `newModel(..., calendar = )`.
Observe how the problem size grows, and explain why the value of storage is
invisible at annual resolution and becomes apparent only once the day acquires an
internal shape.

::: {.callout-tip collapse="true" title="Solution 2.3"}
```{r}
mod_annual_cal <- newModel(
  name     = "RES_annual",
  data     = repo_YR,
  region   = "R1",
  discount = WS_DISCOUNT,
  calendar = calendars$utopia_annual,
  horizon  = horizon_2025_2050_by10
)

mod_seasons <- newModel(
  name     = "RES_seasons",
  data     = repo_YR,
  region   = "R1",
  discount = WS_DISCOUNT,
  calendar = calendars$utopia_seasons,
  horizon  = horizon_2025_2050_by10
)

mod_s4h24 <- newModel(
  name     = "RES_s4h24",
  data     = repo_YR,
  region   = "R1",
  discount = WS_DISCOUNT,
  calendar = calendars$utopia_s4h24,
  horizon  = horizon_2025_2050_by10
)

scen_annual_cal <- solve_model(mod_annual_cal)
scen_seasons    <- solve_model(mod_seasons)
scen_s4h24      <- solve_model(mod_s4h24)

model_size(scen_annual_cal)
model_size(scen_seasons)
model_size(scen_s4h24)

getData(scen_annual_cal, "vObjective", merge = TRUE)
getData(scen_seasons, "vObjective", merge = TRUE)
getData(scen_s4h24, "vObjective", merge = TRUE)
```
More slices, more rows — and a model able to distinguish noon from midnight. On an
annual calendar a battery has no intraday gap to arbitrage, so the optimiser never
builds one; the value it contributes in @sec-storage cannot appear until the
calendar carries sub-daily structure. Resolution is not decoration: it determines
which technologies can even be *seen* to be useful.
:::

## Vintages: one technology, evolving parameters {#sec-vintages}

A coal plant commissioned in 2040 is more efficient than one built in 2020 — yet
it is still *the coal-plant technology*, not a new one. **Vintages** are how a
single technology carries **parameters that depend on the year it was built**:
efficiency, investment cost, fixed O&M, even operating life can each take a
different value per build-year cohort, while the technology stays one object in the
model. A vintage names that cohort — a build window (`start`/`end`) with its own
numbers.

```{r}
# one coal technology whose efficiency improves with each build-year cohort
ECOA_v <- newTechnology(
  name    = "ECOA",
  desc    = "Coal plant, improving by build year",
  input   = list(comm = "COA", unit = "PJ", combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(
    vintage  = c("2020", "2030", "2040"),
    comm     = "COA",
    cinp2use = c(0.40, 0.43, 0.46)   # newer builds burn coal more efficiently
  ),
  invcost = list(invcost = 2000),
  fixom   = 55,
  cap2act = 31.536,
  vintage = data.frame(
    vintage = c("2020", "2030", "2040"),
    start   = c(2020, 2030, 2040),
    end     = c(2029, 2039, 2050),
    olife   = 30
  )
)

draw(ECOA_v)
```

The electrolyser and vehicle technologies in @sec-processes are built exactly this
way — the `FPEM100` electrolyser's hydrogen efficiency climbs from 0.53 (2020) to
0.66 (2050) across its vintages, and the one spec still describes one machine.

A **second**, related use of the same machinery is an existing **fleet** that
retires over time: give a technology a `capacity` stock declared at several years
and energyRt reads it as a declining retirement path, optionally letting the
optimiser retire units early.

```{r}
# the same technology, now also carrying an existing fleet that retires by 2050
ECOA_fleet <- update(
  ECOA_v,
  capacity = data.frame(year = c(2025, 2050), stock = c(1.2, 0)),
  optimizeRetirement = TRUE
)
```

**2.4 — Parameters by build year.** Put `ECOA_v` in the model, solve, and read
`vTechCap` / `vTechNewCap` by vintage. Confirm that capacity built in later years
carries the more efficient coefficients, and use `autoplot(mod, type = "windows")`
to visualise each vintage's build window against its operating life.

::: {.callout-tip collapse="true" title="Solution 2.4"}
```{r}
repo_v <- newRepository(
  "repo_v",
  ELC, COA, GAS, CO2,
  SUP_COA, SUP_GAS, DEM_ELC, ECOA_v, EGAS
)

mod_v <- newModel(
  name     = "VINT",
  data     = repo_v,
  region   = "R1",
  discount = WS_DISCOUNT,
  horizon  = horizon_2025_2050_by10
)

autoplot(mod_v, type = "windows") + theme_ws()   # vintage build windows vs olife

scen_v <- solve_model(mod_v)

getData(scen_v, "vTechCap", merge = TRUE)    # installed capacity, by vintage and year
getData(scen_v, "vTechNewCap", merge = TRUE) # new build, by vintage
```
Each vintage is a distinct cohort: capacity built from 2040 carries
`cinp2use = 0.46` and burns less coal per unit of electricity than the 2020
cohort's 0.40 — even though both are "the coal plant." Vintaging is what lets one
technology track a changing world (improving efficiency, falling costs, an ageing
fleet) without multiplying into a dozen near-duplicate objects.
:::

## Storage and intermittent sources {#sec-storage}

Two technologies acquire meaning only once the day has a shape. **Weather-driven**
generators (solar, wind) produce on a fixed profile rather than on command;
**storage** moves energy between slices.

```{r}
# free renewable resources (no fuel cost), balanced hourly
SOL <- newCommodity("SOL", unit = "PJ", timeframe = "HOUR")
WIN <- newCommodity("WIN", unit = "PJ", timeframe = "HOUR")

SUP_SOL <- newSupply(
  name      = "SUP_SOL",
  commodity = "SOL",
  unit      = "PJ",
  supply    = data.frame(cost = 0)
)

SUP_WIN <- newSupply(
  name      = "SUP_WIN",
  commodity = "WIN",
  unit      = "PJ",
  supply    = data.frame(cost = 0)
)

# capacity-factor profiles for R1 on this calendar (deterministic, shipped)
weather_WSOL <- prof$weather |>
  filter(resource == "WSOL") |>
  select(region, timeslice, wval)

weather_WWIN <- prof$weather |>
  filter(resource == "WWIN") |>
  select(region, timeslice, wval)

WSOL <- newWeather(
  name      = "WSOL",
  timeframe = "HOUR",
  weather   = weather_WSOL
)

WWIN <- newWeather(
  name      = "WWIN",
  timeframe = "HOUR",
  weather   = weather_WWIN
)

# waf.fx = 1 fixes output to the weather profile; because electricity is a
# free-disposal commodity (limtype "LO"), any surplus is dumped and appears in
# vBalance -- which is how we measure curtailment in @sec-curtailment
ESOL <- newTechnology(
  name    = "ESOL",
  desc    = "Solar PV",
  input   = list(comm = "SOL", unit = "PJ"),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "SOL", cinp2use = 1),
  weather = list(weather = "WSOL", comm = "SOL", waf.fx = 1),
  invcost = list(invcost = meur_gw(650)),
  fixom   = 12,
  cap2act = 31.536,
  olife   = 25L
)

EWIN <- update(
  ESOL,
  name    = "EWIN",
  desc    = "Onshore wind",
  input   = list(comm = "WIN", unit = "PJ"),
  ceff    = data.frame(comm = "WIN", cinp2use = 1),
  weather = list(weather = "WWIN", comm = "WIN", waf.fx = 1),
  invcost = list(invcost = meur_gw(1300)),
  fixom   = 35
)

# a 4-hour battery on electricity
STG_ELC <- newStorage(
  name      = "STG_ELC",
  desc      = "Grid battery, 4 h",
  commodity = "ELC",
  invcost   = list(invcost = meur_pj_kwh(200)),
  cap2stg   = 4,
  seff      = data.frame(inpeff = 0.95, outeff = 0.95),
  olife     = 20L
)

draw(ESOL)    # solar: the SOL resource in, electricity out
draw(STG_ELC) # the battery, charging and discharging electricity
```

Assemble the hourly model (electricity now `timeframe = "HOUR"`) and solve.

```{r}
repo_VRE <- newRepository(
  "repo_VRE",
  ELC_h, COA, GAS, CO2, SOL, WIN,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, DEM_ELC,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC
)

mod_VRE <- newModel(
  name     = "VRE",
  data     = repo_VRE,
  region   = "R1",
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = horizon_2025_2050_by10
)

scen_VRE <- solve_model(mod_VRE)

getData(scen_VRE, "vTechCap", merge = TRUE)                           # the mix
getData(scen_VRE, "vStorageInp", merge = TRUE, timeframe = "highest") # charging

autoplot(scen_VRE) + theme_ws() # the generation mix by technology and year
```

**2.5 — Complementarity of wind and solar resources.** Plot the capacity factors
of `WSOL` and `WWIN` across the day and the seasons
(`autoplot(WSOL, calendar = WS_CAL)`). Identify the periods in which one resource
rises as the other falls, and explain how this complementarity alters the demand
placed on storage.

::: {.callout-tip collapse="true" title="Solution 2.5"}
```{r}
autoplot(WSOL, calendar = WS_CAL) + theme_ws()
autoplot(WWIN, calendar = WS_CAL) + theme_ws()
```
Solar is a midday peak that collapses to zero at night; wind is flatter and often
stronger overnight and in winter. Because the two peak at different times, a
combination of them requires *less* storage than either alone — the battery need
only bridge what the pair still leaves uncovered.
:::

## Curtailment {#sec-curtailment}

Because `waf.fx = 1` **fixes** each renewable's output to its weather profile, a
panel or turbine produces its full potential every hour, whether or not it is
needed. Electricity is a **free-disposal** commodity by default
(`limtype = "LO"`, so its balance is *production ≥ consumption*), so the model may
simply dump any surplus. That dumped surplus — electricity generated but not
consumed — *is* the balance variable **`vBalance`**, so curtailment reads straight
off it, with no reconstruction:

```
vBalance(ELC) = Σ vOutTot − Σ vInpTot  ≥ 0    (the curtailed surplus)
```

Had we instead set `waf.up = 1`, the optimiser would *under-produce* rather than
spill, `vBalance` would be identically zero (and so absent from the output, which
records only non-zero balances), and curtailment would have to be inferred from
potential minus output.

There is a catch: whether any surplus arises at all depends on how much renewable
capacity is built. In the least-cost mix above, cheap dispatchable thermal is free
to fill any gap, so the optimiser builds only modest solar and wind and backs the
thermal fleet down when they are plentiful — almost nothing is spilled, and
`vBalance` is essentially zero. **Curtailment is a phenomenon of high renewable
penetration**, not of renewables as such.

**2.6 — Quantifying curtailment.** Confirm that the free least-cost mix curtails
little, then force a large solar fleet — so that its midday output must exceed
demand — re-solve, and read the curtailed surplus from `vBalance`. Explain how the
surplus responds to storage and to further increases in solar capacity.

::: {.callout-tip collapse="true" title="Solution 2.6"}
```{r}
# the free least-cost mix spills little or nothing
getData(scen_VRE, "vBalance", comm = "ELC", merge = TRUE)

# force a high solar penetration: at least 8 GW installed
ESOL_big <- update(
  ESOL,
  capacity = data.frame(year = 2050, cap.lo = 8)
)

repo_curt <- newRepository(
  "repo_curt",
  ELC_h, COA, GAS, CO2, SOL, WIN,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, DEM_ELC,
  ECOA, EGAS, ESOL_big, EWIN, WSOL, WWIN, STG_ELC
)

mod_curt <- newModel(
  name     = "CURT",
  data     = repo_curt,
  region   = "R1",
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = horizon_2025_2050_by10
)

scen_curt <- solve_model(mod_curt)

# now the surplus is real: curtailed electricity by slice, then as an annual total
getData(scen_curt, "vBalance", comm = "ELC", merge = TRUE, timeframe = "highest")
getData(scen_curt, "vBalance", comm = "ELC", merge = TRUE)
```
With cheap thermal on hand the model has no reason to over-build renewables, so the
free mix curtails almost nothing. Force a large solar fleet and the picture
changes: on sunny midday slices its output exceeds what demand and the battery can
absorb, and the surplus is dumped — a positive `vBalance`. Adding storage reduces
it, as energy is shifted to the evening; adding still more solar increases it.
Curtailment appears only when renewables are pushed beyond what demand and
flexibility can take up — which is why the fully renewable system of @sec-value in
the next chapter is where it truly bites.
:::

## What you built

Time turned a snapshot into a trajectory. You met the **horizon** and its
milestones, the **two discount rates** and how the business and social cases
diverge, **sub-annual** calendars and why resolution decides what is even
visible, **vintages** as ageing fleets, and the pair — **storage** and
**intermittent** generation — that exist only once the day has a shape, together
with the **curtailment** that follows. @sec-regions adds space: several regions,
trade between them, and where a new load wants to be.
