---
output: html_document
editor_options:
  chunk_output_type: inline
---
# Regions and trade {#sec-regions}

Everything so far lived in one region. Real systems are several places at once,
each with its own resources, its own demand, and links that let power flow between
them. This chapter adds **space**: a three-region electricity system, a
**transmission** link and a **rest-of-world import**, then two questions that only
space can pose — *where* a new electricity **load** wants to be, and what
**storage and transmission are worth** in a fully renewable system.

```{r}
source("R/workshop-model.R")
library(dplyr)
```

::: {.callout-note appearance="simple"}
Outputs below are illustrative — with `eval: false` nothing runs at render.
:::

## Three regions, three endowments {#sec-multiregion}

The trick to a multi-region model is that every data frame carries a `region`
column, so one object can describe all regions at once with different numbers.
Here the endowments are deliberately asymmetric: `R1` sits on cheap coal, `R2` on
gas, `R3` has the best wind and sun and no domestic fossil fuel.

::: {.callout-note appearance="simple"}
Regions here are flat labels `R1`–`R3`. The **geoscales** package generalises this
to nested spatial hierarchies (zones within regions within a system), just as
**timescales** does for time — the space and time "bricks" of the *optimal2050*
ecosystem.
:::

```{r}
REG <- c("R1", "R2", "R3")

# per-region VRE capacity-factor profiles (deterministic, shipped with energyRt)
prof3 <- utopia_profiles(REG, calendar = "utopia_s4h24")

weather_WSOL <- prof3$weather |>
  filter(resource == "WSOL") |>
  select(region, timeslice, wval)

weather_WWIN <- prof3$weather |>
  filter(resource == "WWIN") |>
  select(region, timeslice, wval)
```

```{r}
ELC <- newCommodity("ELC", unit = "PJ", timeframe = "HOUR")
COA <- newCommodity(
  name      = "COA",
  unit      = "PJ",
  timeframe = "ANNUAL",
  emis      = data.frame(comm = "CO2", unit = "kt/PJ", emis = 95)
)
GAS <- newCommodity(
  name      = "GAS",
  unit      = "PJ",
  timeframe = "ANNUAL",
  emis      = data.frame(comm = "CO2", unit = "kt/PJ", emis = 56)
)
CO2 <- newCommodity("CO2", unit = "kt", timeframe = "ANNUAL")
SOL <- newCommodity("SOL", unit = "PJ", timeframe = "HOUR")
WIN <- newCommodity("WIN", unit = "PJ", timeframe = "HOUR")

# domestic fossil supply exists only where the endowment is (R1 coal, R2 gas)
SUP_COA <- newSupply(
  name      = "SUP_COA",
  commodity = "COA",
  unit      = "PJ",
  supply    = data.frame(region = "R1", cost = 2.5)
)
SUP_GAS <- newSupply(
  name      = "SUP_GAS",
  commodity = "GAS",
  unit      = "PJ",
  supply    = data.frame(region = "R2", cost = 6.0)
)

# free renewable resources everywhere
SUP_SOL <- newSupply(
  name      = "SUP_SOL",
  commodity = "SOL",
  unit      = "PJ",
  supply    = data.frame(region = REG, cost = 0)
)
SUP_WIN <- newSupply(
  name      = "SUP_WIN",
  commodity = "WIN",
  unit      = "PJ",
  supply    = data.frame(region = REG, cost = 0)
)

# demand, larger in R1/R2 than in R3
DEM_ELC <- newDemand(
  name      = "DEM_ELC",
  commodity = "ELC",
  unit      = "PJ",
  demand    = data.frame(region = REG, year = 2050, demand = c(80, 70, 40))
)

WSOL <- newWeather(name = "WSOL", timeframe = "HOUR", weather = weather_WSOL)
WWIN <- newWeather(name = "WWIN", timeframe = "HOUR", weather = weather_WWIN)
```

```{r}
ECOA <- newTechnology(
  name    = "ECOA",
  input   = list(comm = "COA", unit = "PJ", combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "COA", cinp2use = 0.40),
  invcost = list(invcost = 2000),
  fixom   = 55,
  cap2act = 31.536,
  olife   = 30L,
  region  = REG
)
EGAS <- newTechnology(
  name    = "EGAS",
  input   = list(comm = "GAS", unit = "PJ", combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "GAS", cinp2use = 0.58),
  invcost = list(invcost = 900),
  fixom   = 25,
  cap2act = 31.536,
  olife   = 25L,
  region  = REG
)
ESOL <- newTechnology(
  name    = "ESOL",
  input   = list(comm = "SOL", unit = "PJ"),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "SOL", cinp2use = 1),
  weather = list(weather = "WSOL", comm = "SOL", waf.fx = 1),
  invcost = list(invcost = meur_gw(650)),
  fixom   = 12,
  cap2act = 31.536,
  olife   = 25L,
  region  = REG
)
EWIN <- update(
  ESOL,
  name    = "EWIN",
  input   = list(comm = "WIN", unit = "PJ"),
  ceff    = data.frame(comm = "WIN", cinp2use = 1),
  weather = list(weather = "WWIN", comm = "WIN", waf.fx = 1),
  invcost = list(invcost = meur_gw(1300)),
  fixom   = 35
)
STG_ELC <- newStorage(
  name      = "STG_ELC",
  commodity = "ELC",
  invcost   = list(invcost = meur_pj_kwh(200)),
  cap2stg   = 4,
  seff      = data.frame(inpeff = 0.95, outeff = 0.95),
  olife     = 20L,
  region    = REG
)

draw(ECOA) # coal in, electricity out -- the same diagram in every region
draw(ESOL)

hor <- newHorizon(2025:2055, c(1, 10, 10, 10))
```

**3.1 — Regions in isolation.** Assemble the six-technology system across `REG`
with **no** link between regions, solve, and read the generation mix per region.
Determine where each region lands, and explain what isolation costs the
renewables-rich `R3`.

::: {.callout-tip collapse="true" title="Solution 3.1"}
```{r}
repo_iso <- newRepository(
  "repo_iso",
  ELC, COA, GAS, CO2, SOL, WIN,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, DEM_ELC,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC
)

mod_iso <- newModel(
  name     = "ISO",
  data     = repo_iso,
  region   = REG,
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = hor
)

scen_iso <- solve_model(mod_iso)

getData(scen_iso, "vTechCap", merge = TRUE)               # capacity by region and tech
autoplot(scen_iso) + theme_ws()                           # generation mix per region
```
Each region solves its own least-cost problem: `R1` leans on its cheap coal, `R2`
on gas, and `R3` on wind and solar backed by whatever firm capacity it can afford
locally. Isolated, `R3` cannot sell its cheap renewable surplus to anyone, so it
over-builds firm capacity (or spills) — the first hint that a wire would pay.
:::

## A transmission link {#sec-trade}

`newTrade()` connects regions. We give it a **fixed** existing capacity — a wire
that already exists — rather than letting the model build one, so the value we
measure later is clean. `teff` is the throughput efficiency (losses).

```{r}
TBD_R1_R2 <- newTrade(
  name      = "TBD_R1_R2",
  commodity = "ELC",
  routes    = data.frame(src = c("R1", "R2"), dst = c("R2", "R1")),
  trade     = data.frame(src = c("R1", "R2"), dst = c("R2", "R1"), teff = 0.95),
  capacity  = data.frame(stock = 5),
  olife     = 50L
)
TBD_R2_R3 <- newTrade(
  name      = "TBD_R2_R3",
  commodity = "ELC",
  routes    = data.frame(src = c("R2", "R3"), dst = c("R3", "R2")),
  trade     = data.frame(src = c("R2", "R3"), dst = c("R3", "R2"), teff = 0.95),
  capacity  = data.frame(stock = 5),
  olife     = 50L
)
```

**3.2 — The value of the wire.** Add both links, solve, and compare total cost and
the generation mix with the isolated case from 3.1. Determine the direction of
flow and whether `R3` now exports its renewables. Draw the network with
`plot_trade_map()`.

::: {.callout-tip collapse="true" title="Solution 3.2"}
```{r}
repo_TR <- newRepository(
  "repo_TR",
  ELC, COA, GAS, CO2, SOL, WIN,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, DEM_ELC,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC,
  TBD_R1_R2, TBD_R2_R3
)

mod_TR  <- newModel(
  name     = "TR",
  data     = repo_TR,
  region   = REG,
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = hor
)

scen_TR <- solve_model(mod_TR)

getData(scen_TR, "vObjective", merge = TRUE)                   # vs scen_iso
getData(scen_TR, "vTradeIr", merge = TRUE, timeframe = "highest") # flows by slice

# the corridor network on a stylised map (needs the sf package)
plot_trade_map(repo_TR, map = utopia_modules$maps$honeycomb)
```
Total cost falls: cheap renewable energy in `R3` now displaces dearer generation
next door, so the system builds less firm plant overall. The flows run from
wherever power is momentarily cheapest toward wherever it is dearest — often out
of `R3` at midday. A wire is a substitute for both generation and storage.
:::

## Rest-of-world import {#sec-import}

An **import** brings a commodity in from outside the modelled regions at a price —
not a region, just a border. Here, imported gas is available everywhere as a
backstop fuel.

```{r}
IMP_GAS <- newImport(
  name      = "IMP_GAS",
  commodity = "GAS",
  unit      = "PJ",
  import    = data.frame(region = REG, price = 8.0) # dearer than R2's domestic gas
)
```

**3.3 — The marginal fuel.** Add `IMP_GAS` and solve. Determine where the model
chooses to import rather than build, and identify what sets the electricity price
at the margin in regions with no domestic fuel.

::: {.callout-tip collapse="true" title="Solution 3.3"}
```{r}
repo_M <- newRepository(
  "repo_M",
  ELC, COA, GAS, CO2, SOL, WIN,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, IMP_GAS, DEM_ELC,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC,
  TBD_R1_R2, TBD_R2_R3
)

mod_M  <- newModel(
  name     = "M",
  data     = repo_M,
  region   = REG,
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = hor
)

scen_M <- solve_model(mod_M)

getData(scen_M, "vImportRow", merge = TRUE) # imported gas, by region
```
`mod_M` is the model we carry into the policy scenarios of @sec-policy. Imports
fill the gap where building is not worth it — a region short of both fuel and wire
capacity leans on the border, and the import price becomes the ceiling that
domestic options must beat.
:::

## Where does a new load want to be? {#sec-siting}

Now the spatial question in its sharpest form. Suppose policy commits to deploying
a fixed amount of a **new electricity-consuming technology** — an electrolyser, a
fleet of data centres — and lets the model choose *which regions* host it. The
load consumes electricity to make a product `PRD` with an annual demand; a
system-wide **capacity target** (summed over regions, by year) fixes how much is
built; the optimiser places it.

```{r}
PRD     <- newCommodity("PRD", unit = "PJ", timeframe = "ANNUAL") # e.g. hydrogen
DEM_PRD <- newDemand(
  name      = "DEM_PRD",
  commodity = "PRD",
  unit      = "PJ",
  demand    = data.frame(year = 2050, demand = 30)                # met from any region
)

# a new load: consumes ELC (hourly), makes PRD; buildable in every region
LOAD_base <- newTechnology(
  name    = "LOAD",
  desc    = "New base load",
  input   = list(comm = "ELC", unit = "PJ"),
  output  = list(comm = "PRD", unit = "PJ"),
  ceff    = data.frame(comm = "ELC", cinp2use = 1.4), # 1 / efficiency
  af      = data.frame(af.lo = 0.9),                  # must run flat, every hour
  invcost = list(invcost = meur_gw(500)),
  fixom   = 10,
  cap2act = 31.536,
  olife   = 20L,
  region  = REG
)

# system-wide deployment target: total installed capacity over all regions
TARGET <- newConstraint(
  name     = "TARGET",
  eq       = ">=",
  for.each = data.frame(year = 2050),
  term1    = list(variable = "vTechCap", for.sum = list(region = REG, tech = "LOAD")),
  rhs      = data.frame(year = 2050, rhs = 1.2),
  defVal   = 0
)

draw(LOAD_base) # electricity in, product (PRD) out
```

**3.4 — Siting a base load.** Add `PRD`, `DEM_PRD`, `LOAD_base` and `TARGET` to the
model, solve, and read `vTechNewCap` for `LOAD` by region. A base load runs flat
around the clock — determine which region the model selects, and explain why.

::: {.callout-tip collapse="true" title="Solution 3.4"}
```{r}
repo_base <- newRepository(
  "repo_base",
  ELC, COA, GAS, CO2, SOL, WIN, PRD,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, IMP_GAS, DEM_ELC, DEM_PRD,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC,
  TBD_R1_R2, TBD_R2_R3, LOAD_base, TARGET
)

mod_base  <- newModel(
  name     = "SITE_BASE",
  data     = repo_base,
  region   = REG,
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = hor
)

scen_base <- solve_model(mod_base)

getData(scen_base, "vTechNewCap", tech = "LOAD", merge = TRUE) # region allocation
```
A must-run load needs **firm, cheap power every hour**, so it gravitates to the
region that can supply that most cheaply around the clock — the coal/gas regions
rather than renewables-rich `R3`, whose cheap energy comes and goes with the
weather. Flat demand wants flat supply.
:::

**3.5 — Siting a flexible load.** Now make the same load **flexible**: it may
consume whenever power is cheap, provided it makes its annual `PRD` in total. Set
`af.lo = 0, af.up = 1` and re-solve. Determine whether the model moves it to a
different region.

::: {.callout-tip collapse="true" title="Solution 3.5"}
```{r}
LOAD_flex <- update(
  LOAD_base,
  desc = "New flexible load",
  af   = data.frame(af.lo = 0, af.up = 1) # run any hour, idle any hour
)

repo_flex <- newRepository(
  "repo_flex",
  ELC, COA, GAS, CO2, SOL, WIN, PRD,
  SUP_COA, SUP_GAS, SUP_SOL, SUP_WIN, IMP_GAS, DEM_ELC, DEM_PRD,
  ECOA, EGAS, ESOL, EWIN, WSOL, WWIN, STG_ELC,
  TBD_R1_R2, TBD_R2_R3, LOAD_flex, TARGET
)

mod_flex  <- newModel(
  name     = "SITE_FLEX",
  data     = repo_flex,
  region   = REG,
  discount = WS_DISCOUNT,
  calendar = WS_CAL,
  horizon  = hor
)

scen_flex <- solve_model(mod_flex)

getData(scen_flex, "vTechNewCap", tech = "LOAD", merge = TRUE)
```
Because it can wait for cheap hours, the flexible load is drawn toward **`R3`**: it
absorbs wind and solar that would otherwise be **curtailed** (the surplus of
@sec-curtailment), turning it into product instead of waste. Same target, opposite
siting — flexibility, not the average price, decides where it belongs. This
flexible load is exactly an electrolyser; @sec-processes builds one properly.
:::

## The value of storage and transmission {#sec-value}

The renewables-rich cases hint at it; now measure it. Take a **fully renewable**
system — drop the thermal plants so only wind, solar, storage and wires remain —
and price storage and transmission by **removing** them and reading how much total
cost rises.

```{r}
# renewable-only repository: no ECOA/EGAS, no fossil supply/import
repo_re <- newRepository(
  "repo_re",
  ELC, SOL, WIN, PRD,
  SUP_SOL, SUP_WIN, DEM_ELC, DEM_PRD,
  ESOL, EWIN, WSOL, WWIN, STG_ELC,
  TBD_R1_R2, TBD_R2_R3, LOAD_flex, TARGET
)

# the same system with storage removed, and with transmission removed
repo_re_noSTG <- newRepository(
  "repo_re_noSTG",
  ELC, SOL, WIN, PRD,
  SUP_SOL, SUP_WIN, DEM_ELC, DEM_PRD,
  ESOL, EWIN, WSOL, WWIN,
  TBD_R1_R2, TBD_R2_R3, LOAD_flex, TARGET
)
repo_re_noTR <- newRepository(
  "repo_re_noTR",
  ELC, SOL, WIN, PRD,
  SUP_SOL, SUP_WIN, DEM_ELC, DEM_PRD,
  ESOL, EWIN, WSOL, WWIN, STG_ELC,
  LOAD_flex, TARGET
)
```

**3.6 — What are storage and transmission worth?** Solve the renewable system in
full, then again with storage removed, and again with transmission removed. The
cost increase on removal is the value. **Inspect `vDummyImport`** each time — if it
is non-zero the model is meeting demand with expensive penalty slack, and the
measured "value" is partly that penalty.

::: {.callout-tip collapse="true" title="Solution 3.6"}
```{r}
mod_re     <- newModel("RE_full",  data = repo_re,       region = REG,
                       discount = WS_DISCOUNT, calendar = WS_CAL, horizon = hor)
mod_re_noS <- newModel("RE_noSTG", data = repo_re_noSTG, region = REG,
                       discount = WS_DISCOUNT, calendar = WS_CAL, horizon = hor)
mod_re_noT <- newModel("RE_noTR",  data = repo_re_noTR,  region = REG,
                       discount = WS_DISCOUNT, calendar = WS_CAL, horizon = hor)

scen_re     <- solve_model(mod_re)
scen_re_noS <- solve_model(mod_re_noS)
scen_re_noT <- solve_model(mod_re_noT)

obj_full <- getData(scen_re,     "vObjective", merge = TRUE)$value
obj_noS  <- getData(scen_re_noS, "vObjective", merge = TRUE)$value
obj_noT  <- getData(scen_re_noT, "vObjective", merge = TRUE)$value

obj_noS - obj_full # value of storage: cost increase when removed
obj_noT - obj_full # value of transmission

# is the model quietly buying unmet demand as penalty slack?
getData(scen_re_noS, "vDummyImport", merge = TRUE)
getData(scen_re_noT, "vDummyImport", merge = TRUE)
```
Remove storage and the system must either over-build renewables to cover the
evening or lean on the wire; remove the wire and every region must self-balance its
own weather — either way total cost rises, and that rise is what the missing asset
was worth. Watch `vDummyImport`: if removing an asset leaves demand unservable, the
model quietly buys the shortfall as penalty slack rather than reporting infeasible,
and the cost balloons — a valid ranking of value, but read knowing part of it is
the penalty, not real dispatch. Storage and transmission are partial substitutes:
each buys back the curtailment the other would leave on the table.
:::

## What you built

Space added the questions a single region cannot ask. You assembled a
**multi-region** system from `region`-keyed data, connected it with a fixed
**transmission** link and a **rest-of-world import**, mapped the network with
`plot_trade_map()`, and used it to **site** a new load — discovering that a
*flexible* consumer wants to live where the renewables are while a *firm* one does
not — and to **value** storage and transmission by removing them from a fully
renewable system. @sec-processes turns to the inside of the box: how to model
processes like the electrolyser, a blast furnace, or a vehicle fleet.
@sec-policy takes `mod_M` and puts policy on it.
