---
output: html_document
editor_options:
  chunk_output_type: inline
---

# Electric vehicles: when to charge {#sec-ev}

In @sec-ldv a vehicle was a **technology**: fuel in, passenger-kilometres out,
now. That is the right object for a fleet's *annual* fuel use, and the wrong one
for the question that actually makes electric vehicles interesting — **when** the
electricity is drawn from the grid. A technology consumes in the same timeslice
it produces. A car charges at night and drives in the morning.

Moving energy between timeslices is what a `storage` does, so here the vehicle
**is** the storage:

| | | |
|---|---|---|
| `@input` | grid electricity | the on-board charger, kW |
| `@storage` | the battery | kWh |
| `@output` | kilometres | the motor, km/h |

The whole chapter runs on one day, resolved hourly:

```{r}
hrs <- sprintf("h%02d", 0:23)        # 24 hourly timeslices: h00 ... h23
YF  <- 1 / 365                       # this calendar is ONE DAY out of a year

cal <- newCalendar(
  timetable = make_timetable(struct = list(ANNUAL = "ANNUAL", HOUR = hrs),
                             year_fraction = YF),
  year_fraction = YF, name = "one_day")
```

Saying `year_fraction = 1/365` is what makes a timeslice a real hour: each slice
gets `share = 1/8760`, so `cap.up = 7` below means 7 kW and needs no further
argument. It also sets the slice `weight` to 365, which is why the costs quoted
are a day's bill repeated across the year.

```{r}
car <- newStorage(
  "CAR",
  input    = list(comm = "CHG", cap.up = 7),     # 7 kW on-board charger
  storage  = list(comm = "BAT", cap.up = 60),    # 60 kWh battery
  output   = list(comm = "KM"),                  # kilometres
  seff     = data.frame(inpeff = 0.90, outeff = 6),
  af       = data.frame(cinp.up = 1, cout.up = 1),
  capacity = list(cap.up = 120),                 # motor, km/h
  duration = data.frame(duration.lo = 0, duration.up = 1e6),
  vintage  = data.frame(olife = 10L))
```

Two things in there are worth stopping on.

**`seff` is not a round trip here.** `inpeff = 0.90` is charging loss (grid kWh →
battery kWh) but `outeff = 6` is **six kilometres per kWh** — a conversion
between two different commodities, not an efficiency between 0 and 1. Because
`@input`, `@storage` and `@output` may each name a different commodity, `seff`
became a pair of cross-commodity factors. Only when all three coincide is it the
round-trip efficiency you are used to.

**`duration` is switched off on purpose.** `duration` is the ratio of storing
capacity to output capacity, and it is normally a number of *hours*. Here the
reservoir is in kWh and the motor in km/h — different dimensions — so the ratio
is not an hour count and pinning it would be meaningless. Opening it and bounding
the battery directly with `storage = list(cap.up = 60)` says what we mean.

## 5.1 — Passenger car: the schedule is given {#sec-ev-commute}

*Learning goal:* a storage whose input, stored and output commodities are all
different; charging as a **timing** decision under a price signal.

The driver commutes 30 km at 08:00 and 30 km back at 18:00 — that is fixed. The
grid price is not: cheap before 06:00, expensive between 17:00 and 21:00. The
only decision left is when to charge.

```{r}
price <- ifelse(0:23 < 6, 0.05, ifelse(0:23 >= 17 & 0:23 <= 21, 0.40, 0.15))

m1 <- newModel(
  name = "ev_commute", region = "R1", horizon = newHorizon(2020),
  discount = 0, calendar = cal,          # the one-day calendar above
  repo = newRepository("r1",
    newCommodity("ELC", timeframe = "HOUR"),
    newCommodity("BAT", timeframe = "HOUR"),
    newCommodity("KM",  timeframe = "HOUR"),
    newSupply("GRID", commodity = "ELC",
              supply = data.frame(timeslice = hrs, cost = price)),
    car,
    newDemand("TRIPS", commodity = "KM",
              demand = data.frame(timeslice = c("h08", "h18"), demand = 30))))
```

Solved, this draws **11.11 kWh** — 60 km ÷ 6 km/kWh ÷ 0.90 — and takes all of it
in hours **04 and 05**, the last of the cheap window, for an annual bill of
**202.78** (0.5556 for the day, repeated 365 times). Charging as late as the
cheap tariff allows is not an accident: with nothing else to separate the cheap
hours, any of them would do, and the solver simply reports one optimal answer.

**Exercises.**

**5.1a** Reproduce the numbers. Check the energy chain by hand:
`vStorageInp * inpeff == vStorageOut / outeff`.

**5.1b** Shrink the charger to `cap.up = 3`. How many hours does charging spread
over, and does the bill change? Explain why or why not.

**5.1c** Add a second cheap window at midday (solar). Does the car use it? Add a
`storage = list(cap.up = 15)` small battery and try again — what changed?

**5.1d** The answer above is *degenerate*: several charging schedules cost
exactly the same. Add a tiny per-kWh `varom` and confirm the solution stops
moving between runs. Why is degeneracy worth noticing in a real study?

## 5.2 — Freight van: the schedule is a decision too {#sec-ev-freight}

*Learning goal:* what happens when the service requirement is a **total** rather
than a profile, and how to express that with `newConstraint()`.

A delivery van must cover 240 km in the day. *When* it drives is up to the
operator. Instead of an hourly `newDemand()`, put the requirement on the daily
total and let the optimiser choose both the driving hours and the charging hours:

```{r}
route <- newConstraint(
  name = "ROUTE", eq = ">=",
  for.each = list(region = "R1", year = 2020L),
  term1 = list(variable = "vStorageOut",
               for.sum = list(stg = "VAN", comm = "KM", timeslice = hrs)),
  defVal = 240)
```

The van (22 kW charger, 100 kWh battery) covers its 240 km and charges entirely
in hours **00–05**, at an annual cost of **811.11**. The driving itself spreads across all 24 hours, because
nothing distinguishes them — a *degenerate* schedule again, and a useful prompt.

**Exercises.**

**5.2a** Build it and confirm the 240 km is met and charging sits in the cheap
window.

**5.2b** The driving schedule is degenerate. Add a delivery window — no driving
before 07:00 or after 19:00 — with `af = data.frame(timeslice = <night hours>,
cout.up = 0)`. Now when does it drive, and when does it charge?

**5.2c** Give the van a **depot-only** charger: `cinp.up = 0` except between
22:00 and 06:00. Does the route survive? What is the smallest battery that still
works?

**5.2d** Compare against modelling the same van as a `newTechnology()` with an
annual fuel demand. What question can the storage answer that the technology
cannot?

::: {.callout-note}
## Why a constraint and not a coarse commodity
The tidy way to say "240 km a day, hours free" would be to declare `KM` at a
`DAY` timeframe and let the hourly driving aggregate up. That works for a
**technology** — `eqTechOutTot` has a term that sums finer flows onto a coarser
commodity — but the storage equivalent does not yet exist, so the model comes
out infeasible. Until it does, the constraint above is the way to express it.
:::

## 5.3 — A shared charging post {#sec-ev-shared}

*Learning goal:* shared infrastructure as a **capacity several processes compete
for**, and what congestion costs.

Three cars, one post. Each car keeps its own 7 kW on-board charger — that is
`@input$cap.up`, a property of the vehicle — but all three now draw through a
single `POST` whose rating is the thing under study.

```{r}
post <- newTechnology(
  "POST", input = list(comm = "ELC"), output = list(comm = "CHG"),
  capacity = data.frame(cap.up = post_kw),
  cap2act = 8760,                    # a technology defaults to 1 -> state it
  vintage = data.frame(olife = 10L))
```

Sweeping the post rating, with the three cars needing 33.3 kWh between them:

| post | annual cost | peak draw | hours charging |
|---:|---:|---:|---:|
| 21 kW | 608.33 | 14.00 kW | 3 |
| 10 kW | 608.33 | 10.00 kW | 4 |
| 6 kW | 608.33 | 6.00 kW | 6 |
| 2 kW | **1387.00** | 2.00 kW | 18 |
| 1.2 kW | infeasible | | |

This is the whole story of shared charging in one table. Down to 6 kW the post
costs nothing: there are exactly six cheap hours, and the fleet of three still
fits all its charging inside them. Above that the post stops binding at all — the
cars' own 7 kW chargers become the limit, which is why the peak draw stops rising.
Below 6 kW the fleet cannot fit, charging spills into the expensive evening, and
the bill **more than doubles**. Keep shrinking and the energy cannot be delivered
at all.

**Exercises.**

**5.3a** Reproduce the sweep. Find the rating at which cost first rises — the
point where the post starts binding — and explain why it is 6 kW and not
something else. Then explain the *other* end: why does raising the post above
7 kW change nothing at all?

**5.3b** Give `POST` an `invcost` and let the model *size* it instead of
asserting a rating. Does it choose the knee of the curve? Should it?

**5.3c** Stagger the fleet: give one car a morning commute and another an
evening shift. Does the shared post get cheaper per car? This is the
diversity-factor argument for shared infrastructure — make it quantitative.

**5.3d** Add vehicle-to-grid: let a car discharge back to `ELC` as well as to
`KM`. You will need a second output commodity. Then check whether the model ever
charges *and* discharges in the same hour — and if it does, work out how you
would forbid it. (energyRt has no flag for this: a `@sharedThroughput` slot was
tried and withdrawn, because for a single-commodity store the LP never does it
anyway, and across different commodities the two flows are not in the same units
to begin with.)

## Where this connects

- @sec-ldv modelled vehicles as technologies. Everything there is still true for
  annual fuel accounting; this chapter is about **timing**, which needs storage.
- @sec-processes introduced multi-input/multi-output processes; a storage with
  three different commodities is the same idea applied across time.
- `storage_duration()` in energyRt splits a storage level into how long the
  energy actually sits there — see @sec-duration, where it separates a battery
  that cycles daily from a hydrogen store that holds energy for months. On a car
  it would separate the overnight charge from the buffer that never empties.
