---
output: html_document
editor_options:
  chunk_output_type: inline
---

# Electric vehicles: when to charge {#sec-ev}

In @sec-ldv a vehicle was a **technology**: fuel in, passenger-kilometres out,
now. That is the right object for a fleet's *annual* fuel use, and the wrong one
for the question that actually makes electric vehicles interesting — **when** the
electricity is drawn from the grid. A technology consumes in the same timeslice
it produces. A car charges at night and drives in the morning.

Moving energy between timeslices is what a `storage` does, so here the vehicle
**is** the storage:

| | | |
|---|---|---|
| `@input` | grid electricity | the on-board charger, kW |
| `@storage` | the battery | kWh |
| `@output` | kilometres | the motor, km/h |

```{r}
source("R/workshop-model.R")
```

The whole chapter runs on one day, resolved hourly:

```{r}
hrs <- sprintf("h%02d", 0:23)        # 24 hourly timeslices: h00 ... h23
YF  <- 1 / 365                       # this calendar is ONE DAY out of a year

cal <- newCalendar(
  timetable = make_timetable(struct = list(ANNUAL = "ANNUAL", HOUR = hrs),
                             year_fraction = YF),
  year_fraction = YF, name = "one_day")
```

Saying `year_fraction = 1/365` is what makes a timeslice a real hour: each slice
gets `share = 1/8760`, so `inp.cap.up = 7` below means 7 kW and needs no further
argument. It also sets the slice `weight` to 365, which is why the costs quoted
are a day's bill repeated across the year.

Now the car, one slot at a time:

```{r}
CAR <- newStorage(
  name     = "CAR",
  input    = list(comm = "ELC"),                 # what charges it
  storage  = list(comm = "BAT"),                 # what it holds
  output   = list(comm = "KM"),                  # what it delivers
  seff     = data.frame(inpeff = 0.90, outeff = 6),
  af       = data.frame(timeslice = hrs, inp.af.up = 1),
  capacity = list(inp.cap.up = 7,                # 7 kW on-board charger
                  stg.cap.up = 60,               # 60 kWh battery
                  out.cap.up = 120),             # motor, km/h
  duration = data.frame(duration.lo = 0, duration.up = 1e6),
  vintage  = data.frame(olife = 10L))
```

**The three roles declare; `@capacity` sizes.** `@input`, `@storage` and
`@output` say *what* each side handles. Every capacity lives in `@capacity`
under a prefix naming its part — `inp.` the charger, `stg.` the battery, `out.`
the motor — which is why three numbers in three different units (kW, kWh, km/h)
sit together in one place. A car is the clearest case there is: the thing that
fills it, the thing that holds it and the thing that empties it are three
different pieces of hardware with three different ratings.

Three things in there are worth stopping on.

**`seff` is not a round trip here.** `inpeff = 0.90` is charging loss (grid kWh →
battery kWh) but `outeff = 6` is **six kilometres per kWh** — a conversion
between two different commodities, not an efficiency between 0 and 1. Because
`@input`, `@storage` and `@output` may each name a different commodity, `seff`
became a pair of cross-commodity factors. Only when all three coincide is it the
round-trip efficiency you are used to.

**The `af` rows are what make 7 kW an hourly rating.** A part capacity with no
sub-annual availability row limits the part's **annual throughput** only — the
model could then push the whole charge through one hour. Listing the hours in
`@af` puts the bound where the hardware is: at most `inp.cap.up × 1` kWh may
flow in *each* hour.

**`duration` is switched off on purpose.** `duration` is the ratio of storing
capacity to output capacity, and it is normally a number of *hours*. Here the
reservoir is in kWh and the motor in km/h — different dimensions — so the ratio
is not an hour count and pinning it would be meaningless. Opening it and bounding
the battery directly with `capacity = list(stg.cap.up = 60)` says what we mean.
(Try deleting the `duration` line: the default ties the battery to the motor
rating and the bill jumps from 202.78 to 324.44.)

## 5.1 — Passenger car: the schedule is given {#sec-ev-commute}

*Learning goal:* a storage whose input, stored and output commodities are all
different; charging as a **timing** decision under a price signal.

The driver commutes 30 km at 08:00 and 30 km back at 18:00 — that is fixed. The
grid price is not: cheap before 06:00, expensive between 17:00 and 21:00. The
only decision left is when to charge.

Build the model the way every chapter has: commodities first, then the
processes, then a repository, then the model.

```{r}
ELC <- newCommodity(name = "ELC", desc = "Grid electricity", timeframe = "HOUR")
BAT <- newCommodity(name = "BAT", desc = "Battery charge",   timeframe = "HOUR")
KM  <- newCommodity(name = "KM",  desc = "Kilometres",       timeframe = "HOUR")
```

```{r}
price <- ifelse(0:23 < 6, 0.05, ifelse(0:23 >= 17 & 0:23 <= 21, 0.40, 0.15))

GRID <- newSupply(
  name      = "GRID",
  desc      = "Grid electricity at an hourly tariff",
  commodity = "ELC",
  supply    = data.frame(timeslice = hrs, cost = price))

TRIPS <- newDemand(
  name      = "TRIPS",
  desc      = "The commute: 30 km out, 30 km back",
  commodity = "KM",
  demand    = data.frame(timeslice = c("h08", "h18"), demand = 30))
```

```{r}
repo_COMMUTE <- newRepository("repo_COMMUTE", ELC, BAT, KM, GRID, CAR, TRIPS)

mod_COMMUTE <- newModel(
  name     = "EV_COMMUTE",
  desc     = "One car, a fixed commute, an hourly tariff",
  data     = repo_COMMUTE,
  region   = "R1",
  horizon  = newHorizon(2020),
  discount = 0,
  calendar = cal)

scen_COMMUTE <- interpolate_model(mod_COMMUTE, name = "COMMUTE") |>
  solve_scenario(echo = FALSE)
```

Read the answer straight from the solution:

```{r}
getData(scen_COMMUTE, "vObjective", merge = TRUE)$value      # annual bill

getData(scen_COMMUTE, "vStorageInp", merge = TRUE, timeframe = "highest") |>
  filter(value > 1e-9) |>
  select(timeslice, value)                                   # when it charges
```

The car draws **11.11 kWh** — 60 km ÷ 6 km/kWh ÷ 0.90 — entirely inside the
cheap window, for an annual bill of **202.78** (0.5556 for the day, repeated
365 times). *How* the 11.11 kWh is split across the six cheap hours is
**degenerate**: every split costs the same, so the solver reports one optimal
answer among many — here an even 1.85 kWh per hour, on another machine perhaps
two full-power hours. Exercise 5.1d makes this point explicit.

**Exercises.**

**5.1a** Reproduce the numbers. Check the energy chain by hand:
`vStorageInp * inpeff == vStorageOut / outeff`.

**5.1b** Shrink the charger: rebuild `CAR` with `inp.cap.up = 2`, then `1.8`.
(Replace it in the model with `add(mod_COMMUTE, CAR, overwrite = TRUE)`.)
At 2 kW the bill is unchanged — why? At 1.8 kW it rises to **214.13** — where
did the extra charging go, and why exactly there?

**5.1c** Add a second cheap window at midday (solar). Does the car use it? Set
`stg.cap.up = 15` and try again — what changed?

**5.1d** Add a tiny per-kWh `varom` to the storage and confirm the charging
schedule stops moving between runs. Why is degeneracy worth noticing in a real
study?

## 5.2 — Freight van: the schedule is a decision too {#sec-ev-freight}

*Learning goal:* what happens when the service requirement is a **total** rather
than a profile, and how to express that with `newConstraint()`.

A delivery van must cover 240 km in the day. *When* it drives is up to the
operator. First the van — a bigger charger and battery, same three roles:

```{r}
VAN <- newStorage(
  name     = "VAN",
  input    = list(comm = "ELC"),
  storage  = list(comm = "BAT"),
  output   = list(comm = "KM"),
  seff     = data.frame(inpeff = 0.90, outeff = 6),
  af       = data.frame(timeslice = hrs, inp.af.up = 1),
  capacity = list(inp.cap.up = 22,               # depot charger, kW
                  stg.cap.up = 100,              # kWh
                  out.cap.up = 60),              # km/h
  duration = data.frame(duration.lo = 0, duration.up = 1e6),
  vintage  = data.frame(olife = 10L))
```

Instead of an hourly `newDemand()`, put the requirement on the daily total and
let the optimiser choose both the driving hours and the charging hours:

```{r}
ROUTE <- newConstraint(
  name     = "ROUTE", eq = ">=",
  for.each = list(region = "R1", year = 2020L),
  term1    = list(variable = "vStorageOut",
                  for.sum = list(stg = "VAN", comm = "KM", timeslice = hrs)),
  defVal   = 240)
```

The model reuses the commodities and the grid from 5.1 — a repository is just a
bag of objects, so build a new one from the same pieces:

```{r}
repo_FREIGHT <- newRepository("repo_FREIGHT", ELC, BAT, KM, GRID, VAN)

mod_FREIGHT <- newModel(
  name     = "EV_FREIGHT",
  desc     = "A van with a daily route total",
  data     = repo_FREIGHT,
  region   = "R1",
  horizon  = newHorizon(2020),
  discount = 0,
  calendar = cal)

scen_FREIGHT <- interpolate_model(mod_FREIGHT, name = "FREIGHT", ROUTE) |>
  solve_scenario(echo = FALSE)

getData(scen_FREIGHT, "vObjective", merge = TRUE)$value
```

(Note where `ROUTE` went: a constraint is layered onto a scenario at
`interpolate_model()` — the same pattern the UTOPIA kits use for their policy
levers.)

The van covers its 240 km and charges entirely inside the cheap window, at an
annual cost of **811.11**. The driving itself is spread across the day, because
nothing distinguishes the hours — a *degenerate* schedule again, and a useful
prompt.

**Exercises.**

**5.2a** Build it and confirm the 240 km is met
(`getData(scen_FREIGHT, "vStorageOut", ...)`) and charging sits in the cheap
window.

**5.2b** The driving schedule is degenerate. Add a delivery window — no driving
before 07:00 or after 19:00 — with a second `newConstraint()` (`eq = "<="`,
`defVal = 0`) summing `vStorageOut` over the night hours. Now when does it
drive, and when does it charge?

**5.2c** Give the van a **depot-only** charger: `inp.af.up = 0` except between
22:00 and 06:00. Does the route survive? What is the smallest battery that still
works?

**5.2d** Compare against modelling the same van as a `newTechnology()` with an
annual fuel demand. What question can the storage answer that the technology
cannot?

::: {.callout-note}
## Why a constraint and not a coarse commodity
The tidy way to say "240 km a day, hours free" would be to declare `KM` at a
`DAY` timeframe and let the hourly driving aggregate up. That works for a
**technology** — `eqTechOutTot` has a term that sums finer flows onto a coarser
commodity — but the storage equivalent does not yet exist, so the model comes
out infeasible. Until it does, the constraint above is the way to express it.
:::

## 5.3 — A shared charging post {#sec-ev-shared}

*Learning goal:* shared infrastructure as a **capacity several processes compete
for**, and what congestion costs.

Three cars, one post. Each car keeps its own 7 kW on-board charger — that is a
property of the vehicle — but all three now draw through a single `POST` whose
rating is the thing under study. The post is a plain technology converting grid
electricity into "electricity at the post", a new commodity the cars charge
from:

```{r}
CHG <- newCommodity(name = "CHG", desc = "Electricity at the post",
                    timeframe = "HOUR")

POST <- newTechnology(
  name     = "POST",
  desc     = "The shared charging post",
  input    = list(comm = "ELC"),
  output   = list(comm = "CHG"),
  capacity = data.frame(cap.up = 6),   # the rating under study, kW
  cap2act  = 8760,                     # a technology defaults to 1 -> state it
  vintage  = data.frame(olife = 10L))
```

Each car needs its own kilometres — three commodities, three cars, three
commutes. The cars are the `CAR` of 5.1 with two changes: they charge from
`CHG`, and each delivers its own `KM_*`:

```{r}
KM_A <- newCommodity(name = "KM_A", timeframe = "HOUR")
KM_B <- newCommodity(name = "KM_B", timeframe = "HOUR")
KM_C <- newCommodity(name = "KM_C", timeframe = "HOUR")

CAR_A <- newStorage(
  name     = "CAR_A",
  input    = list(comm = "CHG"),
  storage  = list(comm = "BAT"),
  output   = list(comm = "KM_A"),
  seff     = data.frame(inpeff = 0.90, outeff = 6),
  af       = data.frame(timeslice = hrs, inp.af.up = 1),
  capacity = list(inp.cap.up = 7, stg.cap.up = 60, out.cap.up = 120),
  duration = data.frame(duration.lo = 0, duration.up = 1e6),
  vintage  = data.frame(olife = 10L))
CAR_B <- update(CAR_A, name = "CAR_B", output = list(comm = "KM_B"))
CAR_C <- update(CAR_A, name = "CAR_C", output = list(comm = "KM_C"))

TRIPS_A <- newDemand(name = "TRIPS_A", commodity = "KM_A",
                     demand = data.frame(timeslice = c("h08", "h18"), demand = 30))
TRIPS_B <- update(TRIPS_A, name = "TRIPS_B", commodity = "KM_B")
TRIPS_C <- update(TRIPS_A, name = "TRIPS_C", commodity = "KM_C")
```

```{r}
repo_SHARED <- newRepository("repo_SHARED",
  ELC, CHG, BAT, KM_A, KM_B, KM_C,
  GRID, POST, CAR_A, CAR_B, CAR_C,
  TRIPS_A, TRIPS_B, TRIPS_C)

mod_SHARED <- newModel(
  name     = "EV_SHARED",
  desc     = "Three cars behind one post",
  data     = repo_SHARED,
  region   = "R1",
  horizon  = newHorizon(2020),
  discount = 0,
  calendar = cal)

scen_SHARED <- interpolate_model(mod_SHARED, name = "SHARED") |>
  solve_scenario(echo = FALSE)

getData(scen_SHARED, "vObjective", merge = TRUE)$value
```

To sweep the rating, swap the post and re-solve — `add()` with
`overwrite = TRUE` replaces an object of the same name wherever it sits:

```{r}
POST2 <- update(POST, capacity = data.frame(cap.up = 2))

scen_SHARED2 <- add(mod_SHARED, POST2, overwrite = TRUE) |>
  interpolate_model(name = "SHARED2") |>
  solve_scenario(echo = FALSE)

getData(scen_SHARED2, "vObjective", merge = TRUE)$value
```

Sweeping the post rating, with the three cars needing 33.3 kWh between them:

| post | annual cost | hours charging |
|---:|---:|---:|
| 21 kW | 608.33 | 6 |
| 10 kW | 608.33 | 6 |
| 6 kW | 608.33 | 6 |
| 2 kW | **1387.00** | 19 |
| 1.2 kW | infeasible | |

This is the whole story of shared charging in one table. Down to 6 kW the post
costs nothing: there are exactly six cheap hours, and the fleet of three still
fits all its 33.3 kWh inside them (6 kW × 6 h = 36 kWh). Below 6 kW the fleet
cannot fit, charging spills into the expensive hours, and the bill **more than
doubles**. Keep shrinking and the energy cannot be delivered at all: at 1.2 kW
even 24 hours of charging fall short, and the model is infeasible.

**Exercises.**

**5.3a** Reproduce the sweep with `add(..., overwrite = TRUE)` as above. Find
the rating at which cost first rises — the point where the post starts binding —
and explain why it is 6 kW and not something else.

**5.3b** Give `POST` an `invcost` and let the model *size* it instead of
asserting a rating (drop `cap.up`). Does it choose the knee of the curve?
Should it?

**5.3c** Stagger the fleet: give one car a morning commute and another an
evening shift. Does the shared post get cheaper per car? This is the
diversity-factor argument for shared infrastructure — make it quantitative.

**5.3d** Add vehicle-to-grid: let a car discharge back to `ELC` as well as to
`KM`. You will need a second output commodity. Then check whether the model ever
charges *and* discharges in the same hour — and if it does, work out how you
would forbid it. (energyRt has no flag for this: a `@sharedThroughput` slot was
tried and withdrawn, because for a single-commodity store the LP never does it
anyway, and across different commodities the two flows are not in the same units
to begin with.)

## Where this connects

- @sec-ldv modelled vehicles as technologies. Everything there is still true for
  annual fuel accounting; this chapter is about **timing**, which needs storage.
- @sec-processes introduced multi-input/multi-output processes; a storage with
  three different commodities is the same idea applied across time.
- `storage_duration()` in energyRt splits a storage level into how long the
  energy actually sits there — see @sec-duration, where it separates a battery
  that cycles daily from a hydrogen store that holds energy for months. On a car
  it would separate the overnight charge from the buffer that never empties.
