---
output: html_document
editor_options:
  chunk_output_type: inline
---

# Life-cycle analysis {#sec-lca}

Every chapter so far asked one question of a technology: *what does it cost to
run?* A life-cycle question is different. It asks what the technology costs the
world **before** it runs, **while** it stands, and **after** it stops — land
occupied, steel poured, CO₂ embodied, material recovered.

None of that needs a new kind of object. It all rides on **auxiliary flows**
(@sec-processes): a commodity linked to a process by a coefficient named
`<driver>2a<inp|out>`. What makes it life-cycle analysis is *which driver*:

| coefficient | driver | fires | shape in time |
|---|---|---|---|
| `cap2ainp` / `cap2aout` | **standing** capacity | every year the plant exists | a **plateau** |
| `ncap2ainp` / `ncap2aout` | **new** capacity | the milestone it is built | a **pulse** |
| `pho2ainp` / `pho2aout` | capacity **reaching end of life** | the milestone it ages out | a pulse, once |
| `ret2ainp` / `ret2aout` | capacity **retired early** | the milestone it is scrapped | a pulse, once |

The whole chapter is one wind farm seen through those four lenses. Nothing here
is bigger than a page of code, and everything solves in under a second.

```{r}
source("R/workshop-model.R")
library(tidyr)     # pivot_wider(), for the quantity x milestone tables
```

A single annual timeslice, and six milestones over 2020–2045:

```{r}
cal <- newCalendar(timetable = make_timetable(struct = list(ANNUAL = "ANNUAL")),
                   name = "lca_annual")
hor <- newHorizon(2020:2045, intervals = c(1, 5, 5, 5, 5, 5))
YRS <- hor@intervals$mid
YRS
hor@intervals$end - hor@intervals$start + 1   # period lengths, in years
```

::: {.callout-important appearance="simple"}
**The first period is one year long, the rest are five.** That is deliberate.
Almost every quantity below is a **per-year rate**, and unequal periods are the
only way to see the difference: a phase-out of `4` in a five-year period means
**20** units of capacity left, not 4. Equal periods hide the distinction, and
hiding it is how rate/total mistakes survive.
:::

One reading pattern serves the whole chapter, and it is nothing but the API you
already know: `getData()` accepts several variable names at once and returns
them stacked with a `name` column, and `pivot_wider()` turns that into the
*quantity × milestone* table every section reads. The same five-line pipe
appears after every solve — deliberately, so that by the end of the chapter it
is muscle memory rather than magic:

```r
getData(scen, name = c("vTechCap", ...), merge = TRUE, drop.zeros = FALSE) |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

`drop.zeros = FALSE` is not decoration. A schedule that runs down to **nothing**
records a zero, and a zero dropped is a fleet that never leaves.

## 6.1 — Retiring on a schedule {#sec-lca-schedule}

*Learning goal:* what `capacity$stock` actually means, and the two parameters
energyRt derives from it.

`stock` is the **legacy fleet still standing** at each milestone — plant that
exists because it was built before the model starts. Give it a falling path and
you have declared a decommissioning timetable:

```{r}
FLEET <- data.frame(year  = c(2020L, 2028L, 2033L, 2038L, 2043L),
                    stock = c(100,   100,   60,    20,    0))
```

The farm occupies land while it stands — `cap2ainp`, the plateau coefficient —
at 0.2 ha per MW. Objects first:

```{r}
WND  <- newCommodity(name = "WND",  desc = "Wind resource",   timeframe = "ANNUAL")
ELC  <- newCommodity(name = "ELC",  desc = "Electricity",     timeframe = "ANNUAL")
LAND <- newCommodity(name = "LAND", desc = "Occupied land",   timeframe = "ANNUAL")

SWND  <- newSupply(name = "SWND",  commodity = "WND",
                   supply = data.frame(cost = 0))
SLAND <- newSupply(name = "SLAND", commodity = "LAND",
                   supply = data.frame(cost = 0))

WIND <- newTechnology(
  name    = "WIND",
  input   = list(comm = "WND"),
  output  = list(comm = "ELC"),
  cap2act = 1,
  afs     = data.frame(afs.up = 1),
  capacity = FLEET,
  invcost = data.frame(invcost = 1),
  vintage = data.frame(olife = 40L),
  aux     = data.frame(acomm = "LAND", unit = "ha"),
  aeff    = data.frame(acomm = "LAND", cap2ainp = 0.2))

DEM <- newDemand(name = "DEM", commodity = "ELC",
                 demand = data.frame(demand = 5))
```

Then the repository, the model, and the solve:

```{r}
repo_FLEET <- newRepository("repo_FLEET",
                            WND, ELC, LAND, SWND, SLAND, WIND, DEM)

mod_FLEET <- newModel(
  name     = "LC_FLEET",
  desc     = "A wind fleet retiring on a schedule",
  data     = repo_FLEET,
  region   = "R1",
  horizon  = hor,
  discount = 0,
  calendar = cal)

s1 <- interpolate_model(mod_FLEET, name = "lc_fleet") |>
  solve_scenario(echo = FALSE)
```

```{r}
getData(s1, name = c("vTechCap", "vTechStockCap", "vTechStockPhaseOut",
                     "vTechNewCap", "vTechAInp"),
        merge = TRUE, drop.zeros = FALSE) |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

Read the rows against each other:

- **`vTechStockCap`** is the legacy fleet: `100 100 100 60 20 0` — exactly the
  schedule.
- **`vTechStockPhaseOut`** is what *left*: `8` per year in 2033 and 2038, `4` in
  2043. A per-year rate again — ×5 years gives 40, 40 and 20, and
  40 + 40 + 20 = 100, the whole fleet.
- **`vTechCap`** tracks the fleet until 2043, when the schedule reaches zero and
  the model has to build 1 MW/yr to keep serving the 5 MW demand.
- **`vTechAInp[LAND]`** is `0.2 × vTechCap`, to the last decimal. **Land is not
  released by an event.** The draw simply stops when the capacity does.
  Occupation is a stock expressed as a recurring flow, and that is the whole idea
  behind `cap2ainp`.

### What the model derived from your schedule

`stock` carries two different meanings — *what was originally there* and *what is
still there* — and energyRt splits them into two parameters before solving.
Interpolation alone (no solve) is enough to see them:

```{r}
sc1 <- interpolate_model(mod_FLEET, name = "lc_par")

getData(sc1, name = c("pTechStock", "pTechStockNew", "pTechStockSurv"),
        merge = TRUE) |>
  select(name, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value) |>
  arrange(name)
```

- **`pTechStockNew`** — capacity *commissioned* exogenously: the whole 100 MW in
  the first year, and nothing after, because this schedule only falls.
- **`pTechStockSurv`** — the share of the previous milestone's fleet the schedule
  keeps: `60/100 = 0.6`, then `20/60 = 0.333`, then `0/20 = 0`. Its default is
  **1** ("nothing scheduled to leave"), which is why the early years are simply
  **absent** (`NA`) from the stored parameter — an absent tuple reads as the
  default.

Both use only **neighbouring** milestones. That is what makes them invariant when
you move the model's base year — a share of the *first* year's stock would not be.

**Exercises.**

**6.1a** Change the schedule to `c(100, 100, 60, 20, 20)` — a fleet that stops
retiring rather than disappearing. Predict `pTechStockSurv` in 2043 before running.

**6.1b** Make the fleet *grow*: `stock = c(100, 100, 140, 140, 140)`. Which
parameter picks up the increase, and what does `vTechStockPhaseOut` do? Explain
why a rising `stock` must not read as a negative phase-out.

**6.1c** Move the base year to 2025 and re-run. `pTechStockSurv` should not move.
Why does that matter for a study that gets re-based every few years?

## 6.2 — Land competition, and what a hectare is worth {#sec-lca-rent}

*Learning goal:* an aux commodity as a **shared budget**, and reading a shadow
price without a solver's dual.

Two technologies now compete for one landscape. Wind is cheaper to build but
land-hungry (0.5 ha/MW); solar is dearer and land-light (0.1 ha/MW). The
`ava.up` on the land supply is the budget:

```{r}
SUN  <- newCommodity(name = "SUN", desc = "Solar resource", timeframe = "ANNUAL")
SSUN <- newSupply(name = "SSUN", commodity = "SUN", supply = data.frame(cost = 0))

SLAND50 <- newSupply(name = "SLAND", commodity = "LAND",
                     supply = data.frame(ava.up = 50, cost = 0))

WIND2 <- newTechnology(
  name = "WIND", input = list(comm = "WND"), output = list(comm = "ELC"),
  cap2act = 1, afs = data.frame(afs.up = 1),
  invcost = data.frame(invcost = 100), vintage = data.frame(olife = 40L),
  aux  = data.frame(acomm = "LAND", unit = "ha"),
  aeff = data.frame(acomm = "LAND", cap2ainp = 0.5))

SOLAR <- newTechnology(
  name = "SOLAR", input = list(comm = "SUN"), output = list(comm = "ELC"),
  cap2act = 1, afs = data.frame(afs.up = 1),
  invcost = data.frame(invcost = 160), vintage = data.frame(olife = 40L),
  aux  = data.frame(acomm = "LAND", unit = "ha"),
  aeff = data.frame(acomm = "LAND", cap2ainp = 0.1))

DEM100 <- newDemand(name = "DEM", commodity = "ELC",
                    demand = data.frame(demand = 100))

repo_LAND <- newRepository("repo_LAND", WND, SUN, ELC, LAND,
                           SWND, SSUN, SLAND50, WIND2, SOLAR, DEM100)

mod_LAND <- newModel(
  name = "LC_LAND", data = repo_LAND, region = "R1",
  horizon = hor, discount = 0, calendar = cal)

s_50 <- interpolate_model(mod_LAND, name = "lc_land50") |>
  solve_scenario(echo = FALSE)
```

To tighten the budget, swap the land supply and re-solve — `add()` with
`overwrite = TRUE` replaces the object of the same name, and `update()` gives
you a copy with one slot changed:

```{r}
s_40 <- add(mod_LAND, update(SLAND50, supply = data.frame(ava.up = 40, cost = 0)),
            overwrite = TRUE) |>
  interpolate_model(name = "lc_land40") |> solve_scenario(echo = FALSE)
s_31 <- add(mod_LAND, update(SLAND50, supply = data.frame(ava.up = 31, cost = 0)),
            overwrite = TRUE) |>
  interpolate_model(name = "lc_land31") |> solve_scenario(echo = FALSE)
s_30 <- add(mod_LAND, update(SLAND50, supply = data.frame(ava.up = 30, cost = 0)),
            overwrite = TRUE) |>
  interpolate_model(name = "lc_land30") |> solve_scenario(echo = FALSE)
```

`getData()` takes a **named list of scenarios** and adds a `scenario` column,
so the comparison is one call:

```{r}
runs <- list(L50 = s_50, L40 = s_40, L31 = s_31, L30 = s_30)

getData(runs, name = "vObjective", merge = TRUE) |>
  select(scenario, objective = value)

getData(runs, name = "vTechCap", merge = TRUE) |>
  filter(year == 2043) |>
  group_by(scenario, tech) |>
  summarise(MW = sum(value), .groups = "drop") |>
  pivot_wider(names_from = tech, values_from = MW)
```

At 50 ha the constraint is slack and the model builds **all wind**. Tighten it
and solar is bought in, one hectare at a time, purely to save land.

**The shadow price is the slope.** energyRt exports no duals here, so price the
constraint the way it is *defined* — solve twice and difference:

```{r}
obj <- getData(runs, name = "vObjective", merge = TRUE)
o <- setNames(obj$value, obj$scenario)
c(one_ha = o[["L30"]] - o[["L31"]],
  ten_ha = (o[["L30"]] - o[["L40"]]) / 10)
```

**97.5 per hectare**, from a one-hectare step and from a ten-hectare step alike.
That agreement is the check that matters: it says the answer is on one linear
facet. Where the two disagree, you have crossed a kink and the "shadow price" is
not a single number at all.

**Exercises.**

**6.2a** Find the limit at which wind disappears entirely. Is the rent the same
there?

**6.2b** Step the limit from 50 down to 10 in fives and plot objective against
limit. Where are the kinks, and what switches at each?

**6.2c** A one-hectare step is a *finite difference*, not a derivative. Construct
a case where the two estimates above disagree, and say which one you would quote.

## 6.3 — Four life stages, one technology {#sec-lca-stages}

*Learning goal:* the plateau and the pulse side by side — the conceptual spine of
the whole aux system.

Now the demand **grows**, so the model has to build, and all four coefficients
fire on the same object:

```{r}
STEEL <- newCommodity(name = "STEEL", timeframe = "ANNUAL")
CO2E  <- newCommodity(name = "CO2E",  timeframe = "ANNUAL")
SCRAP <- newCommodity(name = "SCRAP", timeframe = "ANNUAL")
SSTEEL <- newSupply(name = "SSTEEL", commodity = "STEEL",
                    supply = data.frame(cost = 0))

WIND4 <- newTechnology(
  name = "WIND",
  input = list(comm = "WND"), output = list(comm = "ELC"), cap2act = 1,
  afs = data.frame(afs.up = 1), invcost = data.frame(invcost = 1),
  vintage = data.frame(olife = 15L),
  aux  = data.frame(acomm = c("LAND", "STEEL", "CO2E", "SCRAP"),
                    unit  = c("ha",   "t",     "t",    "t")),
  aeff = data.frame(
    acomm     = c("LAND", "STEEL", "CO2E", "SCRAP"),
    cap2ainp  = c(0.2,    NA,      NA,     NA),   # land, while it stands
    ncap2ainp = c(NA,     150,     NA,     NA),   # steel, at construction
    ncap2aout = c(NA,     NA,      700,    NA),   # embodied CO2, at construction
    pho2aout  = c(NA,     NA,      NA,     120))) # scrap, at end of life

DEMG <- newDemand(name = "DEM", commodity = "ELC",
                  demand = data.frame(year = YRS,
                                      demand = c(5, 20, 40, 60, 80, 100)))

repo_LIFE <- newRepository("repo_LIFE", WND, ELC, LAND, STEEL, CO2E, SCRAP,
                           SWND, SLAND, SSTEEL, WIND4, DEMG)

mod_LIFE <- newModel(
  name = "LC_LIFE", data = repo_LIFE, region = "R1",
  horizon = hor, discount = 0, calendar = cal)

s4 <- interpolate_model(mod_LIFE, name = "lc_life") |>
  solve_scenario(echo = FALSE)

getData(s4, name = c("vTechCap", "vTechNewCap", "vTechPhaseOut",
                     "vTechAInp", "vTechAOut"),
        merge = TRUE, drop.zeros = FALSE) |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

Same machinery, opposite shapes:

- **LAND** climbs monotonically with the fleet — `1, 4, 8, 12, 16, 20` = 0.2 ×
  capacity. A **plateau**, present in every year the plant exists.
- **STEEL** tracks *building*, not standing — `750, 450, 600, 600, 1200, 1200`
  = 150 × `vTechNewCap`. A **pulse**: it drops in 2023 even though the fleet
  grew, because the build *rate* fell.
- **CO₂E** is the same pulse at 700 t/MW: embodied carbon is paid once, at
  construction.
- **SCRAP** appears only from 2038, when the 2020 and 2023 vintages reach their
  15-year life: `4` MW/yr aging out × 120 t/MW = `480` t/yr.

Read the first column carefully. 2020 is a **one-year** period, so `750` is
750 t. 2038 is a **five-year** period, so `1200` is 1200 t **per year** — 6,000 t
in the period. Every one of these numbers is a rate.

**Exercises.**

**6.3a** Confirm `vTechAInp[STEEL] == 150 * vTechNewCap` and
`vTechAInp[LAND] == 0.2 * vTechCap` for every milestone.

**6.3b** Compute total steel over the horizon in tonnes, weighting each milestone
by its period length. Compare with the naive column sum. How large is the error?

**6.3c** Shorten `olife` to 10 and re-run. Which rows move, and which do not?

## 6.4 — Embodied against operational {#sec-lca-payback}

*Learning goal:* when a low-carbon plant repays the carbon it was built with.

Two ways to serve 438 GWh a year. Wind carries 700 t CO₂/MW of embodied
emissions and nothing while running; gas carries 30 t/MW embodied and
0.35 t/MWh **every hour it runs**. `cap2act = 8760` makes activity read in MWh,
so `act2aout` is per MWh. The two models share everything except the
technology:

```{r}
FUEL <- newCommodity(name = "FUEL", timeframe = "ANNUAL")
SF   <- newSupply(name = "SF", commodity = "FUEL", supply = data.frame(cost = 0))
DEM438 <- newDemand(name = "DEM", commodity = "ELC",
                    demand = data.frame(demand = 438000))

WINDX <- newTechnology(
  name = "WIND", input = list(comm = "FUEL"), output = list(comm = "ELC"),
  cap2act = 8760, afs = data.frame(afs.up = 0.5),
  vintage = data.frame(olife = 40L), invcost = data.frame(invcost = 1),
  aux  = data.frame(acomm = "CO2E", unit = "t"),
  aeff = data.frame(acomm = "CO2E", ncap2aout = 700, act2aout = 0))

GASX <- update(WINDX, name = "GAS",
               aeff = data.frame(acomm = "CO2E", ncap2aout = 30,
                                 act2aout = 0.35))

repo_WINDX <- newRepository("repo_WINDX", FUEL, ELC, CO2E, SF, WINDX, DEM438)
repo_GASX  <- newRepository("repo_GASX",  FUEL, ELC, CO2E, SF, GASX,  DEM438)

sw <- newModel(name = "LC_CX", data = repo_WINDX, region = "R1",
               horizon = hor, discount = 0, calendar = cal) |>
  interpolate_model(name = "lc_cx_wind") |> solve_scenario(echo = FALSE)
sg <- newModel(name = "LC_CX", data = repo_GASX, region = "R1",
               horizon = hor, discount = 0, calendar = cal) |>
  interpolate_model(name = "lc_cx_gas") |> solve_scenario(echo = FALSE)

getData(list(WIND = sw, GAS = sg), name = "vTechAOut", merge = TRUE,
        drop.zeros = FALSE) |>
  filter(comm == "CO2E") |>
  select(scenario, year, value) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0)
```

Wind emits **35,000 t in 2020 and nothing ever again** — 50 MW × 700 t/MW, once.
Gas emits **153,300 t every single year**, for as long as it runs.

```{r}
pulse <- getData(sw, "vTechAOut", merge = TRUE) |>
  filter(comm == "CO2E", year == 2020) |> pull(value)
rate <- getData(sg, "vTechAOut", merge = TRUE) |>
  filter(comm == "CO2E", year == 2043) |> pull(value)
c(payback_years = pulse / rate, payback_months = 12 * pulse / rate)
```

**Two point seven months.** The entire embodied burden of the wind farm is repaid
before its first summer is out. That is the number the "but what about the
concrete?" argument is usually missing, and it falls out of two coefficients on
two technologies.

Note gas's 2020 figure is 154,800, not 153,300: the extra 1,500 t is its *own*
embodied pulse, 50 MW × 30 t/MW, landing in the same year. Both technologies
carry both kinds of emission. They differ only in the ratio.

**Exercises.**

**6.4a** How high would wind's embodied intensity have to be for the payback to
reach one year? Is that a plausible number?

**6.4b** Add a carbon price and let the model choose between the two. At what
price does it switch? Does the embodied pulse affect the switching point at all?

**6.4c** Cumulate both series over the horizon, weighting by period length, and
plot the crossover.

## 6.5 — Scrap tells you *why* it stopped {#sec-lca-scrap}

*Learning goal:* the same physical event, two different coefficients, because the
material recovered is genuinely different.

A plant that runs to the end of its life is worn out; what comes back is low-grade
and modest. A plant scrapped **early** is largely intact, and what comes back is
worth more. That physical difference is why there are two coefficient pairs and
not one:

```{r}
WIND_S <- newTechnology(
  name = "WIND",
  input = list(comm = "WND"), output = list(comm = "ELC"), cap2act = 1,
  afs = data.frame(afs.up = 1), capacity = FLEET,
  fixom = data.frame(fixom = 8), invcost = data.frame(invcost = 1000),
  vintage = data.frame(olife = 18L), optimizeRetirement = TRUE,
  aux  = data.frame(acomm = "SCRAP", unit = "t"),
  aeff = data.frame(acomm = "SCRAP",
                    pho2aout = 120,    # worn out at end of life
                    ret2aout = 200))   # intact, retired early

repo_SCRAP <- newRepository("repo_SCRAP", WND, ELC, SCRAP, SWND, WIND_S, DEM)
```

The fleet costs 8/MW/yr to keep and only 5 MW of it is needed. Retiring early
should obviously pay. The model-level switch decides whether the optimiser is
*allowed* to; two models share one repository:

```{r}
mod_OFF <- newModel(name = "LC_SCRAP", data = repo_SCRAP, region = "R1",
                    horizon = hor, discount = 0, calendar = cal,
                    optimizeRetirement = FALSE)
mod_ON  <- newModel(name = "LC_SCRAP", data = repo_SCRAP, region = "R1",
                    horizon = hor, discount = 0, calendar = cal,
                    optimizeRetirement = TRUE)

s6a <- interpolate_model(mod_OFF, name = "lc_scrap_off") |>
  solve_scenario(echo = FALSE)
getData(s6a, "vObjective", merge = TRUE)$value

getData(s6a, name = c("vTechCap", "vTechStockCap", "vTechRetiredStock",
                      "vTechStockPhaseOut", "vTechAOut"),
        merge = TRUE, drop.zeros = FALSE) |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

95 MW sits idle for two decades because nobody is allowed to close it, and the
only scrap is the schedule's own: `8 × 120 = 960` t/yr in 2033 and 2038.

Now the run with retirement **on** — same repository, same declining schedule:

```{r}
s6b <- interpolate_model(mod_ON, name = "lc_scrap_on") |>
  solve_scenario(echo = FALSE)
getData(s6b, "vObjective", merge = TRUE)$value

getData(s6b, name = c("vTechCap", "vTechStockCap", "vTechRetiredStock",
                      "vTechStockPhaseOut", "vTechAOut"),
        merge = TRUE, drop.zeros = FALSE) |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

The cost falls from **12,477.8 to 1,651.1**, and both mechanisms are visible at
once:

- **95 MW retired in 2020** → `95 × 200 = 19,000` t of *intact* scrap;
- the schedule still thins the 5 MW that remain → `0.4 × 120 = 48` t/yr of
  *worn-out* scrap in 2033 and 2038, and 24 in 2043.

The schedule was honoured **and** the optimiser was free to act. Neither
overrides the other: `vTechStockPhaseOut` thins whatever is actually standing,
*after* early retirement.

::: {.callout-note appearance="simple"}
**This did not always work.** Until energyRt 0.85 a declining `stock` silently
disabled early retirement: the retirement limit was read against a schedule that
reached zero, and a zero in the final year pinned cumulative retirement to zero
in *every* year. The model still solved — it just refused to retire anything, and
said nothing about it. If you are reading an older study whose fleet was declared
with a falling `stock` and `optimizeRetirement = TRUE`, check whether anything
actually retired.
:::

For contrast, the same model on a **constant** fleet — swap the technology's
schedule with `update()` and `add()`:

```{r}
CONST <- data.frame(year = c(2020L, 2043L), stock = c(100, 100))

s6c <- add(mod_ON, update(WIND_S, capacity = CONST), overwrite = TRUE) |>
  interpolate_model(name = "lc_scrap_const") |>
  solve_scenario(echo = FALSE)
getData(s6c, "vObjective", merge = TRUE)$value
```

**1,040.0** — cheaper still, because nothing is left to phase out and the 5 MW
survivor never has to be replaced.

**Exercises.**

**6.5a** Set `ret2aout` equal to `pho2aout` (120). What happens to the objective,
and what does that tell you about how much of the saving was the *credit* rather
than the avoided `fixom`?

**6.5b** Attach `newExport("SCRAP", export = data.frame(price = ...))` so recovery
actually pays. At what price does the model retire the fleet *for the scrap
value alone* — that is, with `fixom = 0`?

**6.5c** Predict the objective for the declining fleet with retirement on, before
running: 95 MW × 8/yr avoided, over how many years? Reconcile your figure with
1,651.1.

## 6.6 — Closing the loop {#sec-lca-recycling}

*Learning goal:* one commodity produced by end-of-life and consumed by
construction — a material loop written with the coefficients you already have.

`STEEL` is now both: `ncap2ainp = 150` (needed to build) and `pho2aout = 120`
(recovered when scrapped). Primary steel costs 100; recovered steel is free.
The base technology has no recovery; its recycling twin adds one coefficient:

```{r}
SSTEEL100 <- newSupply(name = "SSTEEL", commodity = "STEEL",
                       supply = data.frame(cost = 100))

WIND_R <- newTechnology(
  name = "WIND",
  input = list(comm = "WND"), output = list(comm = "ELC"), cap2act = 1,
  afs = data.frame(afs.up = 1), vintage = data.frame(olife = 15L),
  invcost = data.frame(invcost = 1),
  aux  = data.frame(acomm = "STEEL", unit = "t"),
  aeff = data.frame(acomm = "STEEL", ncap2ainp = 150))

WIND_RC <- update(WIND_R,
  aeff = data.frame(acomm = "STEEL", ncap2ainp = 150, pho2aout = 120))

DEMR <- newDemand(name = "DEM", commodity = "ELC",
                  demand = data.frame(year = YRS,
                                      demand = c(10, 20, 30, 40, 50, 60)))

repo_REC <- newRepository("repo_REC", WND, ELC, STEEL,
                          SWND, SSTEEL100, WIND_R, DEMR)

mod_REC <- newModel(name = "LC_REC", data = repo_REC, region = "R1",
                    horizon = hor, discount = 0, calendar = cal)

s7a <- interpolate_model(mod_REC, name = "lc_rec_off") |>
  solve_scenario(echo = FALSE)
s7b <- add(mod_REC, WIND_RC, overwrite = TRUE) |>
  interpolate_model(name = "lc_rec_on") |>
  solve_scenario(echo = FALSE)

getData(list(no_recycling = s7a, recycling = s7b),
        name = "vObjective", merge = TRUE) |>
  select(scenario, objective = value)
```

```{r}
getData(s7b, name = c("vTechNewCap", "vTechAInp", "vTechAOut", "vSupOut"),
        merge = TRUE, drop.zeros = FALSE) |>
  filter(is.na(comm) | comm == "STEEL") |>
  mutate(series = ifelse(is.na(comm), name, paste0(name, "[", comm, "]"))) |>
  select(series, year, value) |>
  mutate(value = round(value, 4)) |>
  pivot_wider(names_from = year, values_from = value, values_fill = 0) |>
  arrange(series)
```

The build programme is **identical** in both runs — same `vTechNewCap`, same
`vTechAInp[STEEL]`. What changes is where the steel comes from. From 2038 the
retiring fleet returns 480 t/yr and purchases of primary steel fall from 900 to
420; in 2043, 240 returns and purchases fall from 600 to 360.

The saving is **360,000**, and checking it is a rate lesson in itself: 480 + 240
= 720 t/yr displaced, but each of those milestones stands for **five years**, so
3,600 t × 100 = 360,000. Read the columns as totals and you would be off by a
factor of five.

**Exercises.**

**6.6a** Add a recycling *loss*: only 80% of recovered steel is usable. (Hint: it
is a coefficient, not a new process.) How much of the saving survives?

**6.6b** Make recovered steel worth something even when unused, with an export.
Does the build programme change now?

**6.6c** The loop here is within one technology. Route the scrap into a *second*
technology's `ncap2ainp` instead, and confirm the totals still balance.

## 6.7 — The trap: a one-time coefficient on a fine calendar {#sec-lca-slices}

*Learning goal:* the single most expensive mistake in this chapter.

An aux coefficient given **without a `timeslice`** applies in **every timeslice**.
For a rate — fuel per hour, emissions per MWh — that is exactly right. For a
*one-time* quantity it is a disaster, and the disaster scales with your calendar.

The same model, the same `pho2aout = 120`, on calendars of 1, 4 and 12 slices —
three explicit builds, because the calendar (and with it every commodity's
`timeframe`) is what changes:

```{r}
cal4 <- newCalendar(timetable = make_timetable(
  struct = list(ANNUAL = "ANNUAL", SEASON = sprintf("s%02d", 1:4))),
  name = "c4")
cal12 <- newCalendar(timetable = make_timetable(
  struct = list(ANNUAL = "ANNUAL", SEASON = sprintf("s%02d", 1:12))),
  name = "c12")

WIND_T1 <- newTechnology(
  name = "WIND",
  input = list(comm = "WND"), output = list(comm = "ELC"), cap2act = 1,
  afs = data.frame(afs.up = 1), capacity = FLEET,
  vintage = data.frame(olife = 40L),
  aux  = data.frame(acomm = "SCRAP", unit = "t"),
  # no `timeslice` column: the coefficient lands in EVERY slice
  aeff = data.frame(acomm = "SCRAP", pho2aout = 120))

repo_SL1 <- newRepository("repo_SL1", WND, ELC, SCRAP, SWND, WIND_T1, DEM)
s_sl1 <- newModel(name = "LC_SLICE", data = repo_SL1, region = "R1",
                  horizon = hor, discount = 0, calendar = cal) |>
  interpolate_model(name = "lc_slice1") |> solve_scenario(echo = FALSE)

WND_s  <- newCommodity(name = "WND",   timeframe = "SEASON")
ELC_s  <- newCommodity(name = "ELC",   timeframe = "SEASON")
SCRAP_s <- newCommodity(name = "SCRAP", timeframe = "SEASON")
repo_SLs <- newRepository("repo_SLs", WND_s, ELC_s, SCRAP_s,
                          SWND, WIND_T1, DEM)
s_sl4 <- newModel(name = "LC_SLICE", data = repo_SLs, region = "R1",
                  horizon = hor, discount = 0, calendar = cal4) |>
  interpolate_model(name = "lc_slice4") |> solve_scenario(echo = FALSE)
s_sl12 <- newModel(name = "LC_SLICE", data = repo_SLs, region = "R1",
                   horizon = hor, discount = 0, calendar = cal12) |>
  interpolate_model(name = "lc_slice12") |> solve_scenario(echo = FALSE)

getData(list(`1` = s_sl1, `4` = s_sl4, `12` = s_sl12),
        name = "vTechAOut", merge = TRUE) |>
  filter(comm == "SCRAP") |>
  group_by(timeslices = scenario) |>
  summarise(total_SCRAP = sum(value), .groups = "drop")
```

**Exactly linear in the slice count.** The same fleet, the same coefficient, and
an answer that is 12× too large — with no warning, no error, and a model that
solves in the usual time. On the 8,760-slice calendar of @sec-realmodel this
figure would read **21 million tonnes**.

::: {.callout-warning appearance="simple"}
**Rule of thumb.** If the quantity is *per unit of something that happens in a
timeslice* (per MWh, per hour of activity), leave `timeslice` unset. If it is
*per unit of capacity* — built, standing, phased out or retired — name the
timeslice explicitly, or divide by the slice count and know that you have.
:::

This is not a quirk of the end-of-life coefficients. `ncap2ainp` behaves
identically and always has. It is simply easier to notice when the number is a
tonnage of steel than when it is a fraction.

**Exercises.**

**6.7a** Fix the model: add `timeslice = "ANNUAL"` to the `aeff` row and confirm
all three calendars now agree at 2,400.

**6.7b** Which of the eight `<driver>2a<inp|out>` coefficients are exposed to this
trap, and which are not? Justify each from its driver.

**6.7c** Write a check you could run over any model — a function that flags
capacity-driven aux coefficients with an unset `timeslice`.

## Where this connects

- @sec-processes introduced aux flows as "anything riding alongside the
  conversion". This chapter is that idea taken to the ends of the plant's life.
- @sec-realmodel's retirement exercise found that `optimizeRetirement = TRUE`
  changed nothing, and correctly reported it. §6.5 shows the opposite case — and
  the version of the defect that would have made "nothing retired" a *lie*.
- @sec-replication's converted PyPSA-Eur model carries no life-cycle data at all.
  Adding land, materials and end-of-life to it is a genuine open exercise, and a
  good one.

**Takeaways**

- **The driver is the physics.** `cap2a*` is a plateau, `ncap2a*` a pulse at
  build, `pho2a*` and `ret2a*` a pulse at the end. Choose by *when the thing
  happens*, not by which coefficient you used last time.
- **`stock` is what is still standing**, and energyRt derives a commissioning
  stream and a survival share from it. Both are reported; look at them.
- **A schedule and an optimiser can coexist.** Phase-out thins what remains after
  retirement; declaring a decommissioning timetable does not take the decision
  away.
- **Everything is a per-year rate.** Multiply by the period length before you
  total anything, and always test a period of a different length.
- **Duals are not exported — difference two solves.** Agreement between a small
  step and a large one is what tells you the price is real.
- **An unset `timeslice` means every timeslice.** For a one-time quantity that is
  a silent multiplication by the size of your calendar.
