---
output: html_document
editor_options:
  chunk_output_type: inline
---
# A real model, end to end {#sec-realmodel}

Every model so far you built by hand, object by object, small enough to hold in
your head. This chapter starts from the other end: a model **derived from
[PyPSA-Eur](https://pypsa-eur.readthedocs.io)** — Belgium as a single node, a
full year at hourly resolution, with real plant, real weather and real demand —
and does the things you actually do with a model you did not write.

Solve it. Measure it. Shrink it. Sample it. Stretch it to 2050. Let it retire its
own plant. Price its carbon, then cap it. Double its demand and check the answer
against PyPSA.

Two smaller and larger variants of the same conversion ship alongside it, and
@sec-choose shows how to swap them in.

```{r}
library(energyRt)
library(dplyr)
library(ggplot2)
```

::: {.callout-note appearance="simple"}
Nothing executes at render (`eval: false` site-wide). The numbers quoted in the
text come from a local run and are there so you can tell whether yours went
right — not as values to reproduce exactly.
:::

## The dataset {#sec-be-data}

The model ships with the course as one serialised object. It is not a PyPSA file
and needs no Python: the conversion already happened (see @sec-provenance for
how, and @sec-licences for the data licences).

```{r}
cp   <- readRDS("data/belgium_copperplate.rds")
repo <- cp@data[[1]]
summary(repo)
```

Belgium as a **copperplate**: one node, no transmission, the full 8,760-hour
year. Small in objects, large in time.

```{r}
getCalendar(cp)        # 8760 timeslices, year_fraction = 1
getHorizon(cp)         # a single milestone year, 2050
cp@config@region       # one region
```

Three things are worth naming before you use it.

**No transmission.** With one node there are no lines, so nothing in this chapter
depends on how power flow is modelled. @sec-replication shows what changes when
there are 77 AC lines.

**A full year, so the weighting is simply correct.** `year_fraction = 1` and
`pTimesliceWeight = 1`. Operating cost and capital cost are both annual, with no
convention to choose. That is *not* true of a sampled model, and @sec-provenance
explains what it costs when it is not.

**Fuel is free at the supply.** PyPSA puts fuel cost on the generator, not the
fuel, so `SUP_GAS` costs 0 and the price sits in each technology's `varom`.
Carbon, though, rides on the *commodity*:

```{r}
emitters <- Filter(function(x) NROW(x@emis), getObject(cp, class = "commodity"))
lapply(emitters, function(x) cbind(comm = x@name, x@emis))
```

`GAS` emits 0.198 t CO₂ per unit, `OIL` and `WST` 0.2571. That is what @sec-tax
and @sec-cap act on.

## Choosing a model {#sec-choose}

Three converted models ship with the course. They are the *same* conversion at
three sizes, so anything you write for one runs on the others.

| file | regions | timeslices | on disk |
|---|---:|---:|---:|
| `belgium_copperplate.rds` | 1 | 8,760 | 0.28 MB |
| `belgium_model.rds` | 5 | 168 | 0.02 MB |
| `eu41_model.rds` | 41 | 8,760 | 12.7 MB |

```{r}
be5 <- readRDS("data/belgium_model.rds")     # 5 Belgian nodes, one week
eu  <- readRDS("data/eu41_model.rds")        # 41 European nodes, full year
summary(eu@data[[1]])
```

Swapping the model changes what the chapter costs to run:

| model | variables | solve |
|---|---:|---|
| copperplate, full year | ~473,000 | ~4½ minutes |
| copperplate, sampled (below) | ~15,600 | ~20 seconds |
| 5-node Belgium, one week | ~40,000 | ~25 seconds |
| 41-node Europe, full year | ~28 **million** | no open solver will take it |

Two pieces of advice if you go to the European model:

- **Use Julia/HiGHS** (`solver_options$julia_highs`). GLPK is a teaching solver
  and stops being practical well before a continental model — see @sec-backends
  for the measurement.
- **Always sample the calendar.** @sec-replication solves the 41-node model on
  four representative days and reports what that costs.

`model_size()` before `solve_scen()` is a good habit on any model you have not
solved before. It is cheap, and it tells you whether the next command takes a
second or an afternoon.

## 6.1 — Solve it {#sec-solve}

**6.1 — Interpolate and solve.** Turn the model into a scenario, solve it, and
report the objective.

::: {.callout-tip collapse="true" title="Solution 6.1"}
```{r}
scen <- interpolate_model(cp, name = "be_cp")
sol  <- solve_scen(scen, solver = solver_options$julia_highs)

getData(sol, "vObjective", merge = TRUE)
```

**8,260,626,305.45**, in about four and a half minutes.

A bare number until you know its units, and establishing those is the first thing
to do with any model you did not build. Here it is genuinely **annual system
cost** — annualised capital plus a full year of operating cost — because the
calendar covers the whole year. On a sampled model it would not be, and that
distinction is the subject of @sec-provenance.

`interpolate_model()` collects sets from the objects, builds the mapping layer,
interpolates parameters over the milestone years and returns a solver-ready
`scenario`. `solve_scen()` writes the model file, runs the solver, reads the
solution back.

Two habits worth forming now. **Name your scenarios** — the name is what
@sec-compare keys on. And **keep the scenario**, not just the solution: it holds
the parameters you will want when a result surprises you.
:::

## 6.2 — How big is it? {#sec-size}

Before shrinking anything, measure it. `model_size()` estimates variables and
constraints from the gating maps and lists the heaviest parameters.

**6.2 — Measure the model.** Report variables and constraints, and identify the
parameters holding the most rows. Explain what they have in common.

::: {.callout-tip collapse="true" title="Solution 6.2"}
```{r}
model_size(scen)
```

```
  parameters : 143 value, 260 maps, 13 sets
  param rows : 286,655
  estimate   : ~473,133 variables, ~508,159 constraints
  top parameters by rows:
    pTechVarom     113,880      pDemand         8,760
    pTechCinp2use   61,320      pTechAf         8,760
    pWeather        40,941      pTechCap           11
```

Every leader is indexed by `timeslice`. With 8,760 of them, anything carrying
that dimension dwarfs anything that does not: `pTechCap` has **11** rows,
`pTechVarom` has **113,880**.

**Dimensionality, not object count, is what makes a model big.** Forty-odd
objects on one region produce a half-million-variable LP — purely because the
year is resolved hourly. This is why the temporal and regional resolution
decisions of @sec-time and @sec-regions are the expensive ones.
:::

## 6.3 — Shrink it: sparse, prune, fold, trim {#sec-shrink}

`interpolate_model()` has three storage knobs, plus one step you apply afterwards:

| knob | what it does |
|---|---|
| `sparse` | drops rows whose value equals the parameter's default |
| `prune` | drops interpolated rows outside the equation-domain maps |
| `fold` | collapses a whole column of a trimmable dimension to an `NA` wildcard |
| `trim_parameters_by_maps()` | drops rows that no map indexes at all |

None of them changes the solution. All change how much data you carry.

**6.3 — Measure each step.** Interpolate with every knob off, switch them on one
at a time, and record parameter rows. Determine which knob pays here and which
does nothing — and why that is informative rather than disappointing.

::: {.callout-tip collapse="true" title="Solution 6.3"}
```{r}
rows <- function(sc) sum(vapply(sc@modInp@parameters,
  function(p) if (is.null(p)) 0L else NROW(p@data), integer(1)))

raw <- interpolate_model(cp, name = "raw", sparse = FALSE, prune = FALSE, fold = FALSE)
sp  <- interpolate_model(cp, name = "sp",  sparse = TRUE,  prune = FALSE, fold = FALSE)
pr  <- interpolate_model(cp, name = "pr",  sparse = TRUE,  prune = TRUE,  fold = FALSE)
fd  <- interpolate_model(cp, name = "fd",  sparse = TRUE,  prune = TRUE,  fold = TRUE)
tr  <- trim_parameters_by_maps(pr)

data.frame(
  step = c("raw", "+sparse", "+prune", "+fold", "+trim"),
  rows = vapply(list(raw, sp, pr, fd, tr), rows, integer(1)),
  size = vapply(list(raw, sp, pr, fd, tr),
                function(x) format(object.size(x), units = "MB"), character(1))
)
```

A local run gives:

| step | parameter rows | size |
|---|---:|---:|
| raw | 1,942,567 | 116.4 MB |
| +sparse | 1,536,711 | 97.4 MB |
| +prune | 1,536,711 | 97.4 MB |
| +fold | 1,308,977 | 84.5 MB |
| +trim | 1,536,711 | 97.4 MB |

`sparse` pays: it strips 405,856 rows that merely restated a default. `fold` pays
again, collapsing a further 227,734. **`prune` and `trim` remove nothing at all**
— and that is the interesting result, not a failure.

Both exist to catch rows the equations never ask for. Removing nothing means the
mapping layer already built these parameters over exactly their equation domains.
On a model where `prune` strips a lot, it is telling you interpolation is
generating year or region combinations no equation uses — usually a sign of a
horizon or region set wider than the data supports.

So treat these knobs as **diagnostics first, savings second**. A folded scenario
is expanded again at solve time, trading a little CPU for memory: worth it when
parameter tables stop fitting comfortably, not before.
:::

## A sampled calendar {#sec-sample}

The rest of the chapter varies the model — six milestone years, then a sweep of
carbon prices. At four and a half minutes a solve, that is an afternoon.

energyRt separates the **model** from the **calendar it is solved on**. The model
carries the whole year; a scenario can be interpolated onto a *subset* of its
timeslices. One such calendar ships with the package:

```{r}
data(calendars)
cal <- calendars$d365_h24_subset_1day_per_month
nrow(cal@timetable)     # 288 = 12 days x 24 hours
cal@year_fraction       # 0.0329  ->  pTimesliceWeight = 30.42
```

One day per month, all 24 hours. Pass it as an extra argument:

```{r}
scen_s <- interpolate_model(cp, cal, name = "be_s")
model_size(scen_s)      # ~15,600 variables instead of ~473,000
sol_s  <- solve_scen(scen_s, solver = solver_options$julia_highs)
```

::: {.callout-note appearance="simple"}
**What the sample costs.** The 288-slice answer is **8,157,519,902** against the
full year's 8,260,626,305 — **1.25% low** — and solves in about 20 seconds
instead of four and a half minutes.

It is not free, and the error is not a fixed property of "sampling". A *four*-day
sample of this same model is **3.10% off** and picks a different solar technology
entirely. Twelve days sees twelve days of weather, so it still understates
anything that rides out a long lull — seasonal storage most of all.
Representative-period sampling is a modelling decision, not a convenience.
:::

Note also what the weighting now does: `pTimesliceWeight` is 30.42, so the 288
modelled hours stand in for 8,760. The objective is again annual, but by
*scaling* rather than by counting.

## 6.4 — Different back-ends {#sec-backends}

energyRt writes the same model for several solvers, generating the algebra from
one specification per back-end. **The objective must agree.** A real
disagreement is a bug in a template, not a matter of solver taste.

**6.4 — Solve the same scenario on several back-ends.** Compare objectives and
wall-clock time. Determine what you would conclude from a disagreement in the
tenth digit, and what from one in the third.

::: {.callout-tip collapse="true" title="Solution 6.4"}
```{r}
names(solver_options)   # what is available

backends <- c("glpk", "julia_highs", "pyomo_glpk", "pyomo_cbc")

bench <- lapply(backends, function(b) {
  t0 <- Sys.time()
  s  <- try(solve_scen(scen_s, name = paste0("be_", b), solver = solver_options[[b]]),
            silent = TRUE)
  data.frame(
    backend   = b,
    seconds   = round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1),
    objective = if (inherits(s, "try-error")) NA_real_
                else getData(s, "vObjective", merge = TRUE)$value[1]
  )
}) |> bind_rows()

bench
```

On the sampled scenario:

| back-end | seconds | objective |
|---|---:|---:|
| glpk | 2.0 | 8,157,519,901.55 |
| julia_highs | 20.2 | 8,157,519,901.55 |
| pyomo_glpk | 6.4 | 8,157,519,901.55 |
| pyomo_cbc | 5.2 | 8,157,519,9**00.00** |

Three back-ends agree to every digit printed; CBC differs by 1.55 — a relative
difference of **1.9 × 10⁻¹⁰**, far below the precision of any input. Ignore it.

A disagreement in the third digit would be a different matter: either the
templates disagree about an equation, or a solver stopped early. Always check
status before believing a number — a run that hits a time limit can return a
*feasible* point that looks perfectly plausible and is not optimal.

**And note the ranking, which inverts with size.** On this small scenario GLPK is
the *fastest*: Julia's start-up dominates. Run the same comparison on the
**full-year** model and the order reverses hard:

| back-end | 288 slices | 8,760 slices |
|---|---:|---:|
| glpk | **2.0 s** | 1,384 s (23 min) |
| julia_highs | 20.2 s | **275 s** |

GLPK still gets there, and agrees with HiGHS to eleven significant figures
(8,260,626,305.56 against 8,260,626,305.45) — but it takes five times as long.
"The serious solver is faster" is not something to assume at any particular size;
it is something to measure. And the corollary matters more: a back-end that is
fine for a teaching model can be hopeless two orders of magnitude up.

You need each back-end installed; one you do not have raises an error, which is
why the loop wraps `try()`. See the **Installation** chapter.
:::

## 6.5 — A planning horizon to 2050 {#sec-horizon}

The model has one milestone year. Planning studies want a path: when capacity is
built, when it retires, what the system looks like in between.

**6.5 — Extend the horizon.** Solve over 2025–2050 in five-year steps on the
sampled calendar. Report the model size, plot the capacity path — and then look
hard at what it built.

::: {.callout-tip collapse="true" title="Solution 6.5"}
```{r}
hor <- newHorizon(2025:2050, intervals = c(1, 5, 5, 5, 5, 5))
hor@intervals
```

Note where the milestones land: **2025, 2028, 2033, 2038, 2043, 2048**. energyRt
places each milestone at its interval's *mid-point*, so "five-year steps to 2050"
does not put a milestone on 2050. Check `hor@intervals` rather than assuming.

```{r}
scen_2050 <- interpolate_model(cp, cal, hor, name = "be_2050")
sol_2050  <- solve_scen(scen_2050, solver = solver_options$julia_highs)

model_size(scen_2050)      # ~93,900 variables

getData(sol_2050, "vTechCap", merge = TRUE) |>
  group_by(year, tech) |> summarise(cap = sum(value), .groups = "drop") |>
  ggplot(aes(year, cap, fill = tech)) + geom_area() +
  labs(x = NULL, y = "capacity") + theme_bw()
```

Objective **15,187,226,948**, and by 2048 the system is solar-hsat 15,191,
biomass 10,999, onwind 7,067, offwind-dc 6,018. Total CO₂: **zero**.

Now stop and check that, because it is wrong.
:::

**6.5b — Something is off. Find it.** The base model has **113 MW** of biomass
and PyPSA marks it non-extendable. The 2048 solution has **10,999 MW**. Work out
why, and fix it.

::: {.callout-tip collapse="true" title="Solution 6.5b"}
```{r}
bio <- getObject(cp, name = "E_BIOMASS")[[1]]
bio@capacity     # stock 113, cap.up NA, year 2050
bio@vintage      # end = 2049
```

Two things conspire, and both come from the same root: **the converted model
describes a single year, 2050.**

- The converter closes investment on non-extendable plant by setting
  `vintage$end = 2049`, one year before the model's own milestone. Every
  milestone on this horizon — 2025 to 2048 — falls *before* 2049, so the
  investment window is **open** in all of them.
- The capacity row is keyed to `year = 2050`. Any bound you add there simply does
  not apply to the earlier milestones either.

So extending the horizon silently unlocked plant that PyPSA holds fixed. Fixing
it needs the bound *and* a wildcard year:

```{r}
cp_fix <- cp
r <- cp_fix@data[[1]]
for (i in seq_along(r@data)) {
  o <- r@data[[i]]
  if (!is(o, "technology") || !NROW(o@vintage) || is.na(o@vintage$end[1])) next
  st <- o@capacity$stock[1]; if (is.na(st)) next
  o@capacity$cap.up <- st
  o@capacity$year   <- NA_integer_     # NA = every year, not just 2050
  r@data[[i]] <- o
}
cp_fix@data[[1]] <- r

sol_fix <- solve_scen(interpolate_model(cp_fix, cal, hor, name = "be_fix"),
                      solver = solver_options$julia_highs)
```

| | as shipped | fixed |
|---|---:|---:|
| objective | 15,187,226,948 | **21,501,935,951** |
| CO₂ | **0** | **1,155,508 t** |
| biomass in 2048 | 10,999 MW | 113 MW |
| fuels burned | biomass only | gas 4.57 M, waste, oil, biomass |

The naive answer was **42% too cheap** and reported zero emissions for a system
that actually burns 4.6 million units of gas.

Two lessons, and the second is the general one:

- **Setting `cap.up` alone does nothing** if the row is keyed to a year your
  horizon never visits. Wildcard the year.
- **A converted model carries the assumptions of the run it came from.** This one
  was built for a single 2050 snapshot; it does not know what "2030" means. Any
  time you extend a model beyond the horizon it was made for, audit the bounds
  before believing the answer — and a capacity 97× its stated stock is exactly
  the kind of thing to notice.
:::

## 6.6 — Endogenous retirement {#sec-retire}

By default a plant lives out its technical life. With `optimizeRetirement` the
model may retire it early, when paying fixed costs is worse than losing it.

**6.6 — Let the model retire plant.** Solve the 2050 horizon with
`optimizeRetirement = TRUE`. Before running it, predict the direction the
objective must move — then check.

::: {.callout-tip collapse="true" title="Solution 6.6"}
```{r}
cp_ret <- cp
cp_ret@config@optimizeRetirement <- TRUE

sol_ret <- solve_scen(interpolate_model(cp_ret, cal, hor, name = "be_ret"),
                      solver = solver_options$julia_highs)

getData(sol_ret, "vObjective", merge = TRUE)
getData(sol_ret, "vTechRetiredStock", merge = TRUE)
```

The objective is **15,187,226,948 — identical — and nothing retires at all.**

You should have predicted the direction, if not the exact tie: retirement is an
**option**, and adding an option to an optimisation can never make the optimum
worse. It can only improve or stay equal. (If yours got worse, something else
changed too.)

Here it stayed exactly equal, which means the option was **worthless on this
model**. That is a result, and the right thing to do is report it — not tune
parameters until something retires. It makes sense once you look at 6.5: the
fossil fleet is already sitting idle, and an idle plant with no fixed cost worth
avoiding has no reason to be scrapped.

One asymmetry to know: retirement is implemented for **technologies**. Storage
carries the retirement data and mappings, but the equations are not in the
back-ends, so a storage will not retire early whatever you set.
:::

## 6.7 — A carbon tax {#sec-tax}

A tax fixes the **price** of carbon and lets the quantity float. Back to the
single-period scenario, where the fleet still burns fuel.

**6.7 — Price the carbon.** Sweep a range of carbon prices and plot emissions
against price. Determine whether the response is smooth.

::: {.callout-tip collapse="true" title="Solution 6.7"}
```{r}
tax_run <- function(level) {
  TAX <- newTax(name = "CT_CO2", comm = "CO2",
                tax = data.frame(year = 2050, bal = level))
  s <- interpolate_model(cp, cal, TAX, name = paste0("tax", level))
  o <- solve_scen(s, solver = solver_options$julia_highs)
  data.frame(tax = level,
             objective = getData(o, "vObjective", merge = TRUE)$value[1],
             co2 = sum(getData(o, "vEmsFuelTot", merge = TRUE)$value))
}

sweep_tax <- bind_rows(lapply(c(0, 10, 30, 60, 120), tax_run))
sweep_tax
```

| tax | objective | CO₂ | abated |
|---:|---:|---:|---:|
| 0 | 8,157,519,902 | 380,814 | — |
| 10 | 8,270,786,273 | 357,103 | 6% |
| 30 | 8,443,801,834 | 254,244 | 33% |
| 60 | 8,652,652,655 | 216,577 | 43% |
| 120 | 8,967,338,499 | 113,828 | **70%** |

The response is **not smooth**. Between 10 and 30 the price triples and buys 27
percentage points; between 30 and 60 it doubles and buys only 10.

That is normal for a linear program and worth internalising, because it defeats
the intuition people bring from smooth economics. An LP's response to a price is
piecewise linear with kinks — each kink a technology becoming worth switching —
so **you cannot interpolate between two runs** and claim the answer in between.
If you need the shape, sweep it.

Check units before believing any of this. Emission factors are tonnes per unit of
fuel commodity; the tax is money per tonne of `CO2`; both must be in the same
money unit as the objective. Off by a factor of 1000, the model either ignores
the tax or shuts the system down — and both look like plausible model behaviour,
which is what makes unit errors here so dangerous.
:::

## 6.8 — A carbon cap {#sec-cap}

A cap fixes the **quantity** and lets the price float — the mirror image.

**6.8 — Cap the carbon.** Constrain CO₂ to fractions of the unconstrained level
and compare the cost of each. Determine the shape of the resulting curve.

::: {.callout-tip collapse="true" title="Solution 6.8"}
```{r}
base_co2 <- sum(getData(sol_s, "vEmsFuelTot", merge = TRUE)$value)   # 380,814

cap_run <- function(frac) {
  CAP <- newConstraint(
    name = "CO2_CAP", eq = "<=",
    for.each = data.frame(year = 2050, comm = "CO2"),
    term1    = list(variable = "vEmsFuelTot"),
    rhs      = data.frame(year = 2050, rhs = frac * base_co2),
    defVal   = Inf)
  o <- solve_scen(interpolate_model(cp, cal, CAP, name = paste0("cap", frac * 100)),
                  solver = solver_options$julia_highs)
  data.frame(cap = frac,
             objective = getData(o, "vObjective", merge = TRUE)$value[1],
             co2 = sum(getData(o, "vEmsFuelTot", merge = TRUE)$value))
}

sweep_cap <- bind_rows(lapply(c(1, 0.75, 0.5, 0.25), cap_run))
sweep_cap |> mutate(cost_per_t = (objective - objective[1]) / (co2[1] - co2))
```

| cap | CO₂ | objective | average cost per tonne |
|---:|---:|---:|---:|
| 100% | 380,814 | 8,157,519,902 | — |
| 75% | 285,611 | 8,187,324,954 | 313 |
| 50% | 190,407 | 8,316,370,475 | 834 |
| 25% | 95,204 | 8,635,818,535 | 1,675 |

Every cap **binds exactly** — emissions land on the ceiling to the tonne, which
is the first thing to check: a cap that does not bind is not a policy, it is a
no-op.

The cost per tonne rises 313 → 834 → 1,675. That is a **marginal abatement cost
curve**, and its convexity is the whole story of climate policy cost: cheap
options first, expensive ones last. It is steep here because the base case has
already taken the cheap options — this is a 2050 cost world with a largely
decarbonised fleet, so what remains is genuinely expensive to remove.

Set the ceiling *from the unconstrained result*, as here, which is why 6.1 runs
first. A cap below what the system can physically reach makes the model
infeasible, and **infeasibility is information, not an error to code around**: it
says nothing in this system can deliver that ceiling. The useful response is to
ask what would have to be added, not to relax the cap until the solver stops
complaining.
:::

## 6.9 — Compare the scenarios {#sec-compare}

**6.9 — Put tax and cap side by side.** Compare a tax and a cap that deliver
similar abatement. Determine which is cheaper, and explain the difference.

::: {.callout-tip collapse="true" title="Solution 6.9"}
Pick the pair that lands closest together — tax 60 (CO₂ 216,577) and cap 50%
(CO₂ 190,407):

| policy | CO₂ | objective | vs base |
|---|---:|---:|---:|
| base | 380,814 | 8,157,519,902 | — |
| tax 60 | 216,577 | 8,652,652,655 | +495,132,753 |
| cap 50% | 190,407 | 8,316,370,475 | +158,850,573 |

The tax costs **three times more** while abating *less*. That looks like a
paradox and is not one — it is the single most misread number in this kind of
comparison.

The cap's cost increase is pure **resource cost**: the extra expense of running a
cleaner system. The tax's increase is resource cost **plus a transfer** — it
charges the carbon price on every one of the ~216,577 tonnes still emitted, and
those payments go to a treasury, not up a smokestack. At 60 per tonne that
transfer alone is about 13 million.

So:

- **Never compare a tax and a cap on objective value alone.** You are comparing a
  resource cost with a resource cost plus a transfer.
- To compare them properly, strip the tax revenue out, or compare the cap's
  **shadow price** with the tax rate. Set the tax equal to the cap's shadow price
  and the two coincide — they are duals.
- Which instrument you prefer follows from what you want certainty about:
  emissions (cap) or cost (tax).

```{r}
TAX60 <- newTax(name = "CT_CO2", comm = "CO2",
                tax = data.frame(year = 2050, bal = 60))
sol_tax <- solve_scen(interpolate_model(cp, cal, TAX60, name = "cmp_tax"),
                      solver = solver_options$julia_highs)

CAP50 <- newConstraint(
  name = "CO2_CAP", eq = "<=",
  for.each = data.frame(year = 2050, comm = "CO2"),
  term1    = list(variable = "vEmsFuelTot"),
  rhs      = data.frame(year = 2050, rhs = 0.50 * base_co2),
  defVal   = Inf)
sol_cap <- solve_scen(interpolate_model(cp, cal, CAP50, name = "cmp_cap"),
                      solver = solver_options$julia_highs)

scns <- list(base = sol_s, tax = sol_tax, cap = sol_cap)

lapply(names(scns), function(nm) {
  getData(scns[[nm]], "vTechOut", merge = TRUE) |>
    group_by(tech) |> summarise(out = sum(value), .groups = "drop") |>
    mutate(scenario = nm)
}) |> bind_rows() |>
  ggplot(aes(scenario, out, fill = tech)) + geom_col() + theme_bw()
```

Plot the generation mix as well as the totals. An aggregate cost hides which
plant actually moved, and "which plant moved" is usually the finding worth
reporting.
:::

## 6.10 — Demand doubles by 2050 {#sec-growth}

Everything so far dispatched a fleet that already existed. This is the first
question that makes the model **build**: what happens to Belgium if electricity
demand doubles?

**Back to the full year for this one.** @sec-sample traded 1.25% of accuracy for
speed, which was a good bargain while varying horizons and carbon prices. It is a
bad bargain here: this section is checked against PyPSA to the cent, and a
sampled calendar would put a modelling approximation between the two models. With
the full 8,760 hours and no transmission, nothing is left to excuse a difference
— which is what makes the comparison in 6.10b worth anything.

Its starting point:

| | |
|---|---:|
| annual demand | 86,292,102 MWh |
| peak demand | 13,645 MW |
| firm plant | 11,267 MW (CCGT 6,634 + nuclear 4,096 + waste 294 + oil 130 + biomass 113) |

Peak already exceeds firm capacity, covered by wind, solar and storage. Doubling
takes peak to ~27.3 GW, so expansion is not optional.

What can grow, and by how much, is what shapes the answer:

| carrier | now | ceiling |
|---|---:|---:|
| CCGT, nuclear | 6,634 / 4,096 | **unbounded** |
| solar | 938 | 106,452 |
| solar-hsat | 0 | 92,467 |
| onwind | 1,724 | 15,235 |
| offwind-ac | 2,752 | 2,752 (**pinned**) |

**6.10 — Double the demand.** Solve the full-year copperplate as shipped, then
again with demand doubled, and report what gets built. Then compare against the
same experiment run in PyPSA. Predict, before you look, whether the extra demand
is met by firm capacity or by renewables.

Each of these solves takes about four and a half minutes.

::: {.callout-tip collapse="true" title="Solution 6.10"}
Doubling demand means editing the `demand` slot. There is no override shortcut
here — the objects you passed through `...` in @sec-tax and @sec-cap were *new*
objects, and passing a second `DEM_ELC` instead raises
`Duplicated objects in "class - name"`. So reach into the repository and replace
it:

```{r}
cp2 <- cp
i <- which(vapply(cp2@data[[1]]@data, function(x) is(x, "demand"), logical(1)))
d <- cp2@data[[1]]@data[[i]]
d@demand$demand <- d@demand$demand * 2
cp2@data[[1]]@data[[i]] <- d

base <- solve_scen(interpolate_model(cp,  name = "cp_base"),
                   solver = solver_options$julia_highs)
x2   <- solve_scen(interpolate_model(cp2, name = "cp_2x"),
                   solver = solver_options$julia_highs)
```

::: {.callout-warning appearance="simple"}
Note the missing `cal`: these interpolate on the model's **own** 8,760-hour
calendar, not the sample of @sec-sample. That is ~473,000 variables and about
**4½ minutes** per solve. Start it before a coffee, not before a demo.
:::

```{r}
capof <- function(s) {
  getData(s, "vTechCap", merge = TRUE) |>
    group_by(tech) |> summarise(cap = sum(value), .groups = "drop")
}
full_join(capof(base), capof(x2), by = "tech", suffix = c(".base", ".x2")) |>
  mutate(built = cap.x2 - cap.base) |>
  arrange(desc(built))
```

The result:

| technology | base | 2× demand | built |
|---|---:|---:|---:|
| CCGT | 6,634.2 | 16,885.1 | **+10,250.9** |
| solar | 10,664.4 | 19,946.2 | +9,281.7 |
| solar-hsat | 1,427.4 | 10,658.8 | +9,231.4 |
| onwind | 1,724.0 | 4,122.8 | +2,398.8 |
| nuclear, offwind, oil, waste, biomass | | unchanged | 0 |

Objective 8,260,626,305 → 13,387,554,321, a **62% cost increase for 100% more
demand** — the system gets cheaper per MWh, because the existing fleet and the
fixed costs are spread over twice the output.

Neither prediction is quite right: the extra demand is met by **both**. Gas takes
the largest single share because doubling the *peak* needs firm capacity that
wind and solar cannot guarantee, but the three renewable carriers together add
20.9 GW against gas's 10.3 GW. Storage grows to match: the battery roughly
triples (578 → 1,806 MW) and the hydrogen store more than doubles
(292,732 → 694,043 MWh). Note that solar-hsat is now worth building, having been
nearly absent in the base case — at higher renewable shares its flatter output
profile earns its extra capital cost.
:::

**6.10b — Compare against PyPSA.** The same experiment was run in PyPSA and
shipped with the course, so you do not need a Python installation. Compare
capacity, and account for any technology that does *not* match.

::: {.callout-tip collapse="true" title="Solution 6.10b"}
```{r}
py <- readRDS("data/be_copperplate_pypsa.rds")
py$note
py$x2$capacity
```

Every generator matches, to within 0.05 MW:

| | energyRt built | PyPSA built |
|---|---:|---:|
| CCGT | +10,250.883 | +10,250.9 |
| solar | +9,281.715 | +9,281.7 |
| solar-hsat | +9,231.439 | +9,231.4 |
| onwind | +2,398.751 | +2,398.8 |
| **total** | **31,162.8** | **31,162.8** |

and the objectives agree to the cent — 8,260,626,305.45 on both in the base case,
13,387,554,321.05 versus .06 with demand doubled.

**Two models, two languages, two formulations, the same answer.** That is the
whole point of the comparison: it is a validation that the translation preserves
the mathematics, not a competition between the tools.

Two things do *not* line up, and both are instructive rather than wrong:

- **Compare `objective_total`, not `objective`.** PyPSA leaves the capital cost of
  non-extendable plant out of `n.objective` and reports it separately as
  `objective_constant` (5,723,317,658 here). energyRt puts everything in one
  number. Add them before comparing, or you will be comparing 2.5 bn with 8.3 bn
  and concluding something dramatic and false.
- **The hydrogen converters differ.** PyPSA rates a `Link` on its **input** bus,
  energyRt rates a technology on its **activity**. The electrolyser confirms the
  rule exactly: PyPSA 162.058 MW × 0.6994 efficiency = 113.3434, which is
  energyRt's `CHR_H2` to four decimals. The fuel cell does *not* reconcile
  (5,593 versus 5,501) — because its `capital_cost` and `marginal_cost` are both
  zero in PyPSA, so its size is a free variable and any value that carries the
  flow is optimal. A degenerate alternate optimum, not a discrepancy.

That second point is the more useful habit: before treating a mismatch as a bug,
check whether the quantity is actually pinned by the optimisation at all.
:::

## 6.11 — How long is your storage? {#sec-duration}

A storage's *capacity* tells you how much energy it holds. It does not tell you
**how long the energy stays there** — and that is usually the more interesting
number, because it is what separates a battery from a seasonal store.

`storage_duration()` answers it. It takes a solved scenario and decomposes each
storage level into bands by how long the energy actually sits:

```{r}
sd <- storage_duration(sol)     # `sol` from 6.1 -- needs the FULL-YEAR solution
head(sd)
levels(sd$duration)
```

```
[1] "<12h"   "12h-1d" "1d-1w"  "1w-30d" ">30d"
```

The method is a nested floor, not a moving average: for each window width it
finds the energy that never leaves the store across *any* window of that length,
then takes successive differences. Energy in the `>30d` band is energy that was
present throughout every 30-day window — it never came out.

::: {.callout-note appearance="simple"}
**This needs the full year.** The bands go up to 30 days, so a 288-slice sample
(@sec-sample) cannot support them — twelve scattered days contain no 30-day
window at all. The decomposition is shipped precomputed so you need not re-solve:

```{r}
sd <- readRDS("data/belgium_storage_duration.rds")
```
:::

**6.11 — Decompose the two stores.** Belgium has a battery and a hydrogen store.
Compute the share of each one's average level that falls in each band, and say
what distinguishes them.

::: {.callout-tip collapse="true" title="Solution 6.11"}
```{r}
sd |>
  group_by(stg, duration) |>
  summarise(mean = mean(value), .groups = "drop") |>
  group_by(stg) |>
  mutate(share = 100 * mean / sum(mean)) |>
  ungroup()
```

| band | STG_BATTERY | STG_H2 |
|---|---:|---:|
| <12h | 24.3% | 0.01% |
| 12h–1d | 16.5% | 0.02% |
| 1d–1w | 40.8% | 0.41% |
| 1w–30d | 17.1% | 1.11% |
| **>30d** | **1.3%** | **98.4%** |

Two storages, the same units, and almost no overlap. **98.4% of the hydrogen
store's energy sits for more than a month**; the battery has 1.3% there and does
most of its work inside a week.

That is the difference between the two technologies stated as a *result* rather
than an assumption. Nothing in the model declared H₂ to be seasonal — no cycle
count, no duration parameter. It came out that way because the optimiser found
that the cheapest way to cover a winter lull is to fill a store in summer.

Now plot it. The interesting view is over time, stacked:

```{r}
sd |>
  filter(stg == "STG_H2") |>
  ggplot(aes(datetime, value, fill = duration)) +
  geom_area() +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(x = NULL, y = "storage level", fill = "held for",
       title = "STG_H2 — a seasonal store") +
  theme_bw()
```

```{r}
sd |>
  filter(stg == "STG_BATTERY") |>
  ggplot(aes(datetime, value, fill = duration)) +
  geom_area() +
  scale_fill_viridis_d(option = "D", direction = -1) +
  labs(x = NULL, y = "storage level", fill = "held for",
       title = "STG_BATTERY — a daily cycler") +
  theme_bw()
```

The hydrogen chart fills from below with a `>30d` block that swells through
summer and drains through winter. The battery chart is a thin band of short
durations churning all year.
:::

**6.11b — Why this matters for sampling.** Re-read the caution in @sec-sample,
then explain what a twelve-day sample would have done to the hydrogen result —
and what that implies for any study that samples representative days and then
reports on long-duration storage.

::: {.callout-tip collapse="true" title="Solution 6.11b"}
A twelve-day sample has no 30-day window, so the `>30d` band — **98.4% of this
store's energy** — cannot exist in it. Worse, the sample is twelve *disjoint*
days: with `fullYear = TRUE` the state of charge cycles over the slices present
and carries across the seams, so the model sees twelve short cycles rather than
one seasonal one.

The store would still appear in the results. It would simply be small, cheap and
busy — a large battery. Nothing would error, and no diagnostic would fire.

This is the honest limit of representative-period sampling, and it is worth
stating plainly whenever you use one: **a sample can only find storage cycles
shorter than the sample itself.** Studies that sample days and then report on
seasonal hydrogen are, quite often, reporting on an artefact of their own time
resolution.
:::

## How this model was derived {#sec-provenance}

The chain, so you can judge what the numbers mean:

1. **PyPSA-Eur** built a Belgian electricity network with its own Snakemake
   workflow: OpenStreetMap grid topology, a power-plant database, ERA5 and
   SARAH-3 weather reanalysis turned into hourly capacity factors by
   [atlite](https://atlite.readthedocs.io), and an hourly demand series. Belgium
   was clustered to **one node** and kept at its full **8,760 hours**.
2. **A converter** (`PyPSA-Eur-to-energyRt`, not part of this course) mapped each
   PyPSA component to an energyRt object: `Generator` → `technology` + a
   `weather` factor, `Load` → `demand`, `Line`/`Link` → `trade`, `StorageUnit`
   and `Store`+`Link` → `storage`, carriers → `commodity`.
3. The result was serialised to `data/belgium_copperplate.rds` (about 285 KB).

Three things you should know about it:

- **Nothing was changed to make it agree.** On the full year the weighting
  question does not arise: `year_fraction = 1`, so `pTimesliceWeight = 1` is
  simply correct, and both models charge annual capital against annual operating
  cost. The base case reproduces PyPSA **to the cent** (@sec-growth), which is
  the strongest statement available about the conversion.

  That is *not* true of a sampled model, and the difference is not small. The
  5-node one-week dataset shipped alongside (`belgium_model.rds`) keeps PyPSA's
  own `snapshot_weightings.objective = 1`, charging 168 hours of fuel against a
  full year of capital. Let energyRt annualise that week instead — the default
  it would otherwise choose — and capacity becomes cheap relative to fuel, so
  the model builds renewables until **no fossil plant dispatches at all**:

  | weighting | objective | CO₂ | fuels burned |
  |---|---:|---:|---|
  | PyPSA (weight 1) | 164,899,752 | 274,037 t | gas, nuclear, waste, biomass, oil |
  | annualised (52.14) | 245,652,738 | **0** | H₂ charge/discharge only |

  A model with no emissions has nothing for @sec-tax or @sec-cap to act on. On a
  **sampled** run the weighting convention is not a technicality — it decides
  whether the model has a fossil fleet at all. On a full year there is nothing to
  decide.
- **Transport, not power flow.** PyPSA imposes Kirchhoff's voltage law on AC
  cycles; energyRt's `trade` is a transport formulation. It makes no difference
  here — a copperplate has no lines — but it is the largest accepted semantic gap
  between the two tools, and @sec-replication measures it at 3.38% on a 41-node
  network.
- **Excluded entities.** The converter's policy is *detect, name, record,
  exclude* — never clamp, never substitute a plausible value. Whatever was
  dropped is recorded:

```{r}
readRDS("data/belgium_provenance.rds")   # the 5-node dataset's register
```

::: {.callout-important appearance="simple"}
**The model describes one year, 2050, and it does not know that.** Capacity rows
are keyed to `year = 2050` and non-extendable plant is closed off with
`vintage$end = 2049`. Extend the horizon earlier and both silently stop applying
— which is how @sec-horizon ended up building 10,999 MW of biomass against a
stated stock of 113 MW, and reporting zero emissions for a system burning 4.6
million units of gas.

This is the general hazard with any converted model: it carries the assumptions
of the run it came from. Audit the bounds before trusting a horizon it was not
built for.
:::

## Data licences {#sec-licences}

The dataset embeds real data, so its licences travel with it. PyPSA-Eur publishes
a per-source inventory (`doc/data_inventory.csv`); the electricity-only Belgian
build draws on these:

| source | used for | licence |
|---|---|---|
| technology-data (`costs`) | capital, fixed and variable costs | CC-BY-4.0 |
| powerplantmatching | existing plant, capacity, location | CC-BY-4.0 |
| ERA5 (Copernicus / ECMWF) | wind and temperature reanalysis | Copernicus licence — commercial use permitted, attribution required |
| SARAH-3 (EUMETSAT CM SAF) | solar irradiance | free use, attribution required |
| OpenStreetMap (`osm`, `osm_boundaries`) | grid topology, substations | **ODbL-1.0** |
| CORINE / Copernicus land cover | land eligibility | CC-BY-4.0 (CORINE: custom, CC-BY-like) |
| Natura 2000, EEZ, GEBCO | protected areas, maritime zones, bathymetry | CC-BY-4.0 / public domain |
| Eurostat NUTS 2021 | region geometry | EC reuse policy (2011/833/EU) |
| ENTSO-E / OPSD demand series | hourly electricity demand | **listed as "unknown"** |

Two entries need care. Neither is a reason not to use the data — only a reason to
know:

- **ODbL on OpenStreetMap.** ODbL is share-alike for *derived databases*. The
  grid capacities in the `trade` objects descend from OSM, so a redistributed
  derivative carries attribution and share-alike obligations. It is not a
  commercial restriction.
- **The demand series licence is genuinely unresolved.** PyPSA-Eur's own
  inventory records `entsoe_electricity_demand` and `opsd_electricity_demand` as
  licence **unknown**. Treat redistribution of the demand numbers as an open
  question rather than a settled one.

::: {.callout-important appearance="simple"}
**On commercial use.** The two entries in PyPSA-Eur's inventory that *do* restrict
commercial use or redistribution — Eurostat `lau_regions` ("permission to download
only if used for non-commercial purposes") and BGR `aquifer_data` ("right to use
without restriction but no right to redistribute") — are **sector-coupling inputs
and are not present in this electricity-only build**. This dataset therefore
carries no non-commercial clause. The one open item is the demand series'
*unknown* licence, which is an absence of information rather than a known
restriction.
:::

This table is a summary, not legal advice, and licences change. Check
`doc/data_inventory.csv` in the PyPSA-Eur version you actually use before
redistributing anything derived from it.

## Where this connects

You have now done, on real data, what the earlier chapters did on toys:
@sec-regions' trade became five Belgian nodes, @sec-processes' technologies became
a plant fleet, and @sec-time's storage became a battery and a hydrogen store.
What is new is everything *around* the model — measuring it, shrinking it, moving
it between solvers, stretching its horizon, and knowing where its numbers came
from.

That last part matters most. A model you cannot trace is a model you cannot
defend, and "it came from PyPSA-Eur" is the beginning of an answer, not the end
of one.
