---
output: html_document
editor_options:
  chunk_output_type: inline
---
# Replication of PyPSA-Eur in energyRt {#sec-replication}

This chapter is a case study rather than an exercise. It takes a model built in
another language, by another team, for another purpose, and rebuilds it in
energyRt — then checks whether the rebuilt model does the same arithmetic.

```{r}
library(energyRt)
library(reneuro)
library(dplyr)
library(ggplot2)
```

## Two model families, one problem {#sec-families}

**[PyPSA](https://pypsa.org)** is a capacity-expansion model of the electric
power sector, with extensions and links into other sectors. It is built for
energy-transition studies with electricity at the centre, and that focus is a
strength: it is an *ad-hoc and efficient* solution for exactly what it was made
to do. Python.

**energyRt** is a macro energy-system model **generator**. It is designed for
flexibility in describing technological processes and chains, for long-term
energy and environmental policy evaluation. R.

The two overlap enough to be comparable and differ enough to be interesting. The
translation exercise has three purposes, in order of importance:

1. **Demonstrate an energy-modelling workflow in R** using a model many people
   already know.
2. **Validate** that the translated model does the same mathematics. That is what
   the comparisons below are — a correctness check, not a benchmark and not a
   competition.
3. Give course participants a European model to take away and investigate.

There is a fourth thing worth naming up front, because it shapes everything in
@sec-price: converting an ad-hoc model with roughly **ten sets of constraints**
into a general system with roughly **a hundred pre-built ones** buys
expandability and a broader range of applications, and costs model size. That
cost was the intent from the beginning, not a defect discovered afterwards.

::: {.callout-note appearance="simple"}
This is **not** the first attempt.
[**revaluation**](https://opendecarbonization.github.io/revaluation/) (2023) translated PyPSA models into both energyRt and
[Switch](https://switch-model.org). This version was completed with the help of
AI — Claude, by Anthropic.
:::

## What was converted {#sec-mapping}

The conversion reads a PyPSA-Eur network at the `base_s_{clusters}_elec.nc` cut
point — after the network is built, before `prepare_network.py` applies policy
overlays and temporal averaging — and maps each component onto energyRt objects.

| PyPSA | energyRt | note |
|---|---|---|
| `Generator` | `technology` + `weather` | grouped by carrier; the time series becomes a weather factor |
| `Load` | `demand` | |
| `Line` (AC) | `trade` | merged by bus pair; **KVL not reproduced** |
| `Link` (DC) | `trade` | merged by bus pair |
| `Link` (converter) | folded into `storage` | charge / discharge efficiency |
| `StorageUnit` | `storage` | with inflow: decomposed into supply + storage + technology |
| `Store` | `storage` | reservoir |
| carriers | `commodity` (+ `supply`) | |

For the 41-node European network — 123 buses, 77 AC lines, 213 links, 397
generators, 59 storage units, 82 stores, 8,760 hourly snapshots — that produces
**143 energyRt objects**:

```{r}
eu <- pypsa_eur_41       # from reneuro; nothing to convert, nothing to download
summary(eu)
```

```
Regions: 41 (AL0 0, AT0 0, BA0 0, BE0 0, BG0 0, ...)
Objects: 143
 commodity  demand  storage  supply  technology  trade  weather
        13       1        4       9          17     92        7
```

The converter folds each `Store` and its two `Link`s into a single `storage`
object rather than emitting separate charger and discharger technologies, which
is why the technology count is 17 and not 21; @sec-storage-shape is what that
choice is about.

### Nothing disappears quietly {#sec-issues}

The conversion has one rule that matters more than any mapping decision:
**detect, name, record, exclude — never clamp, never invent.** A parameter that
cannot be represented is reported and its entity dropped; it is never quietly
replaced with a plausible number.

Every PyPSA row is therefore either mapped or listed as dropped. On the 41-node
network, 18 entities were excluded and 16 more kept with a warning:

| rule | severity | n | |
|---|---|---:|---|
| `link_zero_capacity` | blocker | 16 | transmission link with no capacity (TYNDP candidate project) |
| `storage_subhourly_duration` | blocker | 1 | duration below one hour; cannot cycle at hourly resolution |
| `storage_zero_duration` | blocker | 1 | reservoir with no storage capacity (`max_hours = 0`) |
| `line_zero_impedance` | warning | 13 | no impedance data; `teff` cannot be derived, set to 1 |
| `storage_multiyear_duration` | warning | 3 | duration exceeds a year; usually a small `p_nom` on a real reservoir |

This is the part of a conversion most worth reading. A model that silently
clamped a 28-second pumped-hydro plant to half an hour would build, solve, and
mean nothing.

## Seeing the model {#sec-seeing}

Once converted, the objects are ordinary energyRt objects and the usual tools
work on them.

`getObject()` fetches by name or by class, named without the repository prefix:

```{r}
draw(getObject(eu, name = "E_CCGT")[[1]])   # the chain: GAS in, ELC out
draw(getObject(eu, name = "STG_H2")[[1]])   # the three-role storage
```

```{r}
autoplot(getObject(eu, class = "supply")[[1]], interpolate = TRUE)

# `style = "line"`; the default "area" style needs a year to stack on, and this
# model carries `year = NA` as a wildcard, which yields "no rows to aggregate".
autoplot(getObject(eu, name = "DEM_ELC")[[1]], style = "line")
```

The model also carries **geography**. PyPSA-Eur keeps its region polygons in
GeoJSON sidecars rather than inside the `.nc`, and the converter turns them into a
`geoscales` hierarchy — `country -> bus`, 36 countries over 41 buses — attached to
the model. `plot_trade_map()` then needs no arguments:

```{r}
plot_trade_map(eu)
```

41 nodes and 90 trade routes over the region polygons. This is the quickest check
that the conversion is spatially sane: a name-matching or CRS error puts the
arrows in the sea rather than on the map.

The geometry is simplified for shipping (`rmapshaper::ms_simplify(keep = 0.03)`),
which takes the 41 regions from 2.91 MB compressed to 0.20 MB and is
indistinguishable at this scale.

Weather factors are where a converted model is easiest to sanity-check, because
mistakes are visible. `autoplot.weather()` draws a heatmap by default — hour of
day against day of year — so a solar profile that is bright at midnight tells you
immediately that something is wrong with the time mapping:

```{r}
w <- getObject(eu, class = "weather")

# `region` picks a subset: a number takes that many, or name them explicitly.
# Without it all 41 regions are faceted onto one plot and only the strip labels
# remain legible.
autoplot(w[["W_SOLAR"]], region = 2)          # heatmap is the default style
autoplot(w[["W_ONWIND"]], region = "BE0_0")
```

## Validation {#sec-validation}

Two models solving the same problem should reach the same number. Whether they do
is the whole question.

### Belgium, one node, full year {#sec-validation-be}

The cleanest test available is Belgium as a **copperplate**: one AC bus, no
lines, the full 8,760-hour year. It removes both of the things that could
otherwise excuse a difference — there is no transmission, so no
Kirchhoff-versus-transport gap, and it is a full year, so the annualisation
weighting is simply `1` on both sides.

| | energyRt | PyPSA | difference |
|---|---:|---:|---:|
| base | 8,260,626,305.45 | 8,260,626,305.45 | **0.00** |
| demand doubled | 13,387,554,321.05 | 13,387,554,321.06 | **0.01** |

Identical to the cent. The capacity-expansion answer matches too — **31,162.8 MW
built on both sides** under doubled demand, each technology within 0.05 MW
(@sec-growth works through this case as an exercise).

### Europe, 41 nodes {#sec-validation-eu}

The full European network at hourly resolution is about 28 million variables,
which no open solver will take. Sampling four representative days — mid-January,
April, July and October — gives 96 timeslices and ~326,000 variables, which
solves in under three minutes.

| | |
|---|---:|
| energyRt | 78,474,565,582 |
| PyPSA (transport) | 78,426,977,892 |
| **difference** | **+0.0607%** |

Every capacity matches to four decimal places, demand is exact, and
weather-driven generation agrees to eight significant figures — which is
independent evidence that both models are reading the same 96 hours.

::: {.callout-important appearance="simple"}
**Compare `objective_total`, not `objective`.** PyPSA leaves the capital cost of
non-extendable plant out of `n.objective` and reports it separately as
`objective_constant`. energyRt puts everything in one number. Add PyPSA's two
together before comparing, or the two models will appear to disagree by an order
of magnitude when they do not.
:::

## Three real differences {#sec-differences}

Where the models genuinely differ, they differ for reasons worth understanding.

### 1. Kirchhoff versus transport {#sec-kvl}

PyPSA imposes Kirchhoff's voltage law on AC cycles: power flows according to
impedance, not according to what would be cheapest. energyRt's `trade` is a
**transport** formulation — flow is bounded by capacity and otherwise free.

Transport is a relaxation, so it can only be cheaper. How much cheaper depends
entirely on the network:

| | KVL | transport | gap |
|---|---:|---:|---:|
| Belgium copperplate (0 lines) | — | — | **none possible** |
| Europe, 41 nodes (77 AC lines) | 675,336,409 | 652,485,728 | **3.38%** |

At one node the difference cannot exist. At 41 nodes it is 3.38%, and that is the
honest size of the paradigm gap between the two tools.

### 2. Snapshot weighting versus calendar annualisation {#sec-weighting}

This one is small in appearance and decisive in effect.

PyPSA scales operating cost by `snapshot_weightings.objective` and charges
capital as a full annual annuity regardless. On a *sampled* network PyPSA-Eur
leaves that weighting at **1**, so the model pays a year of capital against only
the sampled hours of fuel. energyRt instead derives `pTimesliceWeight` from the
calendar as `1/year_fraction`, which **annualises** operating cost.

On a full year the two coincide and the question never arises. On a 168-hour
sample of Belgium they do not, and the consequence is not subtle:

| weighting | objective | CO₂ | what runs |
|---|---:|---:|---|
| PyPSA's (w = 1) | 164,899,752 | 274,037 t | gas, nuclear, waste, biomass, oil |
| annualised (w = 52.14) | 245,652,738 | **0** | H₂ charge/discharge only |

The same model, the same data, one convention apart — and in one of them the
system has no fossil fleet at all and nothing for a carbon price to act on.
Neither is wrong. They answer different questions, and a comparison must state
which one it is asking.

### 3. `Store` + two `Link`s versus one storage {#sec-storage-shape}

PyPSA builds a hydrogen system from three components: a `Store` for the gas, a
`Link` for the electrolyser, and a `Link` for the fuel cell. energyRt's `storage`
carries three commodity roles — `@input`, `@storage`, `@output` — so the same
system can be one object.

A consequence worth knowing when reading results: **PyPSA rates a `Link` on its
input bus; energyRt rates a technology on its activity.** The Belgian
electrolyser confirms the rule exactly — PyPSA 162.058 MW × 0.6994 efficiency =
113.3434, which is energyRt's figure to four decimals.

## The price of generality {#sec-price}

energyRt solves the same problem with a substantially larger linear program:

| | rows | columns | after presolve |
|---|---:|---:|---|
| energyRt | 349,557 | 326,345 | 78,688 × 115,820 |
| PyPSA | 207,673 | 102,332 | 51,022 × 81,624 |

**3.2× the columns for identical physics**, and the same ratio shows up in time
and memory: ~146 s and 0.76 GB (R) + 1.93 GB (Julia) against PyPSA's ~12 s and
0.71 GB.

Presolve removes 64% of energyRt's columns, which tells you what they are:
**definitional**. energyRt materialises `vTechInp`, `vTechOut`, `vTechAct`,
`vInpTot`, `vOutTot`, `vTechInpTot` and their relatives as explicit variables
where PyPSA folds the same relationships into expressions.

That is the trade, and it was deliberate:

- **What it costs.** A larger LP, more memory, longer solves.
- **What it buys.** A hundred pre-built constraint sets instead of ten, so the
  same model can be extended into sectors, processes and policies the original
  was never built for — and every flow in the system is directly reportable
  through `getData()`, because it is a variable rather than an expression.

A model built for one job will always beat a generator at that job. The generator
earns its keep on the second job.

## Limitations {#sec-limitations}

Stated plainly, because this material is for training:

- **Training use only.** The validity checks reproduced results in the cases
  shown here; they are not a certification of the converter.
- **The full-year 41-node model has never been solved.** The European comparison
  rests on a four-day sample. A full-year continental solve is future work.
- **No deeper comparison has been done** — no sensitivity analysis, no scenario
  ensemble, no investigation of where the formulations diverge under stress.
- The four-day sample **cannot represent seasonal storage**: state of charge
  carries across the seams between disjoint days, so long-duration storage is
  understated by construction.

The model and converter are handed to participants for further investigation. If
you find something wrong, that is a useful outcome of the exercise.

## Reproduce it yourself {#sec-reproduce}

The model comes from reneuro; the issue register ships with the course:

```{r}
eu <- pypsa_eur_41                             # the converted 41-node model
getCalendar(eu)                                # 8760 timeslices, full year
readRDS("data/eu41_provenance.rds")            # coverage table + issue register
```

To solve it you need a sampled calendar, because the full-resolution model is
intractable. A model holds the **whole** calendar; a scenario may be solved on a
**subset** of its timeslices:

```{r}
yday_sample <- c(15, 105, 196, 288)            # mid-Jan / Apr / Jul / Oct
tt  <- data.table::as.data.table(getCalendar(eu)@timetable)
sub <- tt[YDAY %in% yday2YDAY(yday_sample) & HOUR %in% hour2HOUR(0:23)]
sub <- sub[, .SD, .SDcols = setdiff(names(sub), "weight")]

cal4 <- newCalendar("d365_h24_subset_4days",
                    timetable     = sub,
                    year_fraction = sum(sub$share))

scen <- interpolate_model(eu, cal4, name = "eu41_4d")
model_size(scen)
sol <- solve_scen(scen, solver = solver_options$julia_highs)
```

::: {.callout-warning appearance="simple"}
Two traps in `newCalendar()`, each of which costs a failed run:

- The timetable must stay a **data.table** — `.complete_calendar()` indexes
  `dtf[, weight]` with data.table semantics, and a plain `data.frame` fails with
  `object 'weight' not found`.
- **Keep `timeslice`, drop `weight`.** The check needs `timeslice`; but `weight`
  must be absent so the constructor recomputes `1/sum(share)` for the sample
  instead of inheriting the parent calendar's value. Passing the parent's weight
  through silently gives the sample full-year weighting — @sec-weighting is what
  that costs.
:::

Re-running the conversion itself is outside this course. reneuro's
[About](https://optimal2050.github.io/reneuro/articles/about.html) article
documents how each model was built and from which PyPSA-Eur commit, and
`attr(eu, "reneuro_provenance")` records it per object.

## Credits and licences {#sec-credits}

**Prior art.** [**revaluation**](https://opendecarbonization.github.io/revaluation/) (2023) translated PyPSA models into
energyRt and into Switch. This version was completed with the help of AI —
Claude, by Anthropic.

**Software.** PyPSA-Eur (MIT / CC-BY-4.0 / CC0-1.0 per component, see its
`REUSE.toml`), PyPSA, atlite, energyRt.

**Data.** The converted model embeds real data, so its licences travel with it.
The full table is in @sec-licences; the two entries that need care are repeated
here because they are easy to miss:

- **OpenStreetMap is ODbL-1.0.** The grid capacities in the `trade` objects
  descend from OSM, so a redistributed derivative carries attribution and
  share-alike obligations. Not a commercial restriction.
- **The demand series licence is unresolved.** PyPSA-Eur's own data inventory
  records `entsoe_electricity_demand` and `opsd_electricity_demand` as licence
  **unknown**.

The two entries in that inventory that *do* restrict commercial use or
redistribution — Eurostat `lau_regions` and BGR `aquifer_data` — are
sector-coupling inputs and are **not present** in this electricity-only build.

## Where this connects

@sec-realmodel used a converted Belgian model as a teaching object without asking
where its numbers came from. This chapter is that question answered: what was
mapped, what was dropped, what matches, and what does not.

The result worth remembering is not the agreement — two correct implementations
of the same linear program *should* agree. It is that establishing the agreement
required pinning down three conventions (network paradigm, snapshot weighting,
component rating) that are invisible until you compare, and any one of which
could have been mistaken for a modelling error.
