---
output: html_document
editor_options:
  chunk_output_type: inline
---
# USENSYS: a 26-zone US model {#sec-usensys}

The last chapter replicated a model converted from someone else's framework.
This one hands you a **finished national model**: USENSYS, the United States
electricity system in 26 zones, built for a multi-model intercomparison
(MIP) from the same [PowerGenome](https://github.com/PowerGenome/PowerGenome)
inputs that fed GenX, Switch and TEMOA. Real plant — 1,663 resources
aggregated to ~1,580 technologies — real hourly demand and weather for a
52-week year, 57 transmission corridors, storage, a national CO₂ cap with a
$200/t backstop, and six milestone years from 2027 to 2050.

You will not build it. You will do what modelers do with an inherited model:
load it, look at its map, weigh it, sample it, solve it, question it, and
change it.

```{r}
library(US26MIP)     # loads energyRt
library(dplyr)
library(ggplot2)
```

::: {.callout-note appearance="simple"}
Nothing executes at render (`eval: false` site-wide). The numbers quoted in
the text come from a local run (HiGHS barrier, no crossover, on a desktop
with 64 GB RAM) and are there so you can tell whether yours went right — not
as values to reproduce exactly.
:::

## One package, one model {#sec-us26-load}

The model ships **prebuilt** inside the `US26MIP` package (see the
installation chapter): the ~100 MB download *is* the model, stored the way
energyRt stores models on disk, with the big hourly tables loaded lazily.

```{r}
mod <- us26_model()
summary(mod@data[[1]])
mod@config@region
```

Twenty-six zones — the PowerGenome aggregation of EPA's IPM regions — with
the map attached:

```{r}
plot_geoscale(getGeoscale(mod), type = "map")
```

The package also ships the planning horizon and a family of calendars:

```{r}
us26_horizon@intervals$mid    # 2027 2030 2035 2040 2045 2050
names(us26_calendars)
```

| calendar | timeslices | what it is |
|---|---:|---|
| `w52_h168` | 8,736 | the full year (the model's own calendar) |
| `w8_h168` | 1,344 | 8 sampled weeks |
| `w4_h168` | 672 | 4 sampled weeks |
| `w2_h168` | 336 | 2 sampled weeks |
| `s4d_h24` | 96 | 4 seasonal representative days |

```{r}
# Full (52-weeks) calendar
autoplot(us26_calendars$w52_h168)
```

Sampling here is the same move as in @sec-realmodel: the model keeps its
full calendar; you pass a sampled one to `interpolate_model()` and the
timeslice weights annualize what is left.

```{r}
# 8 weeks calendar
autoplot(us26_calendars$w8_h168)
```

```{r}
# 8 weeks calendar sample on the 52-weeks calendar
autoplot(us26_calendars$w8_h168, reference = us26_calendars$w52_h168)
```


```{r}
report(mod, format = "html")
```



## Sampled solution {#sec-us26-size}

The polite habit from @sec-realmodel — `model_size()` before solving —
matters more here, and this model adds a second cost the Belgian one did not
have: **interpolation time**. Turning ~3,000 declared objects into a
scenario takes real minutes regardless of the calendar, because every
technology, storage and weather table is visited either way.

From the local run, for orientation:

| step | 4 seasonal days (96 ts) | full year (8,736 ts) |
|---|---|---|
| `interpolate_model()`, all 26 zones | ~35 min, ~25 GB RAM | ~35+ min, more RAM |
| HiGHS barrier solve, one year | ~2 min | hours |
| HiGHS barrier solve, 6-step myopic | ~4 h total | overnight+ |

Two consequences. First: on the full 26-zone model you interpolate **once**
and reuse the scenario; you do not iterate. Second: for the exercises, where
you *will* iterate, the package ships a subset accessor — `us26_wecc()`,
the six western-interconnection zones — which cuts every number above by
roughly a factor of four (@sec-us26-ex-solve).

::: {.callout-note appearance="simple"}
Use **Julia/HiGHS** (`solver_options$julia_highs`, or `_barrier` for the
interior-point method without crossover). GLPK is out of its depth at this
scale — keep it for the WECC subset on the 2-week calendar at most.
:::

## Solve one year {#sec-us26-solve}

The cheapest meaningful solve: 2027 only, on the four seasonal days.

```{r}
set_progress_bar("progress") 
show_progress_bar(TRUE)
scen <- interpolate_model(
  mod,
  us26_calendars$s4d_h24,
  newHorizon(period = 2027, intervals = 1),
  name = "us26-2027-s4d", 
  sparse = TRUE, fold = TRUE, # optimizes input data
  verbose = TRUE
)
summary(scen)
model_size(scen)
scen <- solve_scenario(scen, 
                       wait = F, # opens a cmd-window on Windows
                       solver = solver_options$julia_highs_barrier)
scen@path # where is your solution
scen <- read_solution(scen)

```

The local run lands on an objective of **131,331,824,562** (about $131 bn of
discounted system cost for the single year) in **~2 minutes** of solver time
after the ~35-minute interpolation.

What is running? Ask the solution:

```{r}
gen <- getData(scen, name = "vTechOut", comm = "ELC", merge = TRUE,
               process = TRUE, timeframe = "lowest") |>
  mutate(tech = case_when(
    grepl("nuclear", process) ~ "nuclear",
    grepl("natural_?gas|ngcc|ngct|_cc_|_ct_", process) ~ "gas",
    grepl("coal", process) ~ "coal",
    grepl("wind", process) ~ "wind",
    grepl("solar|photovoltaic|utilitypv", process) ~ "solar",
    grepl("hydro", process) ~ "hydro",
    .default = "other")) |>
  summarise(TWh = sum(value) / 1e6, .by = tech) |>
  arrange(desc(TWh))
gen

ggplot(gen, aes(TWh, reorder(tech, TWh))) +
  geom_col() +
  labs(y = NULL, title = "US generation, 2027 (4-day sample)")
```

And because the geoscale rides inside the model, results map without any
`sf` bookkeeping:

```{r}
plot_map(scen, name = "vTechNewCap", facet = "year")
```

## Six myopic steps to 2050 {#sec-us26-myopic}

The intercomparison ran USENSYS **myopically** — each milestone year solved
on its own, installed capacity carried forward — which is one call in
current energyRt:

```{r}
mod0 <- mod
mod0@config@discount$discount <- 0     # myopic runs use no social discounting

myo <- solve_myopic(
  mod0,
  "us26-myopic-s4d",
  us26_calendars$s4d_h24,
  horizon = us26_horizon,
  step = 1L,
  carry = "capacity",
  solver = solver_options$julia_highs_barrier
)
myo$steps
```

The local run solves all six steps in about four hours (the interpolation,
not the solver, is the cost — each step re-interpolates and then solves in
~2 minutes):

| step | year | objective |
|---:|---|---:|
| 1 | 2027 | 1.313 × 10¹¹ |
| 2 | 2030 | 6.823 × 10¹¹ |
| 3 | 2035 | 8.272 × 10¹¹ |
| 4 | 2040 | 9.181 × 10¹¹ |
| 5 | 2045 | 1.028 × 10¹² |
| 6 | 2050 | 1.609 × 10¹² |

(The per-step objectives are not comparable to a perfect-foresight run —
carried capacity is sunk and pays no investment annuity in later steps; use
`myopic_objective()` for a discounted total.)

## The exercise model: WECC {#sec-us26-wecc}

Everything below runs on the western interconnection — six zones, about a
quarter of the model, small enough to interpolate-change-solve in a coffee
break:

```{r}
wecc <- us26_wecc()
wecc@config@region
# "BASN" "CANO" "CASO" "NWPP" "RMRG" "SRSG"
```

One honest caveat: subsetting drops the DC interties to ERCOT and the
Eastern Interconnection (energyRt tells you so, line by line, as it
removes them), so WECC here is an island. For teaching, that is fine; for
analysis you would price the boundary (see `interpolate_model(...,
boundary_prices = )`).

## 9.1 — Solve WECC {#sec-us26-ex-solve}

**9.1 — Interpolate and solve.** One year (2027), four seasonal days, the
WECC subset. Report the objective and how long each step took.

::: {.callout-tip collapse="true" title="Solution 9.1"}
```{r}
scen_w <- interpolate_model(
  wecc,
  us26_calendars$s4d_h24,
  newHorizon(period = 2027, intervals = 1),
  name = "wecc-2027-s4d"
)
scen_w <- solve_scenario(scen_w, solver = solver_options$julia_highs)
scenario_runs(scen_w)$objective
```
The local run interpolates in **⟨WECC_INTERP⟩ minutes**, solves in
**⟨WECC_SOLVE⟩**, and lands on **⟨WECC_OBJ⟩** — your number should match to
solver tolerance.
:::

## 9.2 — Price the carbon differently {#sec-us26-ex-backstop}

**9.2 — Raise the backstop.** The base case caps national CO₂ but lets the
model buy its way out at $200/t (an export of `CO2` at a negative price —
find the `CO2EXP` object). Replace it with a $1,000/t backstop, re-solve
2027, and compare emissions.

::: {.callout-tip collapse="true" title="Solution 9.2"}
```{r}
CO2EXP1000 <- newExport("CO2EXP", commodity = "CO2",
                        exp = list(price = -1000))
scen_hi <- interpolate_model(
  wecc, CO2EXP1000,                       # overrides the model's CO2EXP
  us26_calendars$s4d_h24,
  newHorizon(period = 2027, intervals = 1),
  name = "wecc-2027-co2-1000"
)
scen_hi <- solve_scenario(scen_hi, solver = solver_options$julia_highs)

bind_rows(
  base = getData(scen_w,  name = "vBalance", comm = "CO2",
                 merge = TRUE, timeframe = "lowest"),
  hi   = getData(scen_hi, name = "vBalance", comm = "CO2",
                 merge = TRUE, timeframe = "lowest"),
  .id = "case") |>
  summarise(Mt = sum(value) / 1e6, .by = case)
```
In 2027 the fleet is essentially fixed (stock capacity, one build year), so
expect a *modest* drop — the interesting version of this exercise is
re-running the myopic sequence, where a 5× carbon price reshapes what gets
built after 2030.
:::

## 9.3 — More demand in California {#sec-us26-ex-demand}

**9.3 — +10 % demand in CANO.** Scale the 2027 hourly load in the CANO zone
up by 10 %, re-solve, and find out what serves it: which technologies
generate more, and does anything new get built (`vTechNewCap`)?

::: {.callout-tip collapse="true" title="Solution 9.3"}
```{r}
DEM <- wecc@data[[1]]@data[["DEMELC"]]
dem <- DEM@demand |>
  mutate(demand = if_else(region == "CANO", demand * 1.10, demand))
DEM2 <- update(DEM, demand = dem)

scen_d <- interpolate_model(
  wecc, DEM2,
  us26_calendars$s4d_h24,
  newHorizon(period = 2027, intervals = 1),
  name = "wecc-2027-dem110"
)
scen_d <- solve_scenario(scen_d, solver = solver_options$julia_highs)

getData(scen_d, name = "vTechNewCap", merge = TRUE,
        timeframe = "lowest") |>
  arrange(desc(value)) |>
  head(10)
```
Compare `vTechOut` by zone between `scen_w` and `scen_d` too: some of the
extra CANO load is met locally, some rides the WECC lines in.
:::

## 9.4 — Who generates, what is wasted {#sec-us26-ex-getdata}

**9.4a — Top five.** From the 9.1 solution, list the five technologies with
the largest 2027 generation.

**9.4b — Curtailment.** VRE availability is an *upper* bound
(`waf.up`-style), so what the wind and solar fleet *could* have produced and
what it *did* produce differ. Estimate the curtailed share.

::: {.callout-tip collapse="true" title="Solution 9.4"}
```{r}
out <- getData(scen_w, name = "vTechOut", comm = "ELC", merge = TRUE,
               process = TRUE, timeframe = "lowest")
out |> summarise(TWh = sum(value) / 1e6, .by = process) |>
  arrange(desc(TWh)) |> head(5)
```
Top five in the local run: **⟨WECC_TOP5⟩**.

```{r}
# curtailment: compare vre output with the balance surplus, or query the
# curtailment view directly:
crt <- getData(scen_w, name = "vBalance", comm = "ELC",
               merge = TRUE, timeframe = "lowest")
```
:::

## 9.5 — Draw it {#sec-us26-ex-plot}

**9.5a — Map.** Map 2027 generation by zone (`plot_map()`); try
`type = "capacity"` too.

**9.5b — A day in the west.** Pick the summer representative day and plot
hourly dispatch by technology group against demand for CASO.

::: {.callout-tip collapse="true" title="Solution 9.5"}
```{r}
plot_map(scen_w, type = "generation", comm = "ELC")

dis <- getData(scen_w, name = "vTechOut", comm = "ELC", merge = TRUE,
               process = TRUE) |>
  filter(region == "CASO", grepl("^w32", timeslice)) |>
  mutate(hour = as.integer(sub(".*_h", "", timeslice)))
ggplot(dis, aes(hour, value / 1e3, fill = process)) +
  geom_col(show.legend = FALSE) +
  labs(y = "GWh/h", title = "CASO dispatch, summer day (w32)")
```
:::

## 9.6 — How wrong is a sample? {#sec-us26-ex-calendar}

**9.6 — Sampling error.** Solve 9.1 again on `us26_calendars$w2_h168` (two
whole weeks) and on `w8_h168`. Compare the three objectives with the
four-day run. Which quantity moves most — the objective, the generation
mix, or storage utilization?

::: {.callout-tip collapse="true" title="Solution 9.6"}
```{r}
objs <- sapply(c("s4d_h24", "w2_h168", "w8_h168"), \(cal) {
  s <- interpolate_model(wecc, us26_calendars[[cal]],
                         newHorizon(period = 2027, intervals = 1),
                         name = paste0("wecc-2027-", cal)) |>
    solve_scenario(solver = solver_options$julia_highs)
  scenario_runs(s)$objective[1]
})
objs / objs[["s4d_h24"]]
```
The objective is the most forgiving of the three — annualized weights do a
lot of work — while storage behaviour changes qualitatively with the sample
(a 24-hour day cannot reward inter-day storage at all). That asymmetry is
the whole argument of @sec-time, now at national scale.
:::

## Where this connects

- @sec-realmodel did every one of these moves on a one-zone model you could
  re-solve in seconds — if something here felt rushed, redo it there.
- @sec-replication is the same "inherited model" story with a
  transmission-heavy European system and a validation target.
- The USENSYS model itself, its GenX provenance and the intercomparison
  results live in the [US26MIP](https://github.com/usensys/US26MIP)
  package documentation.
