---
title: "Installing energyRt"
# Optional backends — flip to `true` to include that section in the rendered
# book. GLPK (the default training backend) is always shown.
show-julia: false
show-python: false
show-gams: false
# Belt-and-suspenders: never execute install/solver code while rendering.
execute:
  eval: false
editor_options: 
  chunk_output_type: console
---

```{r}
#| include: false
#| eval: false
# Safety net: even if a chunk sets eval individually, nothing runs at render.
knitr::opts_chunk$set(eval = FALSE)
```

energyRt formulates an optimization model once and hands it to one of several
mathematical-programming **backends**. You only need *one* of them to get
started; you can add more later and cross-check results between them.

::: {.callout-important}
Every command below is shown for you to copy into your **own R session**. This
page renders without executing anything — no packages are installed while the
book builds.
:::

## Prerequisites

Install R (≥ 4.3) and, recommended,
[RStudio](https://posit.co/download/rstudio-desktop/) as the IDE for the
training.

::: {.panel-tabset}
### Windows
Download and run the installer from
[CRAN](https://cran.r-project.org/bin/windows/base/).

### macOS
Install the `.pkg` from [CRAN](https://cran.r-project.org/bin/macosx/), or with
[Homebrew](https://brew.sh/):

```bash
brew install --cask r
```

### Ubuntu
Install a current R from the CRAN apt repository (see the
[CRAN Ubuntu](https://cran.r-project.org/bin/linux/ubuntu/) guide):

```bash
sudo apt update
sudo apt install --no-install-recommends r-base
```
:::

## Quick install (recommended)

One line installs everything — `pak`, the system-library guidance, all
dependencies, and energyRt itself. Paste this into R:

```{r}
source("https://raw.githubusercontent.com/optimal2050/energyRt/master/inst/install.R")
install_energyRt()
```

On Linux it prints the exact `sudo` command for any missing system libraries (it
never runs `sudo` itself) — run that, then re-run `install_energyRt()`.

Once it finishes, verify your setup in a fresh session:

```{r}
library(energyRt)
en_setup()
```

`en_setup()` reports your OS, the system libraries to install (Linux), and the
full dependency status table.

## Step-by-step install (if the quick install fails)

Prefer to install by hand, or need to debug a failure? Install **pak** first —
the installer used throughout this guide:

```{r}
install.packages("pak")
```

### System libraries (Linux)

On Linux, energyRt's dependencies compile from source and need a few system
libraries. `pak` lists exactly which ones — install the reported apt packages
first so the R installs don't fail midway:

```{r}
pak::pkg_sysreqs("optimal2050/energyRt")   # lists the apt packages
```

```bash
# typical set on Debian/Ubuntu:
sudo apt install libcurl4-openssl-dev libssl-dev libxml2-dev
```

### R package dependencies

Install energyRt's CRAN imports up front — one by one, so any failure is
isolated and reported by name. Doing this first makes the energyRt install
itself quick and reliable, and avoids the Linux error
`dependencies '...' are not available for package 'energyRt'`.

```{r}
# energyRt's direct CRAN imports (base packages are omitted):
deps <- c(
  "generics", "data.table", "DBI", "RSQLite", "tibble", "tidyr", "dplyr",
  "rlang", "stringr", "lubridate", "purrr", "arrow", "progressr", "tictoc",
  "cli", "zoo", "registry", "options", "glue", "plyr",
  # suggested -- plots and reports (optional but recommended):
  "ggplot2", "patchwork", "knitr", "rmarkdown", "tinytex", "sf",
  # optional -- the energy-rhapsody finale (plus MuseScore, see below):
  "gm"
)

failed <- character(0)
for (pkg in deps) {
  ok <- tryCatch(
    {
      pak::pkg_install(pkg, ask = FALSE)
      message("  [ok]   ", pkg)
      TRUE
    },
    error = function(e) {
      message("  [FAIL] ", pkg, ": ", conditionMessage(e))
      FALSE
    }
  )
  if (!ok) failed <- c(failed, pkg)
}

if (length(failed) == 0) {
  message("All dependencies installed — you're ready to install energyRt.")
} else {
  message("Failed packages: ", paste(failed, collapse = ", "))
  message("Install their system libraries (above), then re-run.")
}
```

### Install energyRt

With the dependencies in place, installing energyRt is a quick one-liner:

```{r}
pak::pkg_install("optimal2050/energyRt")

# or, with remotes:
# install.packages("remotes")
# remotes::install_github("optimal2050/energyRt")
```

## Check what you already have

```{r}
library(energyRt)
en_check_dependencies()   # solver backends: GLPK / Julia / Python / GAMS / GDX
en_check_packages()       # R packages, training extras, LaTeX engine, MuseScore
```

`en_check_dependencies()` prints a status table — for each backend: installed?,
version, path, and a hint for what to do next — and tells you whether at least
one solver is ready. Each backend also has its own detector, e.g.
`en_check_glpk()`, `en_check_julia()`, `en_check_python()`, `en_check_pyomo()`,
`en_check_gams()`.

`en_check_packages()` covers the rest of the course toolkit: energyRt's own R
dependencies, the plotting/reporting extras (`ggplot2`, `patchwork`, `knitr`,
`rmarkdown`, `tinytex`, `sf`, `gm`), and the external tools they need — a LaTeX
engine for PDF reports and MuseScore for the `gm` music output. (`en_setup()`
runs both checks for you.)

## Choose a backend

| Backend | Software to install | License | Notes |
|---|---|---|---|
| **GLPK** | `glpsol` executable | open-source | Easiest to install; slow on very large models. Used in this training. |
| **Julia / JuMP** | Julia + `JuMP`, `HiGHS` | open-source | Fast (HiGHS barrier); recommended for large models. |
| **Python / Pyomo** | Python + `pyomo` + a solver (CBC) | open-source | Convenient if you already use conda. |
| **GAMS** | GAMS distribution | proprietary | Needs a license; also enables GDX I/O. |

For the training we use **GLPK** — it is open-source and the quickest to set up.
The other three backends are optional; each is covered below when enabled.

## GLPK (recommended for the training)

Install the `glpsol` executable.

::: {.panel-tabset}
### Windows
**Nothing to install.** `glpsol.exe` already ships with
[Rtools](https://cran.r-project.org/bin/windows/Rtools/) 4.5 — which you already
have, since Rtools is required to build energyRt from source. It lives at
`C:/rtools45/x86_64-w64-mingw32.static.posix/bin/glpsol.exe`.

### macOS
```bash
brew install glpk
```

### Ubuntu
```bash
sudo apt install glpk-utils
```
:::

Then confirm energyRt can find it. Set the path only if `glpsol` is not already
detected — on macOS/Ubuntu the package managers put it on your `PATH`
automatically:

```{r}
# Windows (Rtools 4.5): point at the bundled glpsol if not auto-detected.
# On macOS/Ubuntu this is usually unnecessary.
set_glpk_path("C:/rtools45/x86_64-w64-mingw32.static.posix/bin")
en_check_glpk()
```

::: {.content-visible when-meta="show-julia"}
## Julia / JuMP

Install Julia via [juliaup](https://github.com/JuliaLang/juliaup).

::: {.panel-tabset}
### Windows
```powershell
winget install julia -s msstore
```
(or the official installer).

### macOS
```bash
brew install juliaup
# or: curl -fsSL https://install.julialang.org | sh
```

### Ubuntu
```bash
curl -fsSL https://install.julialang.org | sh
```
:::

Then, from R (set the path only if Julia is not on your `PATH`):

```{r}
set_julia_path("~/.juliaup/bin/")
en_install_julia_pkgs()   # JuMP, HiGHS, Cbc, Clp, ... into the Julia env
en_check_julia()
```

::: {.callout-note collapse="true"}
## Manual alternative (if `en_install_julia_pkgs()` fails)

Install the same packages yourself from the Julia REPL (or `julia -e "..."`):

```julia
using Pkg
Pkg.add(["JuMP", "HiGHS", "Cbc", "Clp", "RData", "RCall",
         "CodecBzip2", "Gadfly", "DataFrames", "CSV", "SQLite", "Dates"])
```
:::
:::

::: {.content-visible when-meta="show-python"}
## Python / Pyomo

Install [Miniforge](https://github.com/conda-forge/miniforge) (recommended; it
gives you `conda`).

::: {.panel-tabset}
### Windows
Download and run the `Miniforge3` Windows installer. Avoid the Microsoft Store
"App execution alias" python — `en_check_python()` flags it as non-functional.

### macOS
```bash
brew install miniforge
# or run the Miniforge3 installer for your chip (arm64 / x86_64)
```

### Ubuntu
```bash
curl -L -O https://github.com/conda-forge/miniforge/releases/latest/download/Miniforge3-Linux-x86_64.sh
bash Miniforge3-Linux-x86_64.sh
```
:::

Then, from R:

```{r}
set_python_path("~/.conda/envs/energyRt")     # the env created below
en_install_python_deps()                       # conda env "energyRt": pyomo + cbc
# en_install_python_deps(use_conda = FALSE)    # or pip into the configured python
en_check_pyomo()
```

::: {.callout-note collapse="true"}
## Manual alternative (if `en_install_python_deps()` fails)

Create the environment and install the packages directly with conda (or mamba):

```bash
conda create -y -n energyRt python
conda install -y -n energyRt -c conda-forge pyomo pandas pyarrow coincbc
```

Or with `pip` — note pip cannot supply the CBC solver binary, so install a
Pyomo-compatible solver (e.g. CBC) separately:

```bash
python -m pip install --upgrade pyomo pandas pyarrow
```
:::
:::

::: {.content-visible when-meta="show-gams"}
## GAMS

Download and install GAMS from <https://www.gams.com/download/> (license
required), then point energyRt at the GAMS system directory:

::: {.panel-tabset}
### Windows
```{r}
set_gams_path("C:/GAMS/<version>/")
```

### macOS
```{r}
set_gams_path("/Library/Frameworks/GAMS.framework/Versions/Current/Resources")
```

### Ubuntu
```{r}
set_gams_path("/opt/gams/gams<version>_linux_x64_64_sfx")
```
:::

```{r}
en_check_gams()
```

GDX input/output uses `gdxtools`, which `en_install_deps()` installs
automatically when it detects GAMS.
:::

## Install the library layer

Once your runtime(s) are present, install the packages each backend needs in
one call:

```{r}
en_install_deps()   # detects runtimes, then installs the safe library layer
```

`en_install_deps()` runs a dependency check, installs only the library layer for
the runtimes it finds (skipping — with a warning — any that are missing), and
re-checks at the end.

::: {.callout-note collapse="true"}
## Manual alternative (if `en_install_deps()` fails)

`en_install_deps()` bundles the R, Julia, and Python layers — run them by hand:

**R layer** — the GAMS GDX bridge (not on CRAN) plus optional I/O helpers:

```r
pak::pkg_install("lolow/gdxtools")   # or remotes::install_github("lolow/gdxtools")
install.packages(c("jsonlite", "readxl", "openxlsx"))
```

**Julia** and **Python** layers — use the manual blocks in the *Julia / JuMP*
and *Python / Pyomo* sections above (enable them with the `show-julia` /
`show-python` header params).
:::

## Verify end-to-end

Re-check (everything you installed should now be green), choose a default
solver, and solve a tiny model:

```{r}
en_check_dependencies(solver_pkgs = TRUE)
set_default_solver(solver_options$glpk)   # or julia_highs_barrier, pyomo_cbc, ...

# a minimal model solve confirms the full toolchain works
# (see the modeling chapters for assembling a model)
```

If two or more backends are installed, solving the same scenario with each and
comparing objectives is the strongest check that your setup is correct.

## Optional: the energy-rhapsody finale 🎵

The closing chapter (`@sec-rhapsody`) *sonifies* scenarios — it turns a day of
dispatch into music. It needs two optional pieces; everything else in the
workshop runs without them.

```{r}
# 1. the gm package (composes the score)
install.packages("gm")

# 2. MuseScore (renders score + audio) — download from https://musescore.org
#    then, if gm cannot find it automatically, point gm at the executable:
# options(gm.musescore_path = "C:/Program Files/MuseScore 4/bin/MuseScore4.exe")
```

Without MuseScore you can still *compose* (`energy_rhapsody(scen, play =
FALSE)` returns the score object) — you just can't hear it.
