---
title: "R: setup and resources"
execute:
  eval: false
---

This chapter is for getting **R itself** ready and, if you need it, for
brushing up on the language. Everything energyRt-specific — the package, the
solver backends — is in the next chapter, [Installation](installation.qmd).

You do not need to be an experienced R user to follow the series. Familiarity
helps, but the sessions explain the modeling ideas as they go, and you are
welcome to attend simply to watch.

## Install R and an IDE

| What | Where | Needed? |
|---|---|---|
| **R** (≥ 4.3) | <https://cloud.r-project.org/> | Required |
| **RStudio Desktop** — the IDE we use in the sessions | <https://posit.co/downloads> | Strongly recommended |
| **Quarto** — publishing system for reports and slides | <https://quarto.org/> | Optional |

Install R **first**, then RStudio: RStudio is a front end and looks for an
existing R installation when it starts.

Recent versions of RStudio bundle Quarto, so a separate install is usually
unnecessary — it is listed here because Quarto is worth knowing about in its own
right, and the course materials are written in it.

::: {.callout-note}
## Windows: you also want Rtools

energyRt is installed from source, which needs a compiler toolchain. On Windows
that is [Rtools](https://cran.r-project.org/bin/windows/Rtools/) — install the
version matching your R (Rtools 4.5 for R 4.5.x). A useful side effect: Rtools
ships the GLPK solver we use in the training, so installing it covers two
requirements at once. See [Installation](installation.qmd).
:::

## Learning (or refreshing) R

Many good introductions exist; these are a few short ones that get you to a
working knowledge quickly. Pick one — they cover much the same ground.

- [**Introduction to R**](https://intro2r.com/) — a gentle, complete beginner's
  book: installing, data structures, plotting, basic programming.
- [**R intro (Otoomet, INFO 201)**](https://faculty.washington.edu/otoomet/info201-book/r-intro.html)
  — compact and fast-moving; good if you program already and just want R's
  idioms.
- [**R for Data Science (2e)**](https://r4ds.hadley.nz/) — the standard
  reference for the tidyverse workflow that energyRt's data handling follows.
- [**Hands-On Programming with R**](https://rstudio-education.github.io/hopr/) —
  focuses on the language itself (objects, functions, environments) rather than
  data analysis.
- [**swirl**](https://swirlstats.com/) — learn interactively inside the R
  console, if you prefer doing to reading.

Time is short before the first session — the highest-return preparation is being
comfortable with data frames, `|>` pipes, and subsetting. Everything else you
can pick up as we go.

## Key packages to check first

Install these before Session 1 and confirm they load. They are the ones most
likely to need a compiler or a system library, so finding a problem now rather
than mid-session is worth the few minutes:

```{r}
install.packages(c("dplyr", "ggplot2", "sf"))
```

```{r}
library(dplyr)     # data manipulation — the verbs energyRt returns results for
library(ggplot2)   # plots
library(sf)        # spatial data and maps
```

- **dplyr** — results come back from energyRt as tidy tables; `filter()`,
  `group_by()`, `summarise()` are how you work with them.
- **ggplot2** — used throughout for charts of capacity, dispatch, and costs.
- **sf** — spatial features, for models with regions and maps. It is the one
  most likely to complain, since it links to the GDAL/GEOS/PROJ system
  libraries. On Windows and macOS the CRAN binary includes them; on Linux
  install `libgdal-dev libgeos-dev libproj-dev` (or your distribution's
  equivalents) first.

If `library(sf)` fails, it will not block you from the modeling parts of the
course — only from the mapping examples.

## Cheatsheets

Posit maintains one-page reference cards for most of the tools we use. Printing
or bookmarking a few is the single most useful thing you can do before the
sessions:

**<https://opensource.posit.co/resources/cheatsheets/>**

Worth having open during the course:

- **Data transformation with dplyr** — the verbs, joins, and grouping
- **Data visualization with ggplot2** — geoms, scales, facets
- **RStudio IDE** — panes, shortcuts, projects
- **R Markdown** and **Quarto** — for writing up your own results
- **sf / Simple Features** — if you get to the spatial examples

## Getting help

- `?function_name` in the console opens R's own documentation — always the
  fastest first stop.
- Bring anything unresolved to the [course Q&A thread](index.qmd#questions-and-discussion) —
  it is there precisely for this, and beginner questions are welcome.
