---
title: "Building blocks of an energyRt model"
subtitle: "The `new*()` constructors, and the system around them"
author: "Oleg Lugovoy, et al."
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    transition: fade
    footer: "energyRt · useR! Training"
    fig-align: center
    # Reveal lays every slide out on this virtual canvas and scales it to the
    # screen -- a SMALLER canvas makes everything look BIGGER. This is the one
    # number to tune if the text is the wrong size.
    width: 1050
    height: 720
    margin: 0.06
    code-block-font-size: 0.52em
    # Safety net: a slide that still overflows scrolls instead of being cut off.
    # Slides that fit are unaffected.
    scrollable: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| echo: false
library(energyRt)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 5.2, fig.height = 4.2, dpi = 130)
```

```{=html}
<style>
.badge {
  font-size: 0.62em; font-weight: 600; letter-spacing: .04em;
  padding: .10em .45em; border-radius: .35em; vertical-align: middle;
  text-transform: uppercase; white-space: nowrap;
}
.exp     { background: #fdf0d5; color: #8a5a00; border: 1px solid #e8c98a; }
.planned { background: #eef2f7; color: #3d5a80; border: 1px solid #b8c9dd; }
.limited { background: #fbe9e7; color: #9c3a2e; border: 1px solid #e8b4ac; }

/* Buy vertical room from SPACING, not from type size, so the text stays big. */
.reveal h2            { font-size: 1.5em; margin-bottom: .35em; }
.reveal section p     { margin: .38em 0; line-height: 1.28; }
/* Only touch top/bottom: a shorthand `margin: .25em 0` would zero the LEFT
   margin too, and that is where ordered-list numbers are drawn -- they end up
   outside the column box and get clipped. */
.reveal ul, .reveal ol { margin-top: .25em; margin-bottom: .25em; line-height: 1.3; }
.reveal ol            { list-style-position: outside; }
.reveal li            { margin-top: .18em; margin-bottom: .18em; }
.reveal li > ul       { margin-top: .1em; margin-bottom: .1em; }
/* Columns clip their overflow, so give a list inside one room for its markers. */
.reveal .column ol,
.reveal .column ul    { padding-left: 1.2em; }
.reveal .columns      { gap: .8em; }
.reveal pre           { margin: .45em 0; box-shadow: none; }
.reveal pre code      { padding: .5em .7em; line-height: 1.24; }
.reveal blockquote    { margin: .5em 0; }
.reveal .callout      { margin: .5em 0; font-size: .88em; }

/* Tables: readable by default; .small-table only for the reference matrices. */
.reveal table              { font-size: .9em; margin: .4em 0; }
.reveal table th,
.reveal table td           { padding: .18em .5em; }
.reveal .small-table table { font-size: .74em; }

/* `{.smaller}` is reveal's 0.6x. That is too aggressive here -- soften it so
   dense slides shrink a little instead of a lot. */
.reveal section.smaller > h2 { font-size: 1.4em; }
.reveal .smaller           { font-size: 0.86em; }
</style>
```

::: {.notes}
Reference deck. Parts 1 and 6 are the live-teaching core; the rest is written to
be browsable afterwards. Badges: experimental / planned / backend-limited.
:::

# 1 · The bricks {background-color="#f4f4f4"}

## The vocabulary {.smaller}

An energyRt model is a **set of objects**, each built by a `new*()` function.

::: {.columns}
::: {.column width="50%"}
**Commodities** — the things that flow

- `newCommodity()`

**Processes** — the things that move or transform them

- `newSupply()`, `newImport()`, `newExport()`
- `newDemand()`
- `newTechnology()`
- `newStorage()`, `newTrade()`
:::
::: {.column width="50%"}
**Containers** — how they are assembled

- `newRepository()` — a bag of objects
- `newHorizon()` — the years
- `newModel()` — objects + regions + horizon

**Policy** *(later chapters)*

- `newConstraint()`, `newTax()`, `newSubsidy()`
:::
:::

. . .

Every process can be **drawn**: `draw(obj)` sketches its flows. We use it on
every slide from here on.

## One grammar, everywhere {.smaller}

Every constructor follows the same shape:

```r
newTechnology(
  name    = "ECOA",              # the set element -- UPPERCASE, no spaces
  desc    = "Coal power plant",  # free text
  input   = list(comm = "COA"),  # what it consumes
  output  = list(comm = "ELC"),  # what it produces
  ceff    = data.frame(comm = "COA", cinp2use = 0.4),   # parameters
  region  = character()          # empty = every region
)
```

- **`name`** is the identifier the solver sees.
- **Parameters are `data.frame`s**, and may carry `region`, `year`, `slice`
  columns. Leave a column out and the value applies to *all* of them.
- Values between given years are **interpolated** — you supply anchors, not a
  full grid.

## `newCommodity()` — what flows {.smaller}

A commodity is a *label with a unit*. It has no diagram: it is the thing that
travels along the arrows, not a box.

```{r}
COA <- newCommodity(
  name      = "COA",
  desc      = "Coal",
  unit      = "PJ",
  timeframe = "ANNUAL",                     # ANNUAL | HOUR | ...
  emis      = data.frame(comm = "CO2",      # released when burned
                         unit = "kt/PJ",
                         emis = 95)
)
```

- **`timeframe`** — the finest resolution at which this commodity balances.
  `ANNUAL` pools the year; `HOUR` forces hourly balancing.
- **`emis`** — emission factors travel with the *fuel*, so every technology
  that burns it inherits them (given `combustion = 1` on the input).

## `newSupply()` — a domestic resource {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
SUP_COA <- newSupply(
  name      = "SUP_COA",
  desc      = "Coal supply",
  commodity = "COA",
  unit      = "PJ",
  supply    = data.frame(
    cost   = 2.5,      # MEUR/PJ
    ava.up = 500       # cap, PJ/year
  )
)
```

`supply` accepts `cost` and the availability bounds `ava.lo` / `ava.up` /
`ava.fx`, plus optional `region`, `year`, `slice`.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(SUP_COA)
```
:::
:::

## `newDemand()` — what must be served {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
DEM_ELC <- newDemand(
  name      = "DEM_ELC",
  desc      = "Electricity demand",
  commodity = "ELC",
  unit      = "PJ",
  demand    = data.frame(
    year   = c(2025, 2050),
    demand = c(50, 80)
  )
)
```

Demand is **exogenous**: it is a constraint, not a choice. Values between
2025 and 2050 are interpolated.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(DEM_ELC)
```
:::
:::

## `newImport()` — a flow from outside {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
IMP_GAS <- newImport(
  name      = "IMP_GAS",
  desc      = "LNG import",
  commodity = "GAS",
  unit      = "PJ",
  import    = data.frame(
    price  = 9.0,     # MEUR/PJ
    imp.up = 200      # cap, PJ/year
  ),
  reserve   = 3000    # cumulative cap
)
```

`newExport()` is its mirror: `export`, `price`, `exp.up`.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(IMP_GAS)
```
:::
:::

## `newTechnology()` — the converter {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
ECOA <- newTechnology(
  name    = "ECOA",
  desc    = "Coal power plant",
  input   = list(comm = "COA", unit = "PJ",
                 combustion = 1),
  output  = list(comm = "ELC", unit = "PJ"),
  ceff    = data.frame(comm = "COA",
                       cinp2use = 0.40),
  invcost = list(invcost = 2000),  # MEUR/GW
  fixom   = 55, cap2act = 31.536,
  olife   = 30L
)
```
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(ECOA)
```
:::
:::

Read the diagram left to right: **`inp` → `use` → `act` → `out`**.

## The five stages {.smaller}

Every technology has the same internal chain. Each arrow is a coefficient:

| Stage | Meaning | Coefficient |
|---|---|---|
| `inp` | commodity entering | — |
| `ginp` | *grouped* input, if declared | `cinp2ginp` |
| `use` | what the process actually consumes | `cinp2use`, `ginp2use` |
| `act` | activity — the capacity-linked variable | `use2cact` |
| `out` | commodity leaving | `cact2cout` |

. . .

- **`cap2act`** converts capacity into annual activity
  (1 GW × 8760 h = 31.536 PJ).
- Give **one** efficiency coefficient and the rest default to 1 — that is why
  `cinp2use = 0.40` alone describes a 40 % plant.

## Grouped input: co-firing {.smaller}

::: {.columns}
::: {.column width="58%"}
Two fuels feeding **one** boiler. The group is the thing with an efficiency;
the members compete for share within it.

```{r}
ECOFIRE <- newTechnology(
  name   = "ECOFIRE",
  input  = list(comm = c("COA", "BIO"),
                unit = c("PJ", "PJ"),
                group = c("FUEL", "FUEL"),
                combustion = c(1, 1)),
  output = list(comm = "ELC", unit = "PJ"),
  group  = data.frame(group = "FUEL",
                      desc = "Boiler fuel", unit = "PJ"),
  geff   = data.frame(group = "FUEL",
                      ginp2use = 0.38),
  ceff   = data.frame(comm = c("COA", "BIO"),
                      cinp2ginp = c(1, 1),
                      share.up  = c(1, 0.30)),
  cap2act = 31.536, olife = 30L
)
```
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(ECOFIRE)
```
:::
:::

## Grouped input: what the picture tells you {.smaller}

::: {.columns}
::: {.column width="52%"}
- The **`ginp` column appears** — it is absent for a single-fuel technology.
- `FUEL` is a node: both fuels flow into it, and *it* carries the efficiency
  (`ginp2use: 0.38`).
- `[0, 0.30]` next to BIO is the **share bound**: biomass may supply at most
  30 % of the boiler's fuel.
- `[0, 1]` next to COA: coal may supply all of it.

. . .

**Why group at all?** Without a group each fuel would need its own efficiency
and could not be constrained *relative to the other*. Shares are the point.
:::
::: {.column width="48%"}
```{r}
#| echo: false
draw(ECOFIRE)
```
:::
:::

## Multiple outputs: CHP {.smaller}

::: {.columns}
::: {.column width="58%"}
One activity, two products. `cact2cout` sets how much of each comes out per
unit of activity.

```{r}
ECHP <- newTechnology(
  name   = "ECHP",
  desc   = "Gas CHP: power and heat",
  input  = list(comm = "GAS", unit = "PJ",
                combustion = 1),
  output = list(comm = c("ELC", "HEA"),
                unit = c("PJ", "PJ")),
  ceff   = data.frame(
    comm      = c("ELC", "HEA"),
    cact2cout = c(1, 1.2)      # heat/power ratio
  ),
  cap2act = 31.536, olife = 25L
)
```

Fixing both makes the ratio rigid; bound them instead for a flexible CHP.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(ECHP)
```
:::
:::

## Auxiliary commodities {.smaller}

**Main** commodities (red) are tied to capacity and activity. **Auxiliary**
commodities (blue) are anything else the process consumes or emits — land,
water, steel, labour, pollutants.

They are declared in `aux`, and linked in `aeff` by choosing *which* quantity
they scale with:

| Coefficient | Scales with |
|---|---|
| `cinp2ainp`, `cinp2aout` | main commodity **input** |
| `cout2ainp`, `cout2aout` | main commodity **output** |
| `act2ainp`, `act2aout` | **activity** |
| `cap2ainp`, `cap2aout` | **installed capacity** |
| `ncap2ainp`, `ncap2aout` | **new capacity** (built this year) |

. . .

`…ainp` = the process *consumes* it, `…aout` = the process *produces* it.

## Aux linked to capacity, activity, new capacity {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
ENUC <- newTechnology(
  name   = "ENUC",
  desc   = "Nuclear: land, water, steel",
  input  = list(comm = "NUC", unit = "PJ"),
  output = list(comm = "ELC", unit = "PJ"),
  aux    = data.frame(
    acomm = c("LAND", "WAT", "STEEL"),
    unit  = c("km2", "Mm3", "kt")),
  aeff   = data.frame(
    acomm     = c("LAND", "WAT", "STEEL"),
    cap2ainp  = c(0.5,  NA,   NA),   # per GW held
    act2ainp  = c(NA,   0.02, NA),   # per PJ made
    ncap2ainp = c(NA,   NA,   40)    # per GW built
  ),
  ceff = data.frame(comm = "NUC", cinp2use = 0.35),
  cap2act = 31.536, olife = 50L
)
```
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(ENUC)
```
:::
:::

## Why the distinction matters {.smaller}

::: {.columns}
::: {.column width="52%"}
Three physically different things, and the diagram labels each one:

- **`cap2ainp: 0.50`** — land is occupied for as long as the plant *stands*,
  whether or not it runs.
- **`act2ainp: 0.02`** — cooling water is drawn only when it *generates*.
- **`ncap2ainp: 40.00`** — steel is consumed **once**, in the year the plant is
  *built*.

. . .

Pick the wrong one and the model is quietly wrong: steel on `cap2ainp` would
re-buy the whole plant every year.
:::
::: {.column width="48%"}
```{r}
#| echo: false
draw(ENUC)
```
:::
:::

## Aux as emissions {.smaller}

::: {.columns}
::: {.column width="58%"}
Pollutants that depend on the *plant*, not only the fuel, are auxiliary
**outputs** scaled by fuel input:

```{r}
ECOA_E <- newTechnology(
  name   = "ECOA_E",
  desc   = "Coal plant with NOx and SOx",
  input  = list(comm = "COA", unit = "PJ",
                combustion = 1),
  output = list(comm = "ELC", unit = "PJ"),
  aux    = data.frame(acomm = c("NOX", "SOX"),
                      unit  = c("kt", "kt")),
  aeff   = data.frame(acomm = c("NOX", "SOX"),
                      comm  = "COA",
                      cinp2aout = c(0.30, 0.45)),
  ceff = data.frame(comm = "COA", cinp2use = 0.40),
  cap2act = 31.536, olife = 30L
)
```

Contrast with CO~2~, which sits on the **commodity** — every burner inherits it.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(ECOA_E)
```
:::
:::

## `newStorage()` — moving energy in time {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
STG_BTR <- newStorage(
  name      = "STG_BTR",
  desc      = "4-hour battery",
  commodity = "ELC",
  seff      = data.frame(inpeff = 0.92,
                         outeff = 0.92),
  cap2stg   = 4,          # GWh per GW
  invcost   = list(invcost = 800),
  olife     = 15L
)
```

- `seff` — charge / discharge efficiency, and `stgeff` for standing losses.
- `cap2stg` — how many hours of storage one unit of capacity buys.
- Only meaningful with **sub-annual slices** — see the *Time and storage*
  chapter.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(STG_BTR)
```
:::
:::

## `newTrade()` — moving energy in space {.smaller}

::: {.columns}
::: {.column width="58%"}
```{r}
TBD_ELC <- newTrade(
  name      = "TBD_ELC",
  desc      = "Interregional power line",
  commodity = "ELC",
  routes    = data.frame(          # which links exist
    src = c("R1", "R2"),
    dst = c("R2", "R1")
  ),
  trade     = data.frame(          # their parameters
    src  = c("R1", "R2"),
    dst  = c("R2", "R1"),
    teff = 0.95                    # line losses
  ),
  invcost = list(invcost = 1500)
)
```

Two slots, two jobs: **`routes`** declares the links, **`trade`** parameterises
them. One is not derived from the other — a route with no parameters is
unconstrained, and parameters for an undeclared route are ignored.

Trade is **directional**, so a bidirectional line is *two* rows in each. Note it
has no `region` — the regions are `src` and `dst`.
:::
::: {.column width="42%"}
```{r}
#| echo: false
draw(TBD_ELC)
```
:::
:::

## `newRepository()` — the bag {.smaller}

A repository is an ordered collection of objects. It has no behaviour of its
own; it exists so a model can be assembled from named parts.

```{r}
repo <- newRepository(
  name = "power",
  COA, SUP_COA, ECOA, ECOFIRE, ECHP, ENUC, STG_BTR, DEM_ELC
)
```

```{r}
#| echo: false
cat("class:", class(repo), "\n")
cat("objects:", length(repo), "\n")
```

- Repositories can be **combined** and passed around, so a scenario is
  "base repo + a few replacements".
- `add()` / `update()` return a *modified copy* — nothing is mutated in place.

## `newHorizon()` and `newModel()` {.smaller}

```{r}
mod <- newModel(
  name     = "POWER",
  desc     = "A small power system",
  data     = repo,
  region   = "R1",
  discount = 0.05,
  horizon  = newHorizon(
    period     = 2025:2050,
    intervals  = c(1, 5, 5, 5, 5, 5), # years per period
    mid_is_end = TRUE
  )
)
```

- **`region`** — the model's regions; objects with no `region` apply to all.
- **`horizon`** — milestone years. Periods grow, so the near term is resolved
  finely and the far term cheaply.
- The model is still just a **description**. `interpolate_model()` expands it
  over regions × years × slices; `solve_scenario()` solves it.

## Putting it together {.smaller}

| Function | Makes | Drawn? | Key arguments |
|---|---|:--:|---|
| `newCommodity()` | a flow | — | `unit`, `timeframe`, `emis` |
| `newSupply()` | a resource | ✔ | `supply`: `cost`, `ava.up` |
| `newDemand()` | an obligation | ✔ | `demand` |
| `newImport()` / `newExport()` | rest-of-world flow | ✔ | `price`, `imp.up` / `exp.up` |
| `newTechnology()` | a converter | ✔ | `input`, `output`, `ceff`, `aeff`, `invcost` |
| `newStorage()` | time shifting | ✔ | `seff`, `cap2stg` |
| `newTrade()` | space shifting | ✔ | `routes`, then `trade`: `src`, `dst`, `teff` |
| `newRepository()` | a bag of objects | — | `name`, `...` |
| `newModel()` | the description | — | `data`, `region`, `horizon` |

. . .

**When in doubt, `draw()` it.** If the picture is not the system you meant, the
model is not either.

# 2 · Where objects live: time {background-color="#f4f4f4"}

## Calendars: the time structure {.smaller}

A **calendar** declares the sub-annual structure every flow is balanced on. It
is a set of nested *timeframes*, each split into *slices* carrying a share of
the year.

```{r}
make_timetable(list(SEASON  = c("WI", "SP", "SU", "FA"),
                    DAYNITE = c("D", "N")))  |> head(4)
```

- Nesting is the grammar: `SEASON` above `DAYNITE` gives 8 slices.
- Every parent timeframe currently needs the **same number of children** — a
  regular tree, not a ragged one.
- `newCalendar(name, timetable, year_fraction)` turns a timetable into an object.

## Ready-made calendars {.smaller}

Eight ship with the package — the fastest way to change a model's time
resolution is to swap one in.

```{r}
names(calendars)
```

::: {.columns}
::: {.column width="46%"}
```{r}
#| echo: false
#| fig-width: 4.6
#| fig-height: 3.6
autoplot(calendars$utopia_seasons)
```
:::
::: {.column width="54%"}
```{r}
#| echo: false
#| fig-width: 5.4
#| fig-height: 3.6
autoplot(calendars$utopia_s4h24)
```
:::
:::

Each rectangle is one slice, sized by its share of the year. `ANNUAL` sits on
top; finer timeframes stack below.

## Sampled vs full calendars {.smaller}

::: {.columns}
::: {.column width="52%"}
A **sampled** calendar keeps a representative subset of slices and declares what
fraction of the year it covers:

```{r}
sub  <- calendars$d365_h24_subset_1day_per_month
full <- calendars$d365_h24

data.frame(slices        = c(nrow(sub@timetable), nrow(full@timetable)),
           year_fraction = round(c(sub@year_fraction, full@year_fraction), 4),
           row.names     = c("sampled", "full"))
```

One day per month at hourly resolution — **288 slices instead of 8760**, about
3.3 % of the year.

`year_fraction` is what keeps annual quantities honest when only part of the
year is modelled.
:::
::: {.column width="48%"}
```{r}
#| echo: false
#| fig-width: 5.0
#| fig-height: 4.4
autoplot(sub, reference = full)
```

Passing `reference =` lays out the **full** structure and fills only the slices
the sampled calendar declares.
:::
:::

## Aggregating time in results {.smaller}

Sampling reduces what is *solved*. Aggregation changes what you *read back* —
`getData()` takes a `timeframe` argument:

| `timeframe =` | Returns |
|---|---|
| `"lowest"` *(default)* | flows summed up to the coarsest level, normally `ANNUAL` |
| `"highest"` | native resolution, as stored |
| `"all"` | every level, stacked |
| `"SEASON"`, `"YDAY"`, … | one named calendar level |

. . .

State variables (a storage level, say) are **returned unchanged** — summing a
level over slices is meaningless, and the package knows the difference.

::: {.callout-note}
## Roadmap [planned]{.badge .planned}
Nested timeframes and slice aggregation/disaggregation as a first-class
operation move to the **timescales** package — the temporal companion to
`geoscales`. Today energyRt's calendar machinery is self-contained.
:::

## Horizons: the years {.smaller}

::: {.columns}
::: {.column width="55%"}
```{r}
h <- newHorizon(period    = 2025:2050,
                intervals = c(1, 5, 5, 5, 5, 5))
h@intervals
```

- **`period`** — the full span of years.
- **`intervals`** — either lengths (as here) or a `data.frame` with
  `start`, `mid`, `end`.
- The first interval is forced to a single **base year**
  (`force_BY_interval_to_1_year = TRUE`).
- The **mid** is the milestone year the solver actually optimises.
:::
::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 5.0
#| fig-height: 3.8
autoplot(h)
```
:::
:::

## Where the milestone year sits {.smaller}

The same intervals, three placements of the mid-year — the dashed line:

```{r}
#| echo: false
#| fig-width: 13
#| fig-height: 3.6
library(patchwork)
p1 <- autoplot(newHorizon(period = 2025:2050, intervals = c(1, 5, 5, 5, 5, 5))) +
  labs(title = "default: midpoint")
p2 <- autoplot(newHorizon(period = 2025:2050, intervals = c(1, 5, 5, 5, 5, 5),
                          mid_is_start = TRUE)) + labs(title = "mid_is_start = TRUE")
p3 <- autoplot(newHorizon(period = 2025:2050, intervals = c(1, 5, 5, 5, 5, 5),
                          mid_is_end = TRUE))   + labs(title = "mid_is_end = TRUE")
p1 + p2 + p3
```

. . .

It matters because **costs and demands are evaluated at the mid**. `mid_is_end`
puts the decision at the end of each period — the convention used in the models
we build here. Choose once, and keep it consistent across scenarios you compare.

## Ready-made horizons {.smaller}

```{r}
names(horizons)
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 3.2
autoplot(horizons$Y2020_2060_by_5) + labs(title = "horizons$Y2020_2060_by_5")
```

Single-year horizons (`Y2030`, `Y2050`, …) are useful for a fast structural
check: build the model, solve one year, confirm it is feasible before paying for
the full run.

# 3 · Where objects live: space {background-color="#f4f4f4"}

## Regions {.smaller}

There is **no `newRegion()`**. Regions are a flat character vector declared on
the model:

```{r}
#| eval: false
newModel("POWER", data = repo, region = c("R1", "R2", "R3"))
```

- Objects carry an optional `region` slot; **leaving it empty means every
  region**.
- `trade` is the exception — it has no `region`, only `src` and `dst`.
- `get_region(obj)` walks any object and reports the regions it refers to. It is
  deliberately schema-agnostic, so it keeps working as region information is
  added to classes that do not carry it yet.

```{r}
get_region(TBD_ELC)
```

## Geoscales: nesting regions {.smaller}

A **geoscale** (from the companion `geoscales` package) describes how regions
nest into coarser levels, their weights, and optionally their geometry.

```{r}
gs <- utopia_geoscale()   # UTOPIA's 11 regions: nation > zone > region
class(gs)
```

::: {.callout-important}
## The one sentence to remember
A geoscale is used for **plotting, reporting and subsetting only**. Attaching
one *never changes the optimisation model*. The extra members stay inert until
some commodity names a coarser level.
:::

- `setGeoscale()` / `getGeoscale()` on a `config`, `model` or `scenario`.
- `getData(..., geolevel = )` aggregates results up the hierarchy — `"finest"`,
  `"coarsest"`, `"all"`, or a named level.
- `plot_map()` draws results on the geometry (needs `sf`).
- Inter-regional **flows** are returned unchanged by `geolevel` aggregation —
  summing a flow across regions would double-count it.

# 4 · Commodities in depth {background-color="#f4f4f4"}

## Nine slots, and what each decides {.smaller}

::: {.small-table}
| Slot | Decides |
|---|---|
| `name`, `desc` | identity |
| `limtype` | how the balance closes: `LO` (default, excess allowed), `UP`, `FX` |
| `timeframe` | **where it balances in time** — finest level in the model by default |
| `geolevel` | **where it balances in space** [GLPK / GAMS only]{.badge .limited} |
| `unit` | the commodity's main unit |
| `emis` | emission factors carried by the fuel |
| `agg` | weights aggregating several commodities into this one |
| `misc` | free-form; never read by the model |
:::

. . .

The commodity is deliberately **lean**: it is a label with balancing rules. All
the engineering lives on the processes that consume and produce it.

## Where a commodity lives {.smaller}

::: {.columns}
::: {.column width="50%"}
**In time — `timeframe`**

```{r}
#| eval: false
newCommodity("ELC", timeframe = "HOUR")
newCommodity("COA", timeframe = "ANNUAL")
```

Electricity must balance every hour; coal only has to balance over the year.
Setting `ANNUAL` on a carrier that genuinely needs hourly matching is one of the
easiest ways to build a model that is quietly too optimistic.
:::
::: {.column width="50%"}
**In space — `geolevel`** [experimental]{.badge .exp}

```{r}
#| eval: false
newCommodity("STEEL", geolevel = "nation")
```

Naming a coarser level asserts **free, unlimited transport within that level**.
That suits an integrated goods market; it is wrong for a network-constrained
carrier such as electricity.

Requires a geoscale on the model config.
:::
:::

## Carbon and other emissions {.smaller}

::: {.columns}
::: {.column width="54%"}
Emission factors ride on the **fuel**, so every technology burning it inherits
them:

```{r}
COA2 <- newCommodity("COA", unit = "PJ",
  emis = data.frame(comm = "CO2", unit = "kt/PJ", emis = 95))
GAS2 <- newCommodity("GAS", unit = "PJ",
  emis = data.frame(comm = "CO2", unit = "kt/PJ", emis = 56))
```

The factor only fires where the consuming technology declares
`combustion = 1` on that input.

There is **no separate carbon-content slot** — `@emis` is the single mechanism,
and it generalises to any emitted species.
:::
::: {.column width="46%"}
```{r}
#| echo: false
#| fig-width: 4.8
#| fig-height: 3.6
autoplot(COA2, GAS2)
```
:::
:::

::: {.callout-note}
## Roadmap [planned]{.badge .planned}
Physical properties — mass, volume, heating value — are **not** slots today, and
`@misc` is the documented place to park them meanwhile. Adding them as
first-class, unit-aware slots is a candidate for a future release.
:::

## Plant-specific pollutants are different {.smaller}

A useful contrast, since both end up as "emissions":

| Where it lives | Mechanism | Use when |
|---|---|---|
| **Commodity** `@emis` | factor per unit of fuel burned | CO~2~ — depends only on the fuel's carbon |
| **Technology** `aeff` `cinp2aout` | auxiliary *output* scaled by input | NO~x~, SO~x~ — depend on the burner, not just the fuel |

Two plants burning identical coal emit identical CO~2~ and quite different NO~x~.
That is the whole distinction.

# 5 · Vintages and clusters {background-color="#f4f4f4"}

## One technology, many variants {.smaller}

Real technologies are not one thing. A wind fleet has **resource grades** that
differ in cost and yield, and **build years** that differ in efficiency and
lifetime. energyRt expresses both on a *single* object:

| Dimension | Slot | Means |
|---|---|---|
| **vintage** | `@vintage` | when it was built — investment window and operational life |
| **cluster** | `@cluster` | which grade or variant it is — resource quality, siting class |

. . .

`@vintage` **replaces** the older `start` / `end` / `olife` slots: one row per
(`vintage`, `region`, `cluster`) giving that variant's window and life. A single
row with `vintage = NA` is an ordinary un-vintaged technology.

## Declaring vintages {.smaller}

::: {.columns}
::: {.column width="55%"}
The short form is usually enough — `start` and `end` **default to the vintage
year**, so naming the years defines properly windowed vintages:

```{r}
EWIN <- newTechnology(
  name    = "EWIN",
  input   = list(comm = "WIN"),
  output  = list(comm = "ELC"),
  ceff    = data.frame(comm = "WIN", cinp2use = 1),
  vintage = c("2025", "2040"),
  cap2act = 31.536
)
EWIN@vintage
```
:::
::: {.column width="45%"}
The table form when the variants differ:

```{r}
#| eval: false
vintage = data.frame(
  vintage = c("2025", "2040"),
  olife   = c(25L, 30L)      # newer lasts longer
)
```

Columns available: `vintage`, `region`, `cluster`, `start`, `end`, `olife`.
:::
:::

## Declaring clusters {.smaller}

```{r}
EWINVC <- newTechnology(
  name    = "EWINVC",
  desc    = "Wind: two resource grades, two vintages",
  input   = list(comm = "WIN"),
  output  = list(comm = "ELC"),
  ceff    = data.frame(comm = "WIN", cinp2use = 1),
  cluster = data.frame(cluster = c("best", "mid"),
                       desc    = c("High wind sites", "Medium wind sites"),
                       order   = 1:2),
  vintage = data.frame(vintage = c("2025", "2040"), olife = 25L),
  invcost = data.frame(cluster = c("best", "best", "mid",  "mid"),
                       vintage = c("2025", "2040", "2025", "2040"),
                       invcost = c(1400,   1150,   1600,   1300)),
  cap2act = 31.536
)
```

. . .

**Nothing new was learned.** Parameters vary by variant by gaining one more
selector column — exactly like `region` or `year`. `2 × 2 = 4` cells, one
object.

## `draw()` reads the whole grid {.smaller}

::: {.columns}
::: {.column width="42%"}
Ask for one cell and you get it, with the other vintages behind it as **ghosts**:

```{r}
#| eval: false
draw(EWINVC, vintage = "2040",
             cluster = "best")
```

The faded stack is the *vintage marginal* through the selected cell — the cells
sharing its cluster. Earlier vintages sit down-left, later ones up-right, so the
depth of the stack is the technology's history.
:::
::: {.column width="58%"}
```{r}
#| echo: false
#| fig-width: 6.4
#| fig-height: 4.8
draw(EWINVC, vintage = "2040", cluster = "best")
```
:::
:::

## Clusters: rail, deck, or neither {.smaller}

`cluster_style` changes how the sibling grades are rendered around the selected
cell.

::: {.columns}
::: {.column width="33%"}
```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 3.6
draw(EWINVC, vintage = "2040", cluster_style = "rail")
```
`"rail"` *(default)*
:::
::: {.column width="33%"}
```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 3.6
draw(EWINVC, vintage = "2040", cluster_style = "deck")
```
`"deck"`
:::
::: {.column width="33%"}
```{r}
#| echo: false
#| fig-width: 4.2
#| fig-height: 3.6
draw(EWINVC, vintage = "2040", cluster_style = "none")
```
`"none"`
:::
:::

Cosmetic, not semantic — pick whichever makes the point on your slide.

## Seeing every cell at once {.smaller}

Pass `"all"` to facet a dimension instead of selecting from it:

```{r}
#| fig-width: 11
#| fig-height: 4.6
draw(EWINVC, vintage = "all", cluster = "all")
```

`max_facets = 24L` guards the obvious accident — a 4-vintage × 11-region
technology asked to draw everything.

## From variants to model objects {.smaller}

Variants are **replicated into ordinary technologies before interpolation** —
the solver never sees the vintage/cluster idea, only more technologies.

```{r}
#| eval: false
mod <- expand_variants(mod)     # or expand_tech_variants(mod)
```

Names gain a suffix from `.variant_prefix`, defaulting to `_VIN` and `_CL`:
`EWINVC` becomes `EWINVC_CLbest_VIN2025`, `EWINVC_CLmid_VIN2040`, and so on.

. . .

Reading them back after a solve:

```{r}
#| eval: false
getVariants(scen)                                   # which variants exist
variantSummary(scen, name = "EWINVC", by = "vintage")  # rolled up again
```

The point of the design: **declare once, compare by grade or by build year**,
without hand-maintaining a dozen near-identical technology objects.

## `design()` — the object, back as code {.smaller}

::: {.columns}
::: {.column width="46%"}
Useful when an object came from a dataset or a colleague and you want to see
what it actually says:

```{r}
#| eval: false
design(ECOA)
```

Returns (and can write to a file) the `newTechnology()` call that recreates it.

Note it renders the **modern `vintage` table** even when the object was built
with the older `olife =` argument — a quick way to migrate old code.

[technology only]{.badge .limited} — no `design()` method for other classes yet.
:::
::: {.column width="54%"}
```{r}
#| echo: false
#| class-output: small-table
design(ECOA)
```
:::
:::

# 6 · From description to numbers {background-color="#f4f4f4"}

## What `interpolate_model()` does {.smaller}

The model object is a *description*. Interpolation compiles it into a
solver-ready scenario:

```{r}
#| eval: false
scen <- interpolate_model(mod, name = "BASE")
```

1. collect the sets — regions, years, slices, commodities, processes
2. build the **maps**: membership, calendar, lifespan, value and cost domains
3. extract the numeric parameters and **interpolate them over years**
4. compute equivalent annual costs
5. compile user constraints
6. reduce the data (prune / fold / densify / trim — Part 8)
7. validate

. . .

Only step 3 is interpolation in the literal sense. Regions and slices are
handled by **expanding `NA` wildcards** — an omitted `region` column means *all
regions*, and it is expanded, not interpolated.

## Anchors in, series out {.smaller}

::: {.columns}
::: {.column width="46%"}
You supply anchor years; the pipeline fills the rest:

```{r}
ECOA_Y <- newTechnology(
  name    = "ECOA_Y",
  input   = list(comm = "COA", combustion = 1),
  output  = list(comm = "ELC"),
  ceff    = data.frame(comm = "COA", cinp2use = 0.40),
  invcost = data.frame(year    = c(2025, 2040, 2050),
                       invcost = c(2000, 1800, 1700)),
  cap2act = 31.536, olife = 30L
)
```

`getData()` returns both views — the given rows, and the interpolated series:

```{r}
#| eval: false
getData(ECOA_Y, name = "invcost", merge = FALSE)
getData(ECOA_Y, name = "invcost", merge = FALSE,
        interpolate = TRUE, year = 2025:2050)
```
:::
::: {.column width="54%"}
```{r}
#| echo: false
#| fig-width: 5.6
#| fig-height: 4.4
given <- getData(ECOA_Y, name = "invcost", merge = FALSE)$invcost
itp   <- getData(ECOA_Y, name = "invcost", merge = FALSE,
                 interpolate = TRUE, year = 2025:2050)$invcost
ggplot() +
  geom_line(data = itp, aes(year, invcost), colour = "#2C6E9B", linewidth = .9) +
  geom_point(data = given, aes(year, invcost), size = 3, colour = "#B7410E") +
  labs(title = "ECOA_Y · invcost",
       subtitle = "points: given anchors · line: interpolated",
       x = "year", y = "MEUR/GW") +
  theme_energyRt()
```
:::
:::

## Interpolation rules {.smaller}

Each parameter carries a rule, stored as a composite of three tokens:

| Token | Alias | Applies |
|---|---|---|
| `back` | `bwd` | before the first anchor — backward fill |
| `inter` | `mid` | between anchors — linear |
| `forth` | `fwd` | after the last anchor — forward fill |

The default nearly everywhere is **`back.inter.forth`**: constant outside the
anchors, linear between them. That is why a single value with no `year` column
becomes a flat line across the whole horizon.

. . .

Rules live in the `config@interpolation` table. To change them, build a modified
`config` (or `settings`) and pass the object —
there is **no scalar `interpolate_model(interpolation = ...)` argument yet**
[planned]{.badge .planned}.

# 7 · Scenarios {background-color="#f4f4f4"}

## A scenario is interpolated/solved model with additions {.smaller}

`interpolate_model()` accepts **energyRt objects in `...`** and folds them into
the model before the pipeline runs. That single mechanism is the whole scenario
API:

```{r}
#| eval: false
interpolate_model(mod, "CO2CAP",  co2_cap)                  # a policy constraint
interpolate_model(mod, "SAMPLED", cal_sampled)              # a sampled calendar
interpolate_model(mod, "SHORT",   newHorizon(period = 2025:2035))
interpolate_model(mod, "HIGHDEM", repo_high_demand)         # a whole repository
```

| Pass an object of class | It replaces / adds |
|---|---|
| `calendar` | the scenario's time structure |
| `horizon` | the milestone years |
| `config` / `settings` | the whole settings block |
| `repository` | added to the model |
| `technology`, `commodity`, `constraint`, `tax`, … | bundled into a scenario-specific repository |

Anything unrecognised is **warned about**, not silently ignored.

## The same objects work at solve time {.smaller}

`solve_scenario()` / `solve_mod()` split their `...`: run-control arguments stay,
everything else is routed straight to interpolation.

```{r}
#| eval: false
solve_mod(mod, "COARSE", solver = solver_options$glpk,
          calendar = calendars$utopia_seasons)   # -> forwarded to interpolation
```

. . .

The scenario folder name records what was swapped —
`{scenario}_{model}_{calendar}_{horizon}` — computed from the **final** settings,
so a substituted calendar shows up in the path:

```
COARSE_UTOPIA_utopia-seasons_base/
```

::: {.callout-note}
## Not yet [planned]{.badge .planned}
Swapping **regions**, and named scalar overrides such as
`interpolate_model(mod, "S", discount = 0.07)`. Both are TODOs in the pipeline;
for now, pass a full `config`/`settings` object.
:::

# 8 · Scale: data and models {background-color="#f4f4f4"}

```{r}
#| echo: false
set_scenarios_path(file.path(tempdir(), "deck-scenarios"))
um <- utopia_modules$electricity$R1
mod_u <- newModel("UTOPIA", data = um$repo,
                  calendar = utopia_modules$calendars$utopia_s4h24,
                  region   = um$regions,
                  horizon  = utopia_modules$horizons$base, discount = 0.05)
scen_u <- interpolate_model(mod_u, name = "BASE", verbose = FALSE) |>
  solve_scenario(solver = solver_options$glpk, echo = FALSE)
```

## How big is this model? {.smaller}

```{r}
model_size(scen_u, top_n = 5)
```

`model_size()` counts parameter rows and **estimates** variables and constraints
from the gating maps — enough to decide whether a run is worth starting.

## Four ways to make it smaller {.smaller}

::: {.small-table}
| Operation | What it does | Lossless? |
|---|---|---|
| **prune** | drops rows whose value equals the parameter's default | yes — an absent tuple reads back as the default |
| **fold** | replaces a whole column with `NA` wildcards where the value does not vary across the *entire* allowed membership | yes — reversible with `unfold_*` |
| **trim** | drops rows that no equation-domain map indexes, i.e. data the model never reads | yes |
| **densify** | the **opposite** — materialises default rows, because GAMS has no parameter default and reads an absent tuple as `0` | n/a |
:::

```{r}
#| eval: false
interpolate_model(mod, "BASE",
                  sparse = TRUE,   # the storage knob: drop defaults, enable folding
                  prune  = TRUE,
                  fold   = c("region", "slice"))
```

. . .

[GLPK / JuMP / Pyomo only]{.badge .limited} — folding writes an artificial set
member (`ANYREGION`, `ANYSLICE`, …) into the model source, and the substitution
is not GAMS-ready yet.

## The cheapest reduction is a sampled calendar {.smaller}

Keep the same time *structure*, solve a representative **subset** of its slices —
here every fourth hour of each season, and `year_fraction` records what is
covered:

```{r}
tt    <- calendars$utopia_s4h24@timetable
keep  <- tt[grepl("h(00|04|08|12|16|20)$", tt$HOUR), ]
cal_s <- newCalendar(name = "utopia_s4h24_sampled", timetable = keep,
                     year_fraction = sum(keep$share))
```

```{r}
#| echo: false
scen_s <- interpolate_model(mod_u, "SAMPLED", cal_s, verbose = FALSE)
sz <- function(s, lbl) {
  m <- model_size(s, top_n = 1)
  data.frame(scenario   = lbl,
             calendar   = as.character(s@settings@calendar@name),
             slices     = nrow(s@settings@calendar@timetable),
             param_rows = format(m$param_rows, big.mark = ","),
             variables  = format(m$n_var_est, big.mark = ","))
}
knitr::kable(rbind(sz(scen_u, "BASE"), sz(scen_s, "SAMPLED")), row.names = FALSE)
```

. . .

A factor of roughly **4×** in rows and variables, from one substituted object.
Reduce the *model* before reaching for storage tricks.

## Sampling vs aggregation {.smaller}

Two different things, and only one of them exists today:

| | What it does | Status |
|---|---|---|
| **Sampling** | solve a *subset* of the calendar's slices; `year_fraction` scales annual quantities | works — substitute a sampled calendar |
| **Aggregation** | roll a fine calendar's data *up* onto a coarser one — reweighting profiles, load curves, availability | [in development]{.badge .planned} |

. . .

Passing a structurally different, coarser calendar will build a smaller model,
but the fine-resolution data is **not aggregated onto it** — that machinery is
still being written, and lands with the **timescales** package. Until then,
sampling is the supported way to shrink the time dimension of an existing model.

Aggregation on the **results** side already works: `getData(timeframe = )`
(Part 2).

::: {.callout-tip}
Whichever route you take, `compare_interp_settings()` and
`compare_solve_settings()` rebuild the same model across a grid of
`fold`/`sparse`/`prune` settings and check the **objective is invariant** — the
check worth running before trusting a reduction.
:::

## Large datasets: in memory or on disk {.smaller}

```{r}
#| eval: false
interpolate_model(mod, "BIG", ondisk = TRUE)   # parameters written to disk
save_scenario(scen, path = "scenarios/BIG")    # scenario spilled to parquet
load_scenario("scenarios/BIG")
```

- **`ondisk = TRUE`** writes each parameter to a `modInp/` folder instead of
  holding it in `@data`.
- **`save_scenario()`** stores data-frame slots as **parquet** (zstd by default)
  and empties the in-memory slots; `isInMemory()` tells you which state an object
  is in.

::: {.callout-warning}
## Two honest caveats
`ondisk` reduces peak memory **during the build only** — everything is
materialised back into RAM before the solver files are written.

And `getData()` **collects eagerly**: the Arrow dataset is opened lazily, but
filtering happens in R afterwards. There is no predicate push-down, so this is
not larger-than-memory querying.
:::

## Trimming variables [planned]{.badge .planned} {.smaller}

Everything above operates on **parameter rows**. Removing the *variables* and
*equations* those rows would have gated is a different job — an equation-graph
cascade: mark empty-domain variables, untrim the ones still referenced, drop the
rest.

That cascade lives in the sibling **multimod** package (`trim_model()`,
`mark_empty_variables()`, `untrim_required_elements()`), and energyRt does not
call it yet. A partial removal on the energyRt side would leave dangling
variable references in the equations that read them — which is precisely why it
is not attempted here.

Integration with multimod is in progress and is the route by which this arrives.

# 9 · Reading results {background-color="#f4f4f4"}

## `getData()` — one accessor, many sources {.smaller}

```{r}
#| eval: false
getData(scen,  name = "vTechAct")        # a solved scenario: results
getData(mod,   name = "invcost")         # a model: raw input data, pre-interpolation
getData(ECOA,  name = "ceff")            # a single object: its own slots
getData(list(BASE = s1, CAP = s2))       # a named list: stacked, with a `scenario` column
```

Key arguments: `timeframe` and `geolevel` (aggregation, Parts 2–3), `merge`,
`process`, `parameters`, `variables`, `drop.zeros`, `asTibble`.

. . .

The **named list** form is how many scenarios are compared — there is no need
for a registry to hold them.

## Mixes and comparisons {.smaller}

::: {.columns}
::: {.column width="48%"}
```{r}
#| eval: false
autoplot(scen_u)              # generation / capacity mixes
getMix(list(BASE = scen_u, SAMPLED = scen_s))
```

`getMix()` is the tidy extractor behind the scenario plot. Given a **named
list** it row-binds with a `scenario` column, so any `ggplot2` comparison is a
`facet_wrap(~ scenario)` away.

`autoplot()` itself takes a single scenario — build multi-scenario charts from
`getMix()`.
:::
::: {.column width="52%"}
```{r}
#| echo: false
#| fig-width: 6.0
#| fig-height: 4.6
autoplot(scen_u)
```
:::
:::

## Levelized costs and reports {.smaller}

```{r}
#| eval: false
lc <- levcost(scen_u, name = "ECOA")
autoplot(lc, type = "components")   # also: npv, totals, frontier, input_frontier

report(scen_u)                       # results overview -> HTML
report(mod_u)                        # full model report: config, inventory, every technology
report(ECOA, format = "pdf")         # a single technology datasheet
```

- `report()` works on a `technology`, `repository`, `model` or solved `scenario`.
- Formats: `html`, `pdf`, `tex` — it downgrades to HTML with a warning if no
  LaTeX engine is found.
- **Templates are customisable**: pass the absolute path of your own `.Rmd` to
  `template =`. The built-ins live in `inst/templates/`
  (`report_generic`, `report_model`, `report_scenario`, `report_vehicle`).

# 10 · How it is solved {background-color="#f4f4f4"}

## One solution type today {.smaller}

energyRt writes **one linear program over all milestone years and solves it
once**: full **perfect foresight**. Every backend template contains a single
solve statement.

That is a modelling assumption, not an implementation detail — the model knows
the whole future when it invests. It is the right default for a least-cost
benchmark and the wrong one for studying myopic behaviour.

::: {.callout-note}
## Under development [planned]{.badge .planned}
Myopic · myopic with a guide or target · isolated regional solves ·
geo-aggregated solves · free/no-trade solves.

Also listed alongside these: **coarse-time** solving, which needs the
aggregation machinery described in Part 8 — in development.

**Time-sampled** solving is the one you can already do today, by substituting a
sampled calendar (Parts 2, 7 and 8).
:::

## Backends and solvers {.smaller}

::: {.columns}
::: {.column width="52%"}
| Backend | Needs locally |
|---|---|
| **GLPK / MathProg** | `glpsol` (ships with Rtools) |
| **Julia / JuMP** | Julia + HiGHS |
| **Python / Pyomo** | Python + CBC/HiGHS |
| **GAMS** | GAMS + a license |
| **NEOS** | *nothing* — an email address |

The same model object solves on any of them.
:::
::: {.column width="48%"}
```{r}
#| eval: false
solver_options$glpk
solver_options$julia_highs_barrier
solver_options$neos_gams_cplex
```

26 presets ship with the package. NEOS submits the model as inlined text, so a
CPLEX solve needs no local GAMS at all — subject to its academic/non-commercial
terms and a ~16 MB job cap.

Cross-solving one scenario on two backends and comparing objectives is the
strongest correctness check available.
:::
:::

# 11 · Methods at a glance {background-color="#f4f4f4"}

## What you can call on what {.smaller}

::: {.small-table}
| Method | Returns | Defined for |
|---|---|---|
| `draw()` | grid schematic | `technology`, `storage`, `supply`, `demand`, `import`, `export`, `trade` |
| `autoplot()` | **ggplot** | 19 classes — `calendar`, `horizon`, `commodity`, `weather`, `trade`, `demand`, the processes, `levcost`, `scenario`, `model`, `repository` |
| `plot()` | ggplot | 17 classes — a thin delegate to `autoplot()` |
| `getData()` | tidy data | scenarios (and lists of them), models, repositories, single objects |
| `getObject()` | the objects | repository, model, scenario — filtered by class, name, region, … |
| `report()` | html / pdf / tex | `technology`, `repository`, `model`, `scenario` |
| `design()` | R code | `technology` [only]{.badge .limited} |
| `summary()` / `show()` | console | `model`, `scenario`, `repository` [only]{.badge .limited} |
:::

. . .

Two things worth knowing before you go looking: there is **no `draw()` for a
model or repository** — `draw()` describes a process, and `report(mod)` is what
describes a whole model. And a process `autoplot()` needs year-indexed data in
the economics or capacity slots; with none it returns `NULL` and says so.

## The workhorses {.smaller}

::: {.columns}
::: {.column width="50%"}
**While building**

- `draw(obj)` — is this the system I meant?
- `design(tech)` — what does this object actually say?
- `getObject(repo, class = "technology")` — what is in here?
- `autoplot(cal)`, `autoplot(hor)` — is the time structure right?
:::
::: {.column width="50%"}
**After solving**

- `autoplot(scen)` — the mixes
- `getMix(list(...))` — compare scenarios
- `levcost(scen, name =)` — why did it pick that?
- `model_size(scen)` — why is this so slow?
- `report(scen)` — everything, in one document
:::
:::

# 12 · Extending {background-color="#f4f4f4"}

## Your own constraints {.smaller}

The escape hatch that avoids forking the model code:

```{r}
#| eval: false
newConstraint(
  name  = "RES_SHARE",
  eq    = ">=",
  list(variable = "vTechOut", for.sum = list(tech = c("EWIN", "ESOL"))),
  rhs   = data.frame(year = c(2030, 2050), rhs = c(0.3, 0.6))
)
```

> *In many cases this can be done without writing constraints in the GAMS,
> Julia/JuMP, Python/Pyomo, or GLPK-MathProg languages.*

Each left-hand-side term is a **summand**: a `variable`, a multiplier `mult`, and
what to sum it `for.sum` over.

[experimental]{.badge .exp} — and user constraints currently compile to
GAMS-dialect text; moving them to a language-neutral form so every backend
renders from one definition is in progress.

## The other extension points {.smaller}

| Want to | Use |
|---|---|
| add a relationship between variables | `newConstraint()` [experimental]{.badge .exp} |
| add a cost term to the objective | `newCosts()` [experimental]{.badge .exp} |
| change what a report shows | `report(template = "/path/to/your.Rmd")` |
| carry your own metadata on any object | the `@misc` slot — never read by the model |
| change solver behaviour | `solver_options$...`, or `set_option()` / `en_config_write()` |
| add a new mapping to the pipeline | the mapping registry — a *contributor* path: declare it in `modInp.yml`, write `map_<Name>()`, register it |

. . .

Not available yet: **`newVariable()`** [planned]{.badge .planned}. The variable
catalogue is fixed; a runtime constructor paralleling `newConstraint()` is on the
roadmap. (The `variable` class you can see is the results container, not a
constructor.)

# 13 · Where this is going {background-color="#f4f4f4"}

## Roadmap {.smaller}

::: {.columns}
::: {.column width="50%"}
**Structure**

- Nested **regions** via `geoscales` and nested **time** via `timescales` —
  today both are flat sets with aggregation only on the reporting side
- `newVariable()`, to match `newConstraint()`
- Language-neutral user constraints, rendered to every backend from one
  definition
- Physical properties on `commodity` (mass, volume, heating value)
- Region swapping and scalar setting overrides in scenario definition
:::
::: {.column width="50%"}
**Scale and solving**

- Alternative solution types — myopic first
- multimod integration: the variable/equation **trim** cascade, and one shared
  model definition across packages
- Lazy, push-down querying of on-disk results
- Scenario **registry** — a working index over many saved scenarios
  [experimental today]{.badge .exp}
:::
:::

. . .

Most of the packages around energyRt are pre-release. If a piece of this is on
your critical path, say so — that is the most useful thing you can tell us.

## Recap {.smaller}

::: {.columns}
::: {.column width="50%"}
1. **Objects** describe the system — `new*()`, one grammar throughout
2. **Calendars and horizons** decide where those objects live in time;
   **regions and geoscales**, in space
3. **Vintages and clusters** put many variants on one object
4. **Interpolation** turns anchors into a full grid
:::
::: {.column width="50%"}
5. **Scenarios** are the same model with objects substituted
6. **Scale** is managed first by sampling the calendar, then by pruning, folding
   and trimming its data
7. **Results** come back tidy — `getData()`, `autoplot()`, `levcost()`,
   `report()`
:::
:::

. . .

**When in doubt, `draw()` it** — and when the model is slow, `model_size()` it.
